---
title: "P697: Differential Accessibility, Pseudobulk DGE, and GSEA"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
setwd("/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

library(knitr)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(viridis)
library(stringr)
library(patchwork)
library(Seurat)
library(Signac)
library(edgeR)
library(limma)
library(clusterProfiler)
library(msigdbr)
library(ggrepel)
library(RColorBrewer)

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      axis.text = element_text(colour = "black"),
      axis.ticks = element_line(colour = "black"),
      legend.key = element_blank(),
      strip.text.x = element_text(size = 14, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(fill = "white", colour = "black")))

opts_chunk$set(fig.width = 6, fig.height = 4.0, cache = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE, cache.lazy = FALSE, results = "hide")
opts_knit$set(root.dir = "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

options(stringsAsFactors = FALSE)
knitr::opts_chunk$set(error = FALSE)
options(future.globals.maxSize = 1591289600)
```

# Setup

```{r setUpDirectories}
projNumber <- "P697"
baseDir <- "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2"
dataInputDir <- file.path(baseDir, "data/inputData")
plotDir <- file.path(baseDir, "figures")
dataOutputDir <- file.path(baseDir, "data/outputData")
tableDir <- file.path(baseDir, "tables")
if (!dir.exists(tableDir)) dir.create(tableDir, recursive = TRUE)
dataDate <- Sys.Date()
filenameSuffix <- paste0(projNumber, ".", dataDate)
```

```{r setupFunctions}
source(file.path(getwd(), "code/P697_functions.R"))
```

# Load WNN Bundle

```{r loadBundle}
bundleFiles <- list.files(dataOutputDir,
  pattern = "P697\\..*_WNN_bundle\\.rds$", full.names = TRUE)
stopifnot("No WNN bundle found in data/outputData" = length(bundleFiles) > 0)
bundleFile <- bundleFiles[which.max(file.mtime(bundleFiles))]
cat("Loading WNN bundle:", basename(bundleFile), "\n")

wnnBundle <- readRDS(bundleFile)

seuratList <- list(
  NK = wnnBundle$seuratNK,
  Treg = wnnBundle$seuratTreg
)
compartments <- c("NK", "Treg")

# The downstream updates in this script operate on the existing WNN bundle.
# Peak-calling does not need to be re-run unless the ATAC feature space changes.

# TODO: clean up this hack rename stuff after we regenerate the bundle
# Rename Washout → Followup to match updated nomenclature
for (comp in compartments) {
  lvls <- levels(seuratList[[comp]]$treatmentPhase)
  if ("Washout" %in% lvls) {
    levels(seuratList[[comp]]$treatmentPhase)[lvls == "Washout"] <- "Followup"
    cat(sprintf("  %s: renamed Washout → Followup in factor levels\n", comp))
  }
}

# Palettes
palDonorId <- wnnBundle$palDonorId
palVisit <- wnnBundle$palVisit
palTreatmentPhase <- wnnBundle$palTreatmentPhase
names(palTreatmentPhase) <- gsub("Washout", "Followup", names(palTreatmentPhase))

for (comp in compartments) {
  cat(sprintf("%s: %d cells, %d WNN clusters\n",
    comp, ncol(seuratList[[comp]]),
    length(unique(seuratList[[comp]]$wnn_clusters))))
}

rm(wnnBundle)
gc(verbose = FALSE)
```

# Define Contrasts

```{r defineContrasts}
# Longitudinal contrasts (all relative to baseline and consecutive)
contrastDefs <- list(
  PostIL2_vs_BL    = c("Post-IL2", "Baseline"),
  PostRAPA_vs_BL   = c("Post-RAPA", "Baseline"),
  Followup_vs_BL    = c("Followup", "Baseline"),
  PostRAPA_vs_PostIL2  = c("Post-RAPA", "Post-IL2"),
  Followup_vs_PostRAPA  = c("Followup", "Post-RAPA"),
  PostIL2_vs_Followup  = c("Post-IL2", "Followup"),
  PostRAPA_vs_Followup = c("Post-RAPA", "Followup")
)

primaryContrasts <- c(
  "PostIL2_vs_BL",
  "PostRAPA_vs_BL",
  "Followup_vs_BL",
  "Followup_vs_PostRAPA"
)

cat("Contrasts defined:\n")
for (nm in names(contrastDefs)) {
  cat(sprintf("  %s: %s vs %s\n", nm, contrastDefs[[nm]][1], contrastDefs[[nm]][2]))
}

cat("\nPrimary PI-facing contrasts:\n")
for (nm in primaryContrasts) {
  cat(sprintf("  %s\n", nm))
}
```

# Helper Functions

```{r helperFunctions}
# Pseudobulk aggregation: sum counts per donor × treatment phase
aggregatePseudobulk <- function(countMat, metadata, groupCols) {
  metadata$groupDGE <- apply(metadata[, groupCols, drop = FALSE], 1, paste, collapse = "_")

  # Sum counts per group
  groups <- unique(metadata$groupDGE)
  aggMat <- matrix(0, nrow = nrow(countMat), ncol = length(groups),
    dimnames = list(rownames(countMat), groups))

  for (g in groups) {
    cells <- which(metadata$groupDGE == g)
    if (length(cells) == 1) {
      aggMat[, g] <- countMat[, cells]
    } else {
      aggMat[, g] <- Matrix::rowSums(countMat[, cells])
    }
  }

  # Build design info
  designInfo <- metadata %>%
    distinct(groupDGE, .keep_all = TRUE) %>%
    dplyr::select(all_of(c(groupCols, "groupDGE")))
  designInfo$nCells <- as.integer(table(metadata$groupDGE)[designInfo$groupDGE])
  designInfo <- designInfo[match(colnames(aggMat), designInfo$groupDGE), ]

  list(counts = aggMat, designInfo = designInfo)
}

# Run limma-voom DGE/DA with paired donor design
runLimmaVoom <- function(aggCounts, designInfo, contrastDefs,
                         filterExpr = TRUE, minCellsPerGroup = 3) {
  # Require minimum cells per group
  keep <- designInfo$nCells >= minCellsPerGroup
  if (sum(!keep) > 0) {
    cat(sprintf("  Dropped %d groups with < %d cells\n", sum(!keep), minCellsPerGroup))
    aggCounts <- aggCounts[, keep]
    designInfo <- designInfo[keep, ]
  }

  # Paired design: ~ 0 + treatmentPhase + donorID
  designInfo$treatmentPhase <- factor(designInfo$treatmentPhase,
    levels = c("Baseline", "Post-IL2", "Post-RAPA", "Followup"))
  designInfo$donorID <- factor(designInfo$donorID)

  design <- model.matrix(~ 0 + treatmentPhase + donorID, data = designInfo)
  colnames(design) <- gsub("treatmentPhase", "", colnames(design))
  colnames(design) <- gsub("donorID", "D", colnames(design))
  colnames(design) <- make.names(colnames(design))

  # DGEList + filtering + normalization
  dge <- DGEList(counts = aggCounts)
  if (filterExpr) {
    keep <- filterByExpr(dge, design = design)
    cat(sprintf("  filterByExpr: keeping %d / %d features\n", sum(keep), length(keep)))
    dge <- dge[keep, , keep.lib.sizes = FALSE]
  }
  dge <- calcNormFactors(dge, method = "TMM")

  # voom
  v <- voom(dge, design, plot = FALSE)

  # Fit
  fit <- lmFit(v, design)

  # Build contrast matrix from contrastDefs
  # Only include contrasts where both levels are present in the design
  validContrasts <- list()
  for (nm in names(contrastDefs)) {
    lvl1 <- make.names(contrastDefs[[nm]][1])
    lvl2 <- make.names(contrastDefs[[nm]][2])
    if (lvl1 %in% colnames(design) && lvl2 %in% colnames(design)) {
      validContrasts[[nm]] <- paste0(lvl1, " - ", lvl2)
    }
  }

  if (length(validContrasts) == 0) {
    cat("  No valid contrasts — skipping\n")
    return(NULL)
  }

  contStr <- unlist(validContrasts)
  cont.matrix <- makeContrasts(contrasts = contStr, levels = design)
  colnames(cont.matrix) <- names(validContrasts)

  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2)

  # Extract results
  results <- list()
  for (nm in names(validContrasts)) {
    tt <- topTable(fit2, coef = nm, adjust.method = "BH",
      sort.by = "P", number = Inf)
    tt$feature <- rownames(tt)
    tt$contrast <- nm
    results[[nm]] <- tt
  }

  list(results = results, fit = fit2, voom = v, design = design,
    contMatrix = cont.matrix, designInfo = designInfo)
}

# Volcano plot
plotVolcano <- function(tt, contrastName, featureType = "Gene",
                        pCutoff = 0.05, fcCutoff = 0.5, nLabel = 15) {
  tt$sig <- ifelse(tt$adj.P.Val < pCutoff & abs(tt$logFC) > fcCutoff,
    ifelse(tt$logFC > 0, "Up", "Down"), "NS")
  tt$sig <- factor(tt$sig, levels = c("Up", "Down", "NS"))

  nUp <- sum(tt$sig == "Up")
  nDown <- sum(tt$sig == "Down")

  # Label top features
  topFeatures <- tt %>%
    dplyr::filter(sig != "NS") %>%
    arrange(adj.P.Val) %>%
    head(nLabel)

  ggplot(tt, aes(x = logFC, y = -log10(adj.P.Val), color = sig)) +
    geom_point(size = 0.8, alpha = 0.6) +
    geom_hline(yintercept = -log10(pCutoff), linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = c(-fcCutoff, fcCutoff), linetype = "dashed", color = "grey50") +
    scale_color_manual(values = c("Up" = "#E41A1C", "Down" = "#377EB8", "NS" = "grey70"),
      drop = FALSE) +
    geom_text_repel(data = topFeatures,
      aes(label = feature), size = 3, max.overlaps = 20, color = "black") +
    labs(x = "log2 Fold Change", y = "-log10(Adj. P-value)",
      title = contrastName,
      subtitle = sprintf("%s: %d up, %d down (FDR < %.2f, |logFC| > %.1f)",
        featureType, nUp, nDown, pCutoff, fcCutoff)) +
    theme(legend.position = "none")
}

buildCoverageTable <- function(seuratObj) {
  donorLevels <- sort(unique(as.character(seuratObj$donorID)))
  phaseLevels <- c("Baseline", "Post-IL2", "Post-RAPA", "Followup")

  seuratObj@meta.data %>%
    dplyr::count(donorID, treatmentPhase, name = "nCells") %>%
    tidyr::complete(
      donorID = donorLevels,
      treatmentPhase = factor(phaseLevels, levels = phaseLevels),
      fill = list(nCells = 0)
    ) %>%
    dplyr::mutate(
      donorID = factor(donorID, levels = donorLevels),
      treatmentPhase = factor(treatmentPhase, levels = phaseLevels)
    )
}

summarizeCoverageTable <- function(coverageDf, compartment) {
  coverageDf %>%
    dplyr::group_by(treatmentPhase) %>%
    dplyr::summarise(
      compartment = compartment,
      nDonorsWithCells = sum(nCells > 0),
      medianCellsPerDonor = ifelse(any(nCells > 0), median(nCells[nCells > 0]), NA_real_),
      minCellsPerDonor = ifelse(any(nCells > 0), min(nCells[nCells > 0]), NA_real_),
      maxCellsPerDonor = ifelse(any(nCells > 0), max(nCells[nCells > 0]), NA_real_),
      .groups = "drop"
    ) %>%
    dplyr::select(compartment, dplyr::everything())
}

plotCoverageHeatmap <- function(coverageDf, title) {
  ggplot(coverageDf, aes(x = treatmentPhase, y = donorID, fill = nCells)) +
    geom_tile(color = "white") +
    geom_text(aes(label = nCells), size = 3.5) +
    scale_fill_viridis_c(option = "C", name = "# Cells") +
    labs(x = "Treatment Phase", y = "Donor", title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

classifyPeakByDistance <- function(distanceToTSS) {
  dplyr::case_when(
    is.na(distanceToTSS) ~ "Unassigned",
    distanceToTSS <= 2000 ~ "Promoter (<=2kb)",
    distanceToTSS <= 100000 ~ "Distal (2-100kb)",
    TRUE ~ "Distal (>100kb)"
  )
}

buildPeakAnnotationTable <- function(seuratObj) {
  peakNames <- rownames(seuratObj[["ATAC"]])
  peakGR <- tryCatch(StringToGRanges(peakNames), error = function(e) NULL)

  if (is.null(peakGR)) {
    return(data.frame(
      feature = peakNames,
      seqnames = NA_character_,
      start = NA_integer_,
      end = NA_integer_,
      width = NA_integer_,
      nearestGene = NA_character_,
      distanceToTSS = NA_real_,
      peakClass = "Unassigned",
      stringsAsFactors = FALSE
    ))
  }

  annotationDf <- data.frame(
    feature = peakNames,
    seqnames = as.character(GenomeInfoDb::seqnames(peakGR)),
    start = GenomicRanges::start(peakGR),
    end = GenomicRanges::end(peakGR),
    width = GenomicRanges::width(peakGR),
    stringsAsFactors = FALSE
  )

  nearestDf <- tryCatch(ClosestFeature(seuratObj, regions = peakGR), error = function(e) NULL)
  if (!is.null(nearestDf) && nrow(nearestDf) > 0) {
    matched <- match(annotationDf$feature, nearestDf$query_region)
    annotationDf$nearestGene <- nearestDf$gene_name[matched]
    annotationDf$distanceToTSS <- nearestDf$distance[matched]
  } else {
    annotationDf$nearestGene <- NA_character_
    annotationDf$distanceToTSS <- NA_real_
  }

  annotationDf$peakClass <- classifyPeakByDistance(annotationDf$distanceToTSS)
  annotationDf
}

ensureAtacMotifs <- function(seuratObj) {
  motifsPresent <- tryCatch(!is.null(seuratObj[["ATAC"]]@motifs), error = function(e) FALSE)
  if (motifsPresent) return(seuratObj)

  pfm <- TFBSTools::getMatrixSet(
    JASPAR2020::JASPAR2020,
    opts = list(collection = "CORE", tax_group = "vertebrates", all_versions = FALSE)
  )

  AddMotifs(
    object = seuratObj,
    genome = BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38,
    pfm = pfm
  )
}

buildMultimodalPeakGeneTable <- function(rnaTable, atacTable, peakAnnotation,
                                         maxDistance = 100000,
                                         pCutoff = 0.05,
                                         fcCutoff = 0.5) {
  atacAnnotated <- atacTable %>%
    dplyr::left_join(peakAnnotation, by = "feature") %>%
    dplyr::rename(
      ATAC_logFC = logFC,
      ATAC_FDR = adj.P.Val,
      ATAC_t = t
    )

  rnaCompact <- rnaTable %>%
    dplyr::select(feature, logFC, adj.P.Val, t) %>%
    dplyr::rename(
      nearestGene = feature,
      RNA_logFC = logFC,
      RNA_FDR = adj.P.Val,
      RNA_t = t
    )

  atacAnnotated %>%
    dplyr::filter(!is.na(nearestGene) & !is.na(distanceToTSS) & distanceToTSS <= maxDistance) %>%
    dplyr::inner_join(rnaCompact, by = "nearestGene") %>%
    dplyr::mutate(
      ATAC_sig = ATAC_FDR < pCutoff & abs(ATAC_logFC) > fcCutoff,
      RNA_sig = RNA_FDR < pCutoff & abs(RNA_logFC) > fcCutoff
    ) %>%
    dplyr::filter(ATAC_sig) %>%
    dplyr::mutate(
      multimodalClass = dplyr::case_when(
        RNA_sig & ATAC_logFC > 0 & RNA_logFC > 0 ~ "Concordant up",
        RNA_sig & ATAC_logFC < 0 & RNA_logFC < 0 ~ "Concordant down",
        RNA_sig ~ "Discordant",
        TRUE ~ "DA-only nearby gene"
      ),
      absCombinedLogFC = abs(ATAC_logFC) + abs(RNA_logFC)
    ) %>%
    dplyr::arrange(ATAC_FDR, RNA_FDR, dplyr::desc(absCombinedLogFC))
}
```

# Study Coverage and Primary Contrasts {.tabset}

## Global Donor x Phase Coverage

```{r globalCoverage, fig.height = 5, fig.width = 9}
globalCoverage <- list()
coverageSummary <- list()

for (comp in compartments) {
  coverageDf <- buildCoverageTable(seuratList[[comp]])
  globalCoverage[[comp]] <- coverageDf
  coverageSummary[[comp]] <- summarizeCoverageTable(coverageDf, comp)

  outCsv <- file.path(tableDir,
    paste0(filenameSuffix, "_", comp, "_global_donor_phase_coverage.csv"))
  write.csv(coverageDf, file = outCsv, row.names = FALSE)

  pCoverage <- plotCoverageHeatmap(
    coverageDf,
    paste(comp, "— Cells per Donor x Treatment Phase")
  )

  print(pCoverage)
  savePlot(
    plot = pCoverage,
    plotDir = plotDir,
    filename = paste0(comp, "_global_donor_phase_coverage"),
    height = 5,
    width = 9
  )

  cat(sprintf("\n%s coverage summary:\n", comp))
  print(coverageSummary[[comp]])
}

coverageSummaryDf <- bind_rows(coverageSummary)
write.csv(
  coverageSummaryDf,
  file.path(tableDir, paste0(filenameSuffix, "_global_donor_phase_coverage_summary.csv")),
  row.names = FALSE
)
```

## Primary Contrast Reference

```{r primaryContrastReference, results = "markup"}
primaryContrastReference <- tibble(
  contrast = primaryContrasts,
  comparison = vapply(primaryContrasts, function(nm) {
    paste(contrastDefs[[nm]], collapse = " vs ")
  }, character(1)),
  rationale = c(
    "Immediate low-dose IL-2 response versus baseline",
    "State after rapamycin exposure versus baseline",
    "Net followup remodeling versus baseline",
    "Resolution or persistence after rapamycin cessation"
  )
)

kable(primaryContrastReference, row.names = FALSE,
  caption = "Primary contrasts emphasized for PI-facing interpretation")

write.csv(
  primaryContrastReference,
  file.path(tableDir, paste0(filenameSuffix, "_primary_contrast_reference.csv")),
  row.names = FALSE
)
```

# Global Pseudobulk RNA DGE {.tabset}

## Aggregation and Model Fitting

```{r globalRNADGE}
rnaResults <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Global RNA DGE ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  # Get raw counts
  counts <- GetAssayData(seuratObj, layer = "counts")

  # Filter genes: keep those expressed in >= 30% of cells
  cellCount <- ncol(counts)
  geneFilter <- Matrix::rowSums(counts > 0) >= 0.3 * cellCount
  cat(sprintf("  Gene filter (>=30%% cells): %d / %d genes pass\n",
    sum(geneFilter), length(geneFilter)))
  counts <- counts[geneFilter, ]

  # Build metadata for aggregation
  meta <- seuratObj@meta.data[, c("donorID", "treatmentPhase")]

  # Aggregate by donor × treatment phase
  agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))
  cat(sprintf("  Pseudobulk groups: %d\n", ncol(agg$counts)))

  # Run limma-voom
  rnaResults[[comp]] <- runLimmaVoom(agg$counts, agg$designInfo, contrastDefs)

  if (!is.null(rnaResults[[comp]])) {
    for (nm in names(rnaResults[[comp]]$results)) {
      tt <- rnaResults[[comp]]$results[[nm]]
      nSig <- sum(tt$adj.P.Val < 0.05)
      nFC <- sum(tt$adj.P.Val < 0.05 & abs(tt$logFC) > 0.5)
      cat(sprintf("  %s: %d DE genes (FDR<0.05), %d with |logFC|>0.5\n", nm, nSig, nFC))
    }
  }
}
```

## RNA Volcano Plots

```{r rnaVolcanos, fig.height = 10, fig.width = 14}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next

  plots <- list()
  for (nm in names(rnaResults[[comp]]$results)) {
    plots[[nm]] <- plotVolcano(rnaResults[[comp]]$results[[nm]],
      paste(comp, "RNA:", nm), featureType = "Gene")
  }

  nPlots <- length(plots)
  nCols <- min(3, nPlots)
  nRows <- ceiling(nPlots / nCols)

  combined <- wrap_plots(plots, ncol = nCols)
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_RNA_DGE_volcanos"),
    height = 6 * nRows, width = 6.5 * nCols)
}
```

## RNA Results Tables

```{r rnaResultsTables}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next

  for (nm in names(rnaResults[[comp]]$results)) {
    tt <- rnaResults[[comp]]$results[[nm]]
    outFile <- file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_RNA_DGE_", nm, ".csv"))
    write.csv(tt, file = outFile, row.names = FALSE)
    cat(sprintf("Saved: %s\n", basename(outFile)))
  }
}
```

# Global Pseudobulk ATAC DA {.tabset}

## Aggregation and Model Fitting

```{r globalATACDA}
atacResults <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Global ATAC DA ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  # Get ATAC peak counts
  counts <- GetAssayData(seuratObj, layer = "counts")

  # Filter peaks: keep those accessible in >= 5% of cells
  # (lower threshold for ATAC since peaks are sparser than genes)
  cellCount <- ncol(counts)
  peakFilter <- Matrix::rowSums(counts > 0) >= 0.05 * cellCount
  cat(sprintf("  Peak filter (>=5%% cells): %d / %d peaks pass\n",
    sum(peakFilter), length(peakFilter)))
  counts <- counts[peakFilter, ]

  # Build metadata for aggregation
  meta <- seuratObj@meta.data[, c("donorID", "treatmentPhase")]

  # Aggregate by donor × treatment phase
  agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))
  cat(sprintf("  Pseudobulk groups: %d\n", ncol(agg$counts)))

  # Run limma-voom
  atacResults[[comp]] <- runLimmaVoom(agg$counts, agg$designInfo, contrastDefs)

  if (!is.null(atacResults[[comp]])) {
    for (nm in names(atacResults[[comp]]$results)) {
      tt <- atacResults[[comp]]$results[[nm]]
      nSig <- sum(tt$adj.P.Val < 0.05)
      nFC <- sum(tt$adj.P.Val < 0.05 & abs(tt$logFC) > 0.5)
      cat(sprintf("  %s: %d DA peaks (FDR<0.05), %d with |logFC|>0.5\n", nm, nSig, nFC))
    }
  }
}
```

## ATAC Volcano Plots

```{r atacVolcanos, fig.height = 10, fig.width = 14}
for (comp in compartments) {
  if (is.null(atacResults[[comp]])) next

  plots <- list()
  for (nm in names(atacResults[[comp]]$results)) {
    plots[[nm]] <- plotVolcano(atacResults[[comp]]$results[[nm]],
      paste(comp, "ATAC:", nm), featureType = "Peak")
  }

  nPlots <- length(plots)
  nCols <- min(3, nPlots)
  nRows <- ceiling(nPlots / nCols)

  combined <- wrap_plots(plots, ncol = nCols)
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_ATAC_DA_volcanos"),
    height = 6 * nRows, width = 7 * nCols)
}
```

## Peak Annotation

```{r atacPeakAnnotations}
peakAnnotations <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  peakAnnotations[[comp]] <- buildPeakAnnotationTable(seuratObj)

  outFile <- file.path(tableDir,
    paste0(filenameSuffix, "_", comp, "_ATAC_peak_annotations.csv"))
  write.csv(peakAnnotations[[comp]], file = outFile, row.names = FALSE)

  cat(sprintf("%s: peak annotation saved (%d peaks)\n",
    comp, nrow(peakAnnotations[[comp]])))
}
```

## Primary Contrast Motif Enrichment

```{r atacPrimaryMotifs, fig.height = 8, fig.width = 12}
motifResults <- list()
motifRequiredPkgs <- c("JASPAR2020", "TFBSTools", "BSgenome.Hsapiens.UCSC.hg38")
missingMotifPkgs <- motifRequiredPkgs[
  !vapply(motifRequiredPkgs, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missingMotifPkgs) > 0) {
  cat("Skipping motif enrichment; missing packages:",
    paste(missingMotifPkgs, collapse = ", "), "\n")
} else {
  for (comp in compartments) {
    if (is.null(atacResults[[comp]])) next

    seuratObj <- seuratList[[comp]]
    DefaultAssay(seuratObj) <- "ATAC"
    seuratObj <- ensureAtacMotifs(seuratObj)
    seuratList[[comp]] <- seuratObj
    motifResults[[comp]] <- list()

    plotDf <- data.frame()

    for (nm in intersect(primaryContrasts, names(atacResults[[comp]]$results))) {
      tt <- atacResults[[comp]]$results[[nm]]

      for (direction in c("Up", "Down")) {
        peakSubset <- tt %>%
          dplyr::filter(
            adj.P.Val < 0.05,
            if (direction == "Up") logFC > 0.5 else logFC < -0.5
          ) %>%
          dplyr::pull(feature)

        if (length(peakSubset) < 25) {
          cat(sprintf("%s %s %s: only %d DA peaks for motif enrichment — skipping\n",
            comp, nm, direction, length(peakSubset)))
          next
        }

        motifRes <- tryCatch(
          FindMotifs(seuratObj, features = peakSubset),
          error = function(e) {
            cat(sprintf("%s %s %s motif error: %s\n",
              comp, nm, direction, e$message))
            NULL
          }
        )

        if (is.null(motifRes) || nrow(motifRes) == 0) next

        motifRes$contrast <- nm
        motifRes$direction <- direction
        motifResults[[comp]][[paste(nm, direction, sep = "_")]] <- motifRes

        outFile <- file.path(tableDir,
          paste0(filenameSuffix, "_", comp, "_ATAC_motif_", nm, "_", direction, ".csv"))
        write.csv(motifRes, file = outFile, row.names = FALSE)

        topMotifs <- motifRes %>%
          arrange(pvalue) %>%
          head(5)
        topMotifs$motifLabel <- if ("motif.name" %in% colnames(topMotifs)) {
          topMotifs$motif.name
        } else {
          topMotifs$motif
        }
        topMotifs$contrast <- nm
        topMotifs$direction <- direction
        plotDf <- bind_rows(plotDf, topMotifs)
      }
    }

    if (nrow(plotDf) > 0) {
      pMotif <- ggplot(plotDf,
        aes(x = direction, y = reorder(motifLabel, fold.enrichment))) +
        geom_point(aes(size = fold.enrichment, color = -log10(pvalue))) +
        facet_wrap(~contrast, scales = "free_y") +
        scale_color_viridis_c(name = "-log10(P)") +
        scale_size_area(max_size = 10, name = "Fold enrichment") +
        labs(
          x = "Direction of DA peaks",
          y = "Motif",
          title = paste(comp, "— Primary Contrast ATAC Motif Enrichment")
        ) +
        theme(axis.text.y = element_text(size = 8))

      print(pMotif)
      savePlot(
        plot = pMotif,
        plotDir = plotDir,
        filename = paste0(comp, "_ATAC_primary_motif_enrichment"),
        height = 8,
        width = 12
      )
    }
  }
}
```

## ATAC Results Tables

```{r atacResultsTables}
for (comp in compartments) {
  if (is.null(atacResults[[comp]])) next

  for (nm in names(atacResults[[comp]]$results)) {
    tt <- atacResults[[comp]]$results[[nm]] %>%
      left_join(peakAnnotations[[comp]], by = "feature")
    outFile <- file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_ATAC_DA_", nm, ".csv"))
    write.csv(tt, file = outFile, row.names = FALSE)
    cat(sprintf("Saved: %s\n", basename(outFile)))
  }
}
```

# Per-Cluster RNA DGE {.tabset}

## Aggregation and Model Fitting

```{r perClusterRNADGE}
perClusterRNA <- list()
perClusterRNAStatus <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Per-Cluster RNA DGE ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  clusters <- sort(unique(seuratObj$wnn_clusters))
  perClusterRNA[[comp]] <- list()
  clusterStatus <- data.frame()

  for (cl in clusters) {
    cat(sprintf("\n  Cluster %s:\n", cl))

    # Subset to cluster
    cellIdx <- which(seuratObj$wnn_clusters == cl)
    statusRow <- data.frame(
      cluster = as.character(cl),
      nCells = length(cellIdx),
      nFeaturesPass = NA_integer_,
      nPhasesWithTwoDonors = NA_integer_,
      nPseudobulkGroups = NA_integer_,
      analyzed = FALSE,
      skipReason = NA_character_,
      stringsAsFactors = FALSE
    )

    if (length(cellIdx) < 20) {
      cat(sprintf("    Skipping: only %d cells\n", length(cellIdx)))
      statusRow$skipReason <- "lt20_cells"
      clusterStatus <- bind_rows(clusterStatus, statusRow)
      next
    }

    counts <- GetAssayData(seuratObj, layer = "counts")[, cellIdx]

    # Filter genes: >= 30% of cells in this cluster
    geneFilter <- Matrix::rowSums(counts > 0) >= 0.3 * length(cellIdx)
    counts <- counts[geneFilter, ]
    statusRow$nFeaturesPass <- nrow(counts)
    cat(sprintf("    %d cells, %d genes pass filter\n", length(cellIdx), nrow(counts)))

    if (nrow(counts) < 50) {
      cat("    Skipping: too few genes pass filter\n")
      statusRow$skipReason <- "lt50_genes"
      clusterStatus <- bind_rows(clusterStatus, statusRow)
      next
    }

    meta <- seuratObj@meta.data[cellIdx, c("donorID", "treatmentPhase")]

    # Check: need at least 2 treatment phases with >= 2 donors each
    phaseTab <- table(meta$treatmentPhase, meta$donorID)
    phasesWithData <- rowSums(phaseTab > 0) >= 2
    statusRow$nPhasesWithTwoDonors <- sum(phasesWithData)
    if (sum(phasesWithData) < 2) {
      cat("    Skipping: insufficient donor coverage across phases\n")
      statusRow$skipReason <- "insufficient_donor_phase_coverage"
      clusterStatus <- bind_rows(clusterStatus, statusRow)
      next
    }

    agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))
    statusRow$nPseudobulkGroups <- ncol(agg$counts)

    result <- tryCatch(
      runLimmaVoom(agg$counts, agg$designInfo, contrastDefs),
      error = function(e) {
        cat(sprintf("    Error: %s\n", e$message))
        NULL
      }
    )

    if (!is.null(result)) {
      perClusterRNA[[comp]][[as.character(cl)]] <- result
      statusRow$analyzed <- TRUE
      for (nm in names(result$results)) {
        nSig <- sum(result$results[[nm]]$adj.P.Val < 0.05)
        cat(sprintf("    %s: %d DE genes\n", nm, nSig))
      }
    } else {
      statusRow$skipReason <- "limma_error"
    }

    clusterStatus <- bind_rows(clusterStatus, statusRow)
  }

  perClusterRNAStatus[[comp]] <- clusterStatus
}
```

## Per-Cluster RNA Summary Heatmap

```{r perClusterRNASummary, fig.height = 6, fig.width = 10}
for (comp in compartments) {
  if (length(perClusterRNA[[comp]]) == 0) next

  # Count significant genes per cluster × contrast
  summaryDf <- do.call(rbind, lapply(names(perClusterRNA[[comp]]), function(cl) {
    do.call(rbind, lapply(names(perClusterRNA[[comp]][[cl]]$results), function(nm) {
      tt <- perClusterRNA[[comp]][[cl]]$results[[nm]]
      data.frame(
        cluster = cl,
        contrast = nm,
        nDE = sum(tt$adj.P.Val < 0.05),
        nUp = sum(tt$adj.P.Val < 0.05 & tt$logFC > 0),
        nDown = sum(tt$adj.P.Val < 0.05 & tt$logFC < 0)
      )
    }))
  }))

  pSummary <- ggplot(summaryDf, aes(x = contrast, y = cluster, fill = nDE)) +
    geom_tile(color = "white") +
    geom_text(aes(label = nDE), size = 3.5) +
    scale_fill_gradient(low = "white", high = "#E41A1C", name = "# DE Genes") +
    labs(x = "", y = "WNN Cluster",
      title = paste(comp, "— Per-Cluster DE Genes (RNA, FDR < 0.05)")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pSummary)

  savePlot(plot = pSummary, plotDir = plotDir,
    filename = paste0(comp, "_perCluster_RNA_DGE_summary"),
    height = 6, width = 10)
}
```

## Per-Cluster RNA Tables

```{r perClusterRNATables}
for (comp in compartments) {
  for (cl in names(perClusterRNA[[comp]])) {
    for (nm in names(perClusterRNA[[comp]][[cl]]$results)) {
      tt <- perClusterRNA[[comp]][[cl]]$results[[nm]]
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_cluster", cl, "_RNA_DGE_", nm, ".csv"))
      write.csv(tt, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: per-cluster RNA tables saved\n", comp))
}
```

# Per-Cluster ATAC DA {.tabset}

## Aggregation and Model Fitting

```{r perClusterATACDA}
perClusterATAC <- list()
perClusterATACStatus <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Per-Cluster ATAC DA ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  clusters <- sort(unique(seuratObj$wnn_clusters))
  perClusterATAC[[comp]] <- list()
  clusterStatus <- data.frame()

  for (cl in clusters) {
    cat(sprintf("\n  Cluster %s:\n", cl))

    cellIdx <- which(seuratObj$wnn_clusters == cl)
    statusRow <- data.frame(
      cluster = as.character(cl),
      nCells = length(cellIdx),
      nFeaturesPass = NA_integer_,
      nPhasesWithTwoDonors = NA_integer_,
      nPseudobulkGroups = NA_integer_,
      analyzed = FALSE,
      skipReason = NA_character_,
      stringsAsFactors = FALSE
    )

    if (length(cellIdx) < 20) {
      cat(sprintf("    Skipping: only %d cells\n", length(cellIdx)))
      statusRow$skipReason <- "lt20_cells"
      clusterStatus <- bind_rows(clusterStatus, statusRow)
      next
    }

    counts <- GetAssayData(seuratObj, layer = "counts")[, cellIdx]

    # Filter peaks: >= 5% of cells in this cluster
    peakFilter <- Matrix::rowSums(counts > 0) >= 0.05 * length(cellIdx)
    counts <- counts[peakFilter, ]
    statusRow$nFeaturesPass <- nrow(counts)
    cat(sprintf("    %d cells, %d peaks pass filter\n", length(cellIdx), nrow(counts)))

    if (nrow(counts) < 50) {
      cat("    Skipping: too few peaks pass filter\n")
      statusRow$skipReason <- "lt50_peaks"
      clusterStatus <- bind_rows(clusterStatus, statusRow)
      next
    }

    meta <- seuratObj@meta.data[cellIdx, c("donorID", "treatmentPhase")]

    phaseTab <- table(meta$treatmentPhase, meta$donorID)
    phasesWithData <- rowSums(phaseTab > 0) >= 2
    statusRow$nPhasesWithTwoDonors <- sum(phasesWithData)
    if (sum(phasesWithData) < 2) {
      cat("    Skipping: insufficient donor coverage across phases\n")
      statusRow$skipReason <- "insufficient_donor_phase_coverage"
      clusterStatus <- bind_rows(clusterStatus, statusRow)
      next
    }

    agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))
    statusRow$nPseudobulkGroups <- ncol(agg$counts)

    result <- tryCatch(
      runLimmaVoom(agg$counts, agg$designInfo, contrastDefs),
      error = function(e) {
        cat(sprintf("    Error: %s\n", e$message))
        NULL
      }
    )

    if (!is.null(result)) {
      perClusterATAC[[comp]][[as.character(cl)]] <- result
      statusRow$analyzed <- TRUE
      for (nm in names(result$results)) {
        nSig <- sum(result$results[[nm]]$adj.P.Val < 0.05)
        cat(sprintf("    %s: %d DA peaks\n", nm, nSig))
      }
    } else {
      statusRow$skipReason <- "limma_error"
    }

    clusterStatus <- bind_rows(clusterStatus, statusRow)
  }

  perClusterATACStatus[[comp]] <- clusterStatus
}
```

## Per-Cluster ATAC Summary Heatmap

```{r perClusterATACSummary, fig.height = 6, fig.width = 10}
for (comp in compartments) {
  if (length(perClusterATAC[[comp]]) == 0) next

  summaryDf <- do.call(rbind, lapply(names(perClusterATAC[[comp]]), function(cl) {
    do.call(rbind, lapply(names(perClusterATAC[[comp]][[cl]]$results), function(nm) {
      tt <- perClusterATAC[[comp]][[cl]]$results[[nm]]
      data.frame(
        cluster = cl,
        contrast = nm,
        nDA = sum(tt$adj.P.Val < 0.05),
        nUp = sum(tt$adj.P.Val < 0.05 & tt$logFC > 0),
        nDown = sum(tt$adj.P.Val < 0.05 & tt$logFC < 0)
      )
    }))
  }))

  pSummary <- ggplot(summaryDf, aes(x = contrast, y = cluster, fill = nDA)) +
    geom_tile(color = "white") +
    geom_text(aes(label = nDA), size = 3.5) +
    scale_fill_gradient(low = "white", high = "#FF7F00", name = "# DA Peaks") +
    labs(x = "", y = "WNN Cluster",
      title = paste(comp, "— Per-Cluster DA Peaks (ATAC, FDR < 0.05)")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pSummary)

  savePlot(plot = pSummary, plotDir = plotDir,
    filename = paste0(comp, "_perCluster_ATAC_DA_summary"),
    height = 6, width = 10)
}
```

## Per-Cluster ATAC Tables

```{r perClusterATACTables}
for (comp in compartments) {
  for (cl in names(perClusterATAC[[comp]])) {
    for (nm in names(perClusterATAC[[comp]][[cl]]$results)) {
      tt <- perClusterATAC[[comp]][[cl]]$results[[nm]] %>%
        left_join(peakAnnotations[[comp]], by = "feature")
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_cluster", cl, "_ATAC_DA_", nm, ".csv"))
      write.csv(tt, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: per-cluster ATAC tables saved\n", comp))
}
```

## Per-Cluster Coverage Status

```{r perClusterCoverageStatus, fig.height = 6, fig.width = 10}
for (comp in compartments) {
  statusDf <- perClusterRNAStatus[[comp]] %>%
    dplyr::mutate(analysis = "RNA") %>%
    bind_rows(perClusterATACStatus[[comp]] %>% dplyr::mutate(analysis = "ATAC")) %>%
    dplyr::mutate(
      skipReason = ifelse(is.na(skipReason), "Analyzed", skipReason),
      cluster = factor(cluster, levels = unique(cluster))
    )

  outFile <- file.path(tableDir,
    paste0(filenameSuffix, "_", comp, "_perCluster_analysis_status.csv"))
  write.csv(statusDf, file = outFile, row.names = FALSE)

  pStatus <- ggplot(statusDf, aes(x = cluster, y = analysis, fill = skipReason)) +
    geom_tile(color = "white") +
    geom_text(aes(label = nCells), size = 3.5) +
    scale_fill_brewer(palette = "Set2", name = "Status") +
    labs(
      x = "WNN Cluster",
      y = "Analysis",
      title = paste(comp, "— Per-Cluster Analysis Status"),
      subtitle = "Tile labels show cells per cluster"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pStatus)
  savePlot(
    plot = pStatus,
    plotDir = plotDir,
    filename = paste0(comp, "_perCluster_analysis_status"),
    height = 6,
    width = 10
  )
}
```

# GSEA on Global RNA DGE {.tabset}

## Load Gene Sets

```{r loadGeneSets}
hallmarkSets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

goSets <- msigdbr(species = "Homo sapiens", category = "C5",
  subcategory = "GO:BP") %>%
  dplyr::select(gs_name, gene_symbol)

c7Sets <- msigdbr(species = "Homo sapiens", category = "C7") %>%
  dplyr::select(gs_name, gene_symbol)

# Extract IL2-STAT5 signaling gene set for focused roast analysis
il2Stat5Genes <- hallmarkSets %>%
  dplyr::filter(gs_name == "HALLMARK_IL2_STAT5_SIGNALING") %>%
  dplyr::pull(gene_symbol) %>%
  unique()

cat(sprintf("Hallmark: %d gene sets\n", length(unique(hallmarkSets$gs_name))))
cat(sprintf("GO:BP: %d gene sets\n", length(unique(goSets$gs_name))))
cat(sprintf("C7 Immunologic: %d gene sets\n", length(unique(c7Sets$gs_name))))
cat(sprintf("IL2-STAT5 signaling: %d genes\n", length(il2Stat5Genes)))
```

## Run GSEA

```{r runGSEA}
gseaResults <- list()

for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next
  gseaResults[[comp]] <- list()

  for (nm in names(rnaResults[[comp]]$results)) {
    tt <- rnaResults[[comp]]$results[[nm]]

    # Rank by t-statistic (more stable than logFC for GSEA)
    geneRanks <- tt$t
    names(geneRanks) <- tt$feature
    geneRanks <- sort(geneRanks, decreasing = TRUE)

    # Remove duplicates and NAs
    geneRanks <- geneRanks[!is.na(geneRanks)]
    geneRanks <- geneRanks[!duplicated(names(geneRanks))]

    cat(sprintf("\n%s %s: GSEA with %d ranked genes\n", comp, nm, length(geneRanks)))

    # Hallmark GSEA
    hallmarkGSEA <- tryCatch(
      GSEA(geneRanks, TERM2GENE = hallmarkSets,
        minGSSize = 10, maxGSSize = 500,
        pvalueCutoff = 1.0, verbose = FALSE),
      error = function(e) {
        cat(sprintf("  Hallmark error: %s\n", e$message))
        NULL
      }
    )

    # GO:BP GSEA
    goGSEA <- tryCatch(
      GSEA(geneRanks, TERM2GENE = goSets,
        minGSSize = 10, maxGSSize = 500,
        pvalueCutoff = 1.0, verbose = FALSE),
      error = function(e) {
        cat(sprintf("  GO:BP error: %s\n", e$message))
        NULL
      }
    )

    gseaResults[[comp]][[nm]] <- list(
      hallmark = hallmarkGSEA,
      go_bp = goGSEA
    )

    if (!is.null(hallmarkGSEA)) {
      nSigH <- sum(hallmarkGSEA@result$p.adjust < 0.05)
      cat(sprintf("  Hallmark: %d significant (FDR<0.05)\n", nSigH))
    }
    if (!is.null(goGSEA)) {
      nSigG <- sum(goGSEA@result$p.adjust < 0.05)
      cat(sprintf("  GO:BP: %d significant (FDR<0.05)\n", nSigG))
    }
  }
}
```

## GSEA Bubble Plots

```{r gseaBubblePlots, fig.height = 10, fig.width = 12}
for (comp in compartments) {
  if (is.null(gseaResults[[comp]])) next

  for (nm in names(gseaResults[[comp]])) {
    gseaSets <- gseaResults[[comp]][[nm]]
    plotData <- list()

    # Hallmark results
    if (!is.null(gseaSets$hallmark) && nrow(gseaSets$hallmark@result) > 0) {
      hRes <- gseaSets$hallmark@result %>%
        dplyr::filter(p.adjust < 0.1) %>%
        mutate(GeneSet = "Hallmark",
          Description_clean = gsub("^HALLMARK_", "", Description),
          Description_clean = gsub("_", " ", Description_clean),
          Description_clean = str_to_title(Description_clean))

      if (nrow(hRes) > 0) plotData[["Hallmark"]] <- hRes
    }

    # GO:BP results (top 20 by p.adjust)
    if (!is.null(gseaSets$go_bp) && nrow(gseaSets$go_bp@result) > 0) {
      gRes <- gseaSets$go_bp@result %>%
        dplyr::filter(p.adjust < 0.1) %>%
        arrange(p.adjust) %>%
        head(20) %>%
        mutate(GeneSet = "GO:BP",
          Description_clean = gsub("^GOBP_", "", Description),
          Description_clean = gsub("_", " ", Description_clean),
          Description_clean = str_to_title(Description_clean))

      if (nrow(gRes) > 0) plotData[["GO:BP"]] <- gRes
    }

    if (length(plotData) == 0) {
      cat(sprintf("%s %s: No significant GSEA results to plot\n", comp, nm))
      next
    }

    combined <- bind_rows(plotData)

    pBubble <- ggplot(combined, aes(x = NES, y = reorder(Description_clean, NES))) +
      geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
      geom_segment(aes(y = reorder(Description_clean, NES),
        yend = reorder(Description_clean, NES),
        x = 0, xend = NES), color = "black", linewidth = 0.5) +
      geom_point(aes(size = setSize, fill = -log10(p.adjust)),
        shape = 21, color = "black", stroke = 0.5) +
      facet_grid(GeneSet ~ ., scales = "free_y", space = "free_y") +
      scale_fill_gradient(low = "lightblue", high = "red",
        name = "-log10(FDR)") +
      scale_size_area(max_size = 8,
        guide = guide_legend(override.aes = list(fill = NA, color = "black"))) +
      labs(x = "Normalized Enrichment Score (NES)", y = "",
        size = "Gene Set Size",
        title = paste(comp, "—", nm)) +
      theme(
        strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
        strip.text = element_text(face = "bold", size = 12),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y = element_text(size = 10),
        panel.grid.major.x = element_line(color = "gray95"),
        panel.grid.major.y = element_blank())

    print(pBubble)

    savePlot(plot = pBubble, plotDir = plotDir,
      filename = paste0(comp, "_GSEA_", nm),
      height = max(6, 0.4 * nrow(combined) + 2), width = 12)
  }
}
```

## GSEA Results Tables

```{r gseaResultsTables}
for (comp in compartments) {
  if (is.null(gseaResults[[comp]])) next

  for (nm in names(gseaResults[[comp]])) {
    gseaSets <- gseaResults[[comp]][[nm]]

    for (setType in c("hallmark", "go_bp")) {
      if (is.null(gseaSets[[setType]])) next
      res <- gseaSets[[setType]]@result
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_GSEA_", setType, "_", nm, ".csv"))
      write.csv(res, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: GSEA tables saved\n", comp))
}
```

# Focused Roast: IL2-STAT5 Signaling {.tabset}

## Global Roast

```{r roastIL2STAT5, fig.height = 8, fig.width = 8}
roastPlotDir <- file.path(plotDir, "roast_IL2STAT5")
if (!dir.exists(roastPlotDir)) dir.create(roastPlotDir, recursive = TRUE)

roastResults <- list()
roastSummary <- data.frame()

for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next
  roastResults[[comp]] <- list()

  v <- rnaResults[[comp]]$voom
  design <- rnaResults[[comp]]$design
  contMatrix <- rnaResults[[comp]]$contMatrix

  # Gene index for IL2-STAT5
  idx <- which(rownames(v) %in% il2Stat5Genes)
  cat(sprintf("\n%s: %d / %d IL2-STAT5 genes found in voom object\n",
    comp, length(idx), length(il2Stat5Genes)))

  if (length(idx) < 10) {
    cat("  Too few genes overlap — skipping roast\n")
    next
  }

  for (nm in colnames(contMatrix)) {
    roastObj <- roast(y = v, index = idx, design = design,
      contrast = contMatrix[, nm], nrot = 9999)
    roastPV <- roastObj$p.value  # data.frame with rows Down/Up/UpOrDown/Mixed

    roastResults[[comp]][[nm]] <- roastPV

    roastSummary <- rbind(roastSummary, data.frame(
      compartment = comp,
      contrast = nm,
      NGenes = length(idx),
      PropDown = roastPV["Down", "Active.Prop"],
      PropUp = roastPV["Up", "Active.Prop"],
      Direction_Up_PValue = roastPV["Up", "P.Value"],
      Direction_Down_PValue = roastPV["Down", "P.Value"],
      Mixed_PValue = roastPV["Mixed", "P.Value"]
    ))

    cat(sprintf("  %s: Mixed P = %.4f, Up P = %.4f, Down P = %.4f\n",
      nm, roastPV["Mixed", "P.Value"],
      roastPV["Up", "P.Value"], roastPV["Down", "P.Value"]))
  }
}
```

## Global Volcano + Barcode Plots

```{r roastIL2VolcanoBarcode, fig.height = 8, fig.width = 8}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]]) || is.null(roastResults[[comp]])) next

  v <- rnaResults[[comp]]$voom
  fit2 <- rnaResults[[comp]]$fit
  idx <- which(rownames(v) %in% il2Stat5Genes)

  for (nm in names(roastResults[[comp]])) {
    tt <- rnaResults[[comp]]$results[[nm]]
    roastRes <- roastResults[[comp]][[nm]]

    # Classify genes
    tt$inSet <- tt$feature %in% il2Stat5Genes
    tt$highlight <- ifelse(tt$inSet, "IL2-STAT5", "Other")
    tt$highlight <- factor(tt$highlight, levels = c("IL2-STAT5", "Other"))

    # Top IL2-STAT5 genes to label (by significance within the set)
    topSetGenes <- tt %>%
      dplyr::filter(inSet) %>%
      arrange(P.Value) %>%
      head(10)

    # Volcano panel
    pVolcano <- ggplot(tt, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = dplyr::filter(tt, !inSet),
        color = "grey80", size = 0.6, alpha = 0.5) +
      geom_point(data = dplyr::filter(tt, inSet),
        color = "#8B0000", size = 1.5, alpha = 0.8) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
      geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey50") +
      geom_text_repel(data = topSetGenes,
        aes(label = feature), size = 3, color = "#8B0000",
        max.overlaps = 20, fontface = "italic") +
      labs(x = NULL, y = expression(-log[10](P[adj])),
        title = paste(comp, "—", nm),
        subtitle = "IL2-STAT5 Signaling genes highlighted") +
      theme(plot.margin = margin(5, 5, 0, 5))

    # Barcode panel — rug strip showing logFC distribution
    # All genes as light ticks, IL2-STAT5 as dark red ticks
    allGenes <- data.frame(logFC = tt$logFC, inSet = tt$inSet)
    bgGenes <- dplyr::filter(allGenes, !inSet)
    setGenes <- dplyr::filter(allGenes, inSet)

    # Roast annotation text
    roastPV <- roastResults[[comp]][[nm]]  # already the $p.value data.frame
    mixedP <- roastPV["Mixed", "P.Value"]
    upP <- roastPV["Up", "P.Value"]
    downP <- roastPV["Down", "P.Value"]
    bestDir <- ifelse(upP < downP, "Up", "Down")
    bestP <- min(upP, downP)
    annotText <- sprintf("Roast: P = %.2e (%s)", bestP, bestDir)

    pBarcode <- ggplot() +
      geom_segment(data = bgGenes,
        aes(x = logFC, xend = logFC, y = 0, yend = 0.4),
        color = "grey80", linewidth = 0.15, alpha = 0.3) +
      geom_segment(data = setGenes,
        aes(x = logFC, xend = logFC, y = 0, yend = 1),
        color = "#8B0000", linewidth = 0.4, alpha = 0.8) +
      annotate("text", x = max(tt$logFC) * 0.7, y = 0.5,
        label = annotText, size = 3.5, hjust = 0.5) +
      scale_y_continuous(breaks = NULL, limits = c(0, 1.1)) +
      labs(x = expression(log[2] ~ fold ~ change), y = "GSEA") +
      theme_minimal(base_size = 12) +
      theme(
        panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 10),
        plot.margin = margin(0, 5, 5, 5))

    # Combine with patchwork
    combined <- pVolcano / pBarcode + plot_layout(heights = c(5, 1))
    print(combined)

    savePlot(plot = combined, plotDir = roastPlotDir,
      filename = paste0(comp, "_roast_IL2STAT5_volcano_", nm),
      height = 8, width = 8)
  }
}
```

## Global Roast Summary Table

```{r roastIL2Table, results = "markup"}
if (nrow(roastSummary) > 0) {
  kable(roastSummary, digits = 4, row.names = FALSE,
    caption = "Roast test of HALLMARK_IL2_STAT5_SIGNALING across contrasts")

  outFile <- file.path(tableDir,
    paste0(filenameSuffix, "_roast_IL2STAT5_global_summary.csv"))
  write.csv(roastSummary, file = outFile, row.names = FALSE)
  cat(sprintf("Saved: %s\n", basename(outFile)))
}
```

## Per-Cluster Roast

```{r roastIL2PerCluster}
roastPerCluster <- list()
roastPerClusterSummary <- data.frame()

for (comp in compartments) {
  if (length(perClusterRNA[[comp]]) == 0) next
  roastPerCluster[[comp]] <- list()

  for (cl in names(perClusterRNA[[comp]])) {
    clResult <- perClusterRNA[[comp]][[cl]]
    v <- clResult$voom
    design <- clResult$design
    contMatrix <- clResult$contMatrix

    idx <- which(rownames(v) %in% il2Stat5Genes)
    if (length(idx) < 5) {
      cat(sprintf("%s cluster %s: only %d IL2-STAT5 genes — skipping\n",
        comp, cl, length(idx)))
      next
    }

    roastPerCluster[[comp]][[cl]] <- list()

    for (nm in colnames(contMatrix)) {
      roastPV <- tryCatch(
        {
          roastObj <- roast(y = v, index = idx, design = design,
            contrast = contMatrix[, nm], nrot = 9999)
          roastObj$p.value
        },
        error = function(e) {
          cat(sprintf("  %s cl%s %s error: %s\n",
            comp, cl, nm, e$message))
          NULL
        }
      )

      if (is.null(roastPV)) next
      roastPerCluster[[comp]][[cl]][[nm]] <- roastPV

      roastPerClusterSummary <- rbind(roastPerClusterSummary, data.frame(
        compartment = comp,
        cluster = cl,
        contrast = nm,
        NGenes = length(idx),
        PropDown = roastPV["Down", "Active.Prop"],
        PropUp = roastPV["Up", "Active.Prop"],
        Direction_Up_PValue = roastPV["Up", "P.Value"],
        Direction_Down_PValue = roastPV["Down", "P.Value"],
        Mixed_PValue = roastPV["Mixed", "P.Value"]
      ))
    }

    cat(sprintf("%s cluster %s: roast complete (%d genes)\n", comp, cl, length(idx)))
  }
}
```

## Per-Cluster Roast Heatmap

```{r roastIL2PerClusterHeatmap, fig.height = 8, fig.width = 12}
for (comp in compartments) {
  compSummary <- roastPerClusterSummary %>%
    dplyr::filter(compartment == comp)
  if (nrow(compSummary) == 0) next

  # Determine direction for annotation
  compSummary <- compSummary %>%
    mutate(
      bestP = pmin(Direction_Up_PValue, Direction_Down_PValue),
      direction = ifelse(Direction_Up_PValue < Direction_Down_PValue, "Up", "Down"),
      dirLabel = case_when(
        bestP < 0.01 ~ ifelse(direction == "Up", "\u2191\u2191", "\u2193\u2193"),
        bestP < 0.05 ~ ifelse(direction == "Up", "\u2191", "\u2193"),
        bestP < 0.1  ~ ifelse(direction == "Up", "\u2191?", "\u2193?"),
        TRUE ~ ""
      ),
      negLog10P = -log10(Mixed_PValue)
    )

  pHeat <- ggplot(compSummary, aes(x = contrast, y = cluster)) +
    geom_tile(aes(fill = negLog10P), color = "white") +
    geom_text(aes(label = paste0(dirLabel, "\n",
      sprintf("P=%.3f", Mixed_PValue))), size = 3) +
    scale_fill_gradient(low = "white", high = "#8B0000",
      name = expression(-log[10](P))) +
    labs(x = "", y = "WNN Cluster",
      title = paste(comp, "— Per-Cluster Roast: IL2-STAT5 Signaling")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pHeat)

  savePlot(plot = pHeat, plotDir = roastPlotDir,
    filename = paste0(comp, "_roast_IL2STAT5_perCluster_heatmap"),
    height = 8, width = 12)
}
```

## Per-Cluster Roast Table

```{r roastIL2PerClusterTable, results = "markup"}
if (nrow(roastPerClusterSummary) > 0) {
  kable(roastPerClusterSummary, digits = 4, row.names = FALSE,
    caption = "Per-cluster Roast: HALLMARK_IL2_STAT5_SIGNALING")

  outFile <- file.path(tableDir,
    paste0(filenameSuffix, "_roast_IL2STAT5_perCluster_summary.csv"))
  write.csv(roastPerClusterSummary, file = outFile, row.names = FALSE)
  cat(sprintf("Saved: %s\n", basename(outFile)))
}
```

# GSEA: C7 Immunologic Signatures {.tabset}

## Global C7 GSEA

```{r runC7GSEA}
c7PlotDir <- file.path(plotDir, "C7_GSEA")
if (!dir.exists(c7PlotDir)) dir.create(c7PlotDir, recursive = TRUE)

c7GseaResults <- list()

for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next
  c7GseaResults[[comp]] <- list()

  for (nm in names(rnaResults[[comp]]$results)) {
    tt <- rnaResults[[comp]]$results[[nm]]

    geneRanks <- tt$t
    names(geneRanks) <- tt$feature
    geneRanks <- sort(geneRanks, decreasing = TRUE)
    geneRanks <- geneRanks[!is.na(geneRanks)]
    geneRanks <- geneRanks[!duplicated(names(geneRanks))]

    cat(sprintf("\n%s %s: C7 GSEA with %d ranked genes\n",
      comp, nm, length(geneRanks)))

    c7GSEA <- tryCatch(
      GSEA(geneRanks, TERM2GENE = c7Sets,
        minGSSize = 10, maxGSSize = 500,
        pvalueCutoff = 1.0, verbose = FALSE),
      error = function(e) {
        cat(sprintf("  C7 error: %s\n", e$message))
        NULL
      }
    )

    c7GseaResults[[comp]][[nm]] <- c7GSEA

    if (!is.null(c7GSEA)) {
      nSig <- sum(c7GSEA@result$p.adjust < 0.05)
      cat(sprintf("  C7: %d significant (FDR<0.05)\n", nSig))
    }
  }
}
```

## Global C7 Bubble Plots

```{r c7BubblePlots, fig.height = 14, fig.width = 14}
for (comp in compartments) {
  if (is.null(c7GseaResults[[comp]])) next

  for (nm in names(c7GseaResults[[comp]])) {
    c7Res <- c7GseaResults[[comp]][[nm]]
    if (is.null(c7Res) || nrow(c7Res@result) == 0) next

    plotDf <- c7Res@result %>%
      dplyr::filter(p.adjust < 0.1) %>%
      arrange(p.adjust) %>%
      head(30) %>%
      mutate(
        Description_clean = gsub("^GSE\\d+_", "", Description),
        Description_clean = gsub("_", " ", Description_clean),
        Description_clean = str_to_title(Description_clean),
        Description_clean = str_wrap(Description_clean, width = 50))

    if (nrow(plotDf) == 0) {
      cat(sprintf("%s %s: No significant C7 results to plot\n", comp, nm))
      next
    }

    pBubble <- ggplot(plotDf, aes(x = NES,
      y = reorder(Description_clean, NES))) +
      geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
      geom_segment(aes(y = reorder(Description_clean, NES),
        yend = reorder(Description_clean, NES),
        x = 0, xend = NES), color = "black", linewidth = 0.5) +
      geom_point(aes(size = setSize, fill = -log10(p.adjust)),
        shape = 21, color = "black", stroke = 0.5) +
      scale_fill_gradient(low = "lightblue", high = "red",
        name = "-log10(FDR)") +
      scale_size_area(max_size = 8,
        guide = guide_legend(override.aes = list(fill = NA, color = "black"))) +
      labs(x = "Normalized Enrichment Score (NES)", y = "",
        size = "Gene Set Size",
        title = paste(comp, "— C7 Immunologic:", nm)) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y = element_text(size = 9),
        panel.grid.major.x = element_line(color = "gray95"),
        panel.grid.major.y = element_blank())

    print(pBubble)

    savePlot(plot = pBubble, plotDir = c7PlotDir,
      filename = paste0(comp, "_C7_GSEA_", nm),
      height = max(8, 0.5 * nrow(plotDf) + 3), width = 14)
  }
}
```

## Global C7 Results Tables

```{r c7ResultsTables}
for (comp in compartments) {
  if (is.null(c7GseaResults[[comp]])) next

  for (nm in names(c7GseaResults[[comp]])) {
    c7Res <- c7GseaResults[[comp]][[nm]]
    if (is.null(c7Res)) next
    res <- c7Res@result
    outFile <- file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_GSEA_C7_", nm, ".csv"))
    write.csv(res, file = outFile, row.names = FALSE)
  }
  cat(sprintf("%s: C7 GSEA tables saved\n", comp))
}
```

## Per-Cluster C7 GSEA

```{r runC7GSEAPerCluster}
c7GseaPerCluster <- list()

for (comp in compartments) {
  if (length(perClusterRNA[[comp]]) == 0) next
  c7GseaPerCluster[[comp]] <- list()

  for (cl in names(perClusterRNA[[comp]])) {
    clResult <- perClusterRNA[[comp]][[cl]]
    c7GseaPerCluster[[comp]][[cl]] <- list()

    for (nm in names(clResult$results)) {
      tt <- clResult$results[[nm]]

      geneRanks <- tt$t
      names(geneRanks) <- tt$feature
      geneRanks <- sort(geneRanks, decreasing = TRUE)
      geneRanks <- geneRanks[!is.na(geneRanks)]
      geneRanks <- geneRanks[!duplicated(names(geneRanks))]

      if (length(geneRanks) < 50) next

      c7Res <- tryCatch(
        GSEA(geneRanks, TERM2GENE = c7Sets,
          minGSSize = 10, maxGSSize = 500,
          pvalueCutoff = 1.0, verbose = FALSE),
        error = function(e) NULL
      )

      c7GseaPerCluster[[comp]][[cl]][[nm]] <- c7Res

      if (!is.null(c7Res)) {
        nSig <- sum(c7Res@result$p.adjust < 0.05)
        cat(sprintf("%s cluster %s %s: %d significant C7 terms\n",
          comp, cl, nm, nSig))
      }
    }
  }
}
```

## Per-Cluster C7 Bubble Plots

```{r c7PerClusterBubblePlots, fig.height = 12, fig.width = 14}
for (comp in compartments) {
  if (is.null(c7GseaPerCluster[[comp]])) next

  for (cl in names(c7GseaPerCluster[[comp]])) {
    for (nm in names(c7GseaPerCluster[[comp]][[cl]])) {
      c7Res <- c7GseaPerCluster[[comp]][[cl]][[nm]]
      if (is.null(c7Res) || nrow(c7Res@result) == 0) next

      plotDf <- c7Res@result %>%
        dplyr::filter(p.adjust < 0.1) %>%
        arrange(p.adjust) %>%
        head(20) %>%
        mutate(
          Description_clean = gsub("^GSE\\d+_", "", Description),
          Description_clean = gsub("_", " ", Description_clean),
          Description_clean = str_to_title(Description_clean),
          Description_clean = str_wrap(Description_clean, width = 50))

      if (nrow(plotDf) == 0) next

      pBubble <- ggplot(plotDf, aes(x = NES,
        y = reorder(Description_clean, NES))) +
        geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
        geom_segment(aes(y = reorder(Description_clean, NES),
          yend = reorder(Description_clean, NES),
          x = 0, xend = NES), color = "black", linewidth = 0.5) +
        geom_point(aes(size = setSize, fill = -log10(p.adjust)),
          shape = 21, color = "black", stroke = 0.5) +
        scale_fill_gradient(low = "lightblue", high = "red",
          name = "-log10(FDR)") +
        scale_size_area(max_size = 8,
          guide = guide_legend(override.aes = list(fill = NA, color = "black"))) +
        labs(x = "Normalized Enrichment Score (NES)", y = "",
          size = "Gene Set Size",
          title = paste(comp, "cluster", cl, "— C7:", nm)) +
        theme(
          plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
          axis.text.y = element_text(size = 9),
          panel.grid.major.x = element_line(color = "gray95"),
          panel.grid.major.y = element_blank())

      print(pBubble)

      savePlot(plot = pBubble, plotDir = c7PlotDir,
        filename = paste0(comp, "_cluster", cl, "_C7_GSEA_", nm),
        height = max(8, 0.5 * nrow(plotDf) + 3), width = 14)
    }
  }
}
```

## Per-Cluster C7 Results Tables

```{r c7PerClusterTables}
for (comp in compartments) {
  if (is.null(c7GseaPerCluster[[comp]])) next

  for (cl in names(c7GseaPerCluster[[comp]])) {
    for (nm in names(c7GseaPerCluster[[comp]][[cl]])) {
      c7Res <- c7GseaPerCluster[[comp]][[cl]][[nm]]
      if (is.null(c7Res)) next
      res <- c7Res@result
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_cluster", cl, "_GSEA_C7_", nm, ".csv"))
      write.csv(res, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: per-cluster C7 GSEA tables saved\n", comp))
}
```

# Cross-Modality Integration Summary {.tabset}

## DA-DGE Overlap

```{r daDgeOverlap, fig.height = 6, fig.width = 8}
# For each contrast, identify peaks near DE genes and vice versa
for (comp in compartments) {
  if (is.null(rnaResults[[comp]]) || is.null(atacResults[[comp]])) next

  overlapSummary <- data.frame()

  for (nm in intersect(names(rnaResults[[comp]]$results),
    names(atacResults[[comp]]$results))) {
    rna <- rnaResults[[comp]]$results[[nm]]
    atac <- atacResults[[comp]]$results[[nm]]

    nDE <- sum(rna$adj.P.Val < 0.05)
    nDA <- sum(atac$adj.P.Val < 0.05)

    overlapSummary <- rbind(overlapSummary, data.frame(
      contrast = nm,
      DE_genes = nDE,
      DA_peaks = nDA
    ))
  }

  cat(sprintf("\n%s: DA-DGE Summary\n", comp))
  print(overlapSummary)

  if (nrow(overlapSummary) > 0) {
    overlapLong <- overlapSummary %>%
      pivot_longer(cols = c(DE_genes, DA_peaks),
        names_to = "modality", values_to = "count")

    pOverlap <- ggplot(overlapLong, aes(x = contrast, y = count, fill = modality)) +
      geom_col(position = "dodge") +
      scale_fill_manual(values = c("DE_genes" = "#377EB8", "DA_peaks" = "#FF7F00"),
        labels = c("DE Genes (RNA)", "DA Peaks (ATAC)")) +
      labs(x = "", y = "# Significant Features (FDR < 0.05)",
        fill = "", title = paste(comp, "— Significant Features per Contrast")) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    print(pOverlap)

    savePlot(plot = pOverlap, plotDir = plotDir,
      filename = paste0(comp, "_DA_DGE_overlap_barplot"),
      height = 6, width = 8)
  }
}
```

## Primary Multimodal Concordance

```{r multimodalConcordance, fig.height = 6, fig.width = 10}
multimodalSummary <- list()
multimodalCounts <- data.frame()

for (comp in compartments) {
  if (is.null(rnaResults[[comp]]) || is.null(atacResults[[comp]])) next
  multimodalSummary[[comp]] <- list()

  for (nm in intersect(primaryContrasts,
    intersect(names(rnaResults[[comp]]$results), names(atacResults[[comp]]$results)))) {
    joined <- buildMultimodalPeakGeneTable(
      rnaTable = rnaResults[[comp]]$results[[nm]],
      atacTable = atacResults[[comp]]$results[[nm]],
      peakAnnotation = peakAnnotations[[comp]]
    )

    if (nrow(joined) == 0) next

    multimodalSummary[[comp]][[nm]] <- joined
    write.csv(
      joined,
      file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_multimodal_peak_gene_", nm, ".csv")),
      row.names = FALSE
    )

    multimodalCounts <- bind_rows(
      multimodalCounts,
      joined %>%
        dplyr::count(multimodalClass, name = "n") %>%
        dplyr::mutate(compartment = comp, contrast = nm)
    )
  }
}

if (nrow(multimodalCounts) > 0) {
  pMulti <- ggplot(multimodalCounts,
    aes(x = contrast, y = n, fill = multimodalClass)) +
    geom_col(position = "stack") +
    facet_wrap(~compartment, scales = "free_y") +
    scale_fill_brewer(palette = "Set2", name = "Peak-gene class") +
    labs(
      x = "Primary Contrast",
      y = "# Significant DA peaks with nearby genes",
      title = "Primary Multimodal Concordance",
      subtitle = "Significant ATAC peaks linked to nearest genes within 100 kb"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pMulti)
  savePlot(
    plot = pMulti,
    plotDir = plotDir,
    filename = "primary_multimodal_concordance",
    height = 6,
    width = 10
  )
}
```

## LogFC Correlation (Top DE Peaks Nearest Genes)

```{r logFCCorrelation, fig.height = 6, fig.width = 8}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]]) || is.null(atacResults[[comp]])) next
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  for (nm in intersect(names(rnaResults[[comp]]$results),
    names(atacResults[[comp]]$results))) {
    atac <- atacResults[[comp]]$results[[nm]]
    rna <- rnaResults[[comp]]$results[[nm]]

    # Get significant DA peaks
    sigPeaks <- atac %>% dplyr::filter(adj.P.Val < 0.1)
    if (nrow(sigPeaks) < 5) next

    # Find nearest gene for each significant peak
    peakGR <- StringToGRanges(sigPeaks$feature)

    nearestGene <- tryCatch(
      {
        nearest <- ClosestFeature(seuratObj, regions = peakGR)
        nearest
      },
      error = function(e) NULL)

    if (is.null(nearestGene) || nrow(nearestGene) == 0) next

    # Merge ATAC logFC with RNA logFC for nearest gene
    sigPeaks$nearestGene <- nearestGene$gene_name[
      match(sigPeaks$feature, nearestGene$query_region)]
    sigPeaks$distance <- nearestGene$distance[
      match(sigPeaks$feature, nearestGene$query_region)]

    # Filter: only peaks within 100kb of a gene
    sigPeaks <- sigPeaks %>%
      dplyr::filter(!is.na(nearestGene) & distance <= 100000)

    if (nrow(sigPeaks) < 5) next

    # Join with RNA results
    merged <- sigPeaks %>%
      inner_join(rna %>% dplyr::select(feature, logFC, adj.P.Val) %>%
        dplyr::rename(RNA_logFC = logFC, RNA_FDR = adj.P.Val),
      by = c("nearestGene" = "feature"))

    if (nrow(merged) < 5) next

    corTest <- cor.test(merged$logFC, merged$RNA_logFC, method = "spearman")

    pCorr <- ggplot(merged, aes(x = logFC, y = RNA_logFC)) +
      geom_point(aes(color = -log10(adj.P.Val)), size = 2, alpha = 0.7) +
      geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.5) +
      scale_color_viridis_c(name = "-log10(ATAC FDR)") +
      geom_text_repel(
        data = merged %>% arrange(adj.P.Val) %>% head(10),
        aes(label = nearestGene), size = 3, max.overlaps = 15) +
      labs(x = "ATAC Peak logFC", y = "Nearest Gene RNA logFC",
        title = paste(comp, nm),
        subtitle = sprintf("Spearman rho = %.2f, p = %.2e (n = %d peaks)",
          corTest$estimate, corTest$p.value, nrow(merged)))

    print(pCorr)

    savePlot(plot = pCorr, plotDir = plotDir,
      filename = paste0(comp, "_logFC_correlation_", nm),
      height = 6, width = 8)
  }
}
```

# LogFC Scatterplots (Baseline vs Followup Reference) {.tabset}

## RNA LogFC Scatterplots

```{r rnaLogfcScatter, fig.height = 8, fig.width = 10}
scatterPairs <- list(
  list(xContrast = "PostIL2_vs_BL", yContrast = "PostIL2_vs_Followup",
    treatment = "Post-IL2", xRef = "Baseline", yRef = "Followup"),
  list(xContrast = "PostRAPA_vs_BL", yContrast = "PostRAPA_vs_Followup",
    treatment = "Post-RAPA", xRef = "Baseline", yRef = "Followup")
)

pCutoff <- 0.05

for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next

  for (sp in scatterPairs) {
    xTT <- rnaResults[[comp]]$results[[sp$xContrast]]
    yTT <- rnaResults[[comp]]$results[[sp$yContrast]]
    if (is.null(xTT) || is.null(yTT)) next

    merged_df <- merge(
      xTT[, c("feature", "logFC", "adj.P.Val")],
      yTT[, c("feature", "logFC", "adj.P.Val")],
      by = "feature", suffixes = c("_xRef", "_yRef"))

    merged_df <- merged_df %>%
      mutate(
        sig_x = adj.P.Val_xRef <= pCutoff,
        sig_y = adj.P.Val_yRef <= pCutoff,
        sig = sig_x | sig_y,
        distance_from_diagonal = abs(logFC_yRef - logFC_xRef)
      )

    top30 <- merged_df %>%
      filter(sig) %>%
      arrange(desc(distance_from_diagonal)) %>%
      slice(1:30)

    axMax <- max(abs(c(merged_df$logFC_xRef, merged_df$logFC_yRef)), na.rm = TRUE) * 1.05

    p <- ggplot(merged_df, aes(x = logFC_xRef, y = logFC_yRef)) +
      geom_point(data = filter(merged_df, !sig),
        aes(fill = adj.P.Val_yRef, color = adj.P.Val_xRef),
        shape = 21, size = 2, stroke = 0.8, alpha = 0.4) +
      geom_point(data = filter(merged_df, sig),
        aes(fill = adj.P.Val_yRef, color = adj.P.Val_xRef),
        shape = 21, size = 2.5, stroke = 0.8) +
      geom_text_repel(data = top30,
        aes(label = feature), size = 2.8, max.overlaps = 30,
        fontface = "italic", segment.color = "grey50", segment.size = 0.3) +
      scale_fill_gradient(
        low = "blue", high = "grey", limits = c(0, pCutoff),
        oob = scales::squish,
        name = paste0("FDR (vs ", sp$yRef, ")")) +
      scale_color_gradient(
        low = "red", high = "grey", limits = c(0, pCutoff),
        oob = scales::squish,
        name = paste0("FDR (vs ", sp$xRef, ")")) +
      geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "grey40") +
      geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
      geom_vline(xintercept = 0, color = "grey70", linewidth = 0.3) +
      annotate("text", x = axMax * 0.7, y = -axMax * 0.95,
        label = paste("Up in", sp$treatment), size = 3.2, color = "grey40") +
      annotate("text", x = -axMax * 0.7, y = -axMax * 0.95,
        label = paste("Up in", sp$xRef), size = 3.2, color = "grey40") +
      annotate("text", x = -axMax * 0.95, y = axMax * 0.7,
        label = paste("Up in", sp$treatment), size = 3.2, color = "grey40",
        angle = 90) +
      annotate("text", x = -axMax * 0.95, y = -axMax * 0.7,
        label = paste("Up in", sp$yRef), size = 3.2, color = "grey40",
        angle = 90) +
      coord_equal(xlim = c(-axMax, axMax), ylim = c(-axMax, axMax)) +
      labs(
        x = paste0("logFC (", sp$treatment, " vs ", sp$xRef, ")"),
        y = paste0("logFC (", sp$treatment, " vs ", sp$yRef, ")"),
        title = paste0(comp, " RNA — ", sp$treatment),
        subtitle = sprintf(
          "Fill = FDR vs %s, Outline = FDR vs %s\nSig: %d vs %s, %d vs %s, %d both",
          sp$yRef, sp$xRef,
          sum(merged_df$sig_x), sp$xRef,
          sum(merged_df$sig_y), sp$yRef,
          sum(merged_df$sig_x & merged_df$sig_y)))

    print(p)

    savePlot(plot = p, plotDir = plotDir,
      filename = paste0(comp, "_RNA_logFC_scatter_",
        gsub("-", "", sp$treatment)),
      height = 8, width = 12)
  }
}
```

## ATAC LogFC Scatterplots

```{r atacLogfcScatter, fig.height = 8, fig.width = 10}
for (comp in compartments) {
  if (is.null(atacResults[[comp]])) next
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  for (sp in scatterPairs) {
    xTT <- atacResults[[comp]]$results[[sp$xContrast]]
    yTT <- atacResults[[comp]]$results[[sp$yContrast]]
    if (is.null(xTT) || is.null(yTT)) next

    merged_df <- merge(
      xTT[, c("feature", "logFC", "adj.P.Val")],
      yTT[, c("feature", "logFC", "adj.P.Val")],
      by = "feature", suffixes = c("_xRef", "_yRef"))

    merged_df <- merged_df %>%
      mutate(
        sig_x = adj.P.Val_xRef <= pCutoff,
        sig_y = adj.P.Val_yRef <= pCutoff,
        sig = sig_x | sig_y,
        distance_from_diagonal = abs(logFC_yRef - logFC_xRef)
      )

    # Annotate significant peaks with nearest gene
    sigIdx <- which(merged_df$sig)
    if (length(sigIdx) == 0) next

    peakGR <- tryCatch(StringToGRanges(merged_df$feature[sigIdx]),
      error = function(e) NULL)
    if (is.null(peakGR)) next

    nearestInfo <- tryCatch(ClosestFeature(seuratObj, regions = peakGR),
      error = function(e) NULL)
    if (is.null(nearestInfo) || nrow(nearestInfo) == 0) next

    merged_df$nearestGene <- NA_character_
    merged_df$geneDistance <- NA_real_
    matched <- match(merged_df$feature[sigIdx], nearestInfo$query_region)
    merged_df$nearestGene[sigIdx] <- nearestInfo$gene_name[matched]
    merged_df$geneDistance[sigIdx] <- nearestInfo$distance[matched]

    # Top 15 off-diagonal sig peaks within 50kb of a gene
    labelDf <- merged_df %>%
      filter(sig & !is.na(nearestGene) & geneDistance <= 50000) %>%
      arrange(desc(distance_from_diagonal)) %>%
      slice(1:15)

    labelDf$label <- ifelse(
      labelDf$geneDistance == 0,
      labelDf$nearestGene,
      paste0(labelDf$nearestGene, " (", round(labelDf$geneDistance / 1000), "kb)"))

    axMax <- max(abs(c(merged_df$logFC_xRef, merged_df$logFC_yRef)),
      na.rm = TRUE) * 1.05

    p <- ggplot(merged_df, aes(x = logFC_xRef, y = logFC_yRef)) +
      geom_point(data = filter(merged_df, !sig),
        aes(fill = adj.P.Val_yRef, color = adj.P.Val_xRef),
        shape = 21, size = 1.2, stroke = 0.5, alpha = 0.3) +
      geom_point(data = filter(merged_df, sig),
        aes(fill = adj.P.Val_yRef, color = adj.P.Val_xRef),
        shape = 21, size = 1.8, stroke = 0.7) +
      geom_text_repel(data = labelDf,
        aes(x = logFC_xRef, y = logFC_yRef, label = label),
        size = 2.4, max.overlaps = 20,
        fontface = "italic", segment.color = "grey50", segment.size = 0.3) +
      scale_fill_gradient(
        low = "blue", high = "grey", limits = c(0, pCutoff),
        oob = scales::squish,
        name = paste0("FDR (vs ", sp$yRef, ")")) +
      scale_color_gradient(
        low = "red", high = "grey", limits = c(0, pCutoff),
        oob = scales::squish,
        name = paste0("FDR (vs ", sp$xRef, ")")) +
      geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "grey40") +
      geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
      geom_vline(xintercept = 0, color = "grey70", linewidth = 0.3) +
      annotate("text", x = axMax * 0.6, y = -axMax * 0.95,
        label = paste("More accessible in", sp$treatment),
        size = 2.8, color = "grey40") +
      annotate("text", x = -axMax * 0.6, y = -axMax * 0.95,
        label = paste("More accessible in", sp$xRef),
        size = 2.8, color = "grey40") +
      annotate("text", x = -axMax * 0.95, y = axMax * 0.6,
        label = paste("More accessible in", sp$treatment),
        size = 2.8, color = "grey40", angle = 90) +
      annotate("text", x = -axMax * 0.95, y = -axMax * 0.6,
        label = paste("More accessible in", sp$yRef),
        size = 2.8, color = "grey40", angle = 90) +
      coord_equal(xlim = c(-axMax, axMax), ylim = c(-axMax, axMax)) +
      labs(
        x = paste0("logFC (", sp$treatment, " vs ", sp$xRef, ")"),
        y = paste0("logFC (", sp$treatment, " vs ", sp$yRef, ")"),
        title = paste0(comp, " ATAC — ", sp$treatment),
        subtitle = sprintf(
          "Fill = FDR vs %s, Outline = FDR vs %s\nSig: %d vs %s, %d vs %s, %d both",
          sp$yRef, sp$xRef,
          sum(merged_df$sig_x), sp$xRef,
          sum(merged_df$sig_y), sp$yRef,
          sum(merged_df$sig_x & merged_df$sig_y)))

    print(p)

    savePlot(plot = p, plotDir = plotDir,
      filename = paste0(comp, "_ATAC_logFC_scatter_",
        gsub("-", "", sp$treatment)),
      height = 8, width = 12)
  }
}
```

# Export for SCENIC+

```{r scenicExport}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  exportDir <- file.path(dataOutputDir, paste0("scenic_export_", comp))
  if (!dir.exists(exportDir)) dir.create(exportDir, recursive = TRUE)

  # RNA counts (sparse matrix)
  DefaultAssay(seuratObj) <- "RNA"
  rnaCounts <- GetAssayData(seuratObj, layer = "counts")
  Matrix::writeMM(rnaCounts, file.path(exportDir, "RNA_counts.mtx"))
  writeLines(rownames(rnaCounts), file.path(exportDir, "RNA_genes.txt"))
  writeLines(colnames(rnaCounts), file.path(exportDir, "RNA_barcodes.txt"))

  # ATAC counts (sparse matrix)
  DefaultAssay(seuratObj) <- "ATAC"
  atacCounts <- GetAssayData(seuratObj, layer = "counts")
  Matrix::writeMM(atacCounts, file.path(exportDir, "ATAC_counts.mtx"))
  writeLines(rownames(atacCounts), file.path(exportDir, "ATAC_peaks.txt"))
  writeLines(colnames(atacCounts), file.path(exportDir, "ATAC_barcodes.txt"))

  # Metadata
  metaCols <- c("donorID", "visitNum", "treatmentPhase", "wnn_clusters",
    "RNA.weight", "ATAC.weight")
  metaCols <- intersect(metaCols, colnames(seuratObj@meta.data))
  meta <- seuratObj@meta.data[, metaCols]
  write.csv(meta, file.path(exportDir, "metadata.csv"))

  # Embeddings
  for (redName in c("umap.wnn", "harmony", "harmony_lsi")) {
    if (redName %in% names(seuratObj@reductions)) {
      emb <- Embeddings(seuratObj, reduction = redName)
      write.csv(emb, file.path(exportDir, paste0("embedding_", redName, ".csv")))
    }
  }

  cat(sprintf("%s: SCENIC+ export saved to %s\n", comp, exportDir))
  cat(sprintf("  RNA: %d genes × %d cells\n", nrow(rnaCounts), ncol(rnaCounts)))
  cat(sprintf("  ATAC: %d peaks × %d cells\n", nrow(atacCounts), ncol(atacCounts)))
}
```

# Save DA Bundle

```{r saveDABundle}
daBundle_path <- file.path(dataOutputDir,
  paste0(filenameSuffix, "_DA_bundle.rds"))

daBundle <- list(
  # Global results
  rnaResults = rnaResults,
  atacResults = atacResults,

  # Coverage and interpretation helpers
  primaryContrasts = primaryContrasts,
  globalCoverage = globalCoverage,
  coverageSummary = coverageSummaryDf,
  peakAnnotations = peakAnnotations,
  motifResults = motifResults,
  multimodalSummary = multimodalSummary,

  # Per-cluster results
  perClusterRNA = perClusterRNA,
  perClusterATAC = perClusterATAC,
  perClusterRNAStatus = perClusterRNAStatus,
  perClusterATACStatus = perClusterATACStatus,

  # GSEA
  gseaResults = gseaResults,

  # Roast IL2-STAT5
  roastResults = roastResults,
  roastPerCluster = roastPerCluster,

  # C7 Immunologic GSEA
  c7GseaResults = c7GseaResults,
  c7GseaPerCluster = c7GseaPerCluster,

  # Contrasts used
  contrastDefs = contrastDefs,

  # Carry forward palettes
  palDonorId = palDonorId,
  palVisit = palVisit,
  palTreatmentPhase = palTreatmentPhase,

  # Metadata
  dataDate = dataDate,
  bundle_created = Sys.time(),
  source_script = "P697_DA.rmd"
)

saveRDS(daBundle, file = daBundle_path)
cat("DA bundle saved to:", daBundle_path, "\n")
cat("Bundle contains:", paste(names(daBundle), collapse = ", "), "\n")
```

# Session Info

```{r sessionInfo, results = "markup"}
sessionInfo()
```
