---
title: "P697: NK ATAC Peak Calling (MACS2 via Signac)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
setwd("/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

library(knitr)
library(dplyr)
library(tibble)
library(ggplot2)
library(viridis)
library(stringr)
library(patchwork)
library(Seurat)
library(Signac)
library(GenomicRanges)
library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggpointdensity)
library(RColorBrewer)
library(cowplot)
library(harmony)

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      axis.text = element_text(colour = "black"),
      axis.ticks = element_line(colour = "black"),
      legend.key = element_blank(),
      strip.text.x = element_text(size = 14, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(fill = "white", colour = "black")))

opts_chunk$set(fig.width = 6, fig.height = 4.0, cache = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE, cache.lazy = FALSE, results = "hide")
opts_knit$set(root.dir = "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

options(stringsAsFactors = FALSE)
options(future.globals.maxSize = 1591289600)
```

# Setup

```{r setUpDirectories}
projNumber <- "P697"
baseDir <- "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2"
dataInputDir <- file.path(baseDir, "data/inputData")
plotDir <- file.path(baseDir, "figures")
dataOutputDir <- file.path(baseDir, "data/outputData")
dataDate <- Sys.Date()
filenameSuffix <- paste0(projNumber, ".", dataDate)
```

```{r setupFunctions}
source(file.path(getwd(), "code/P697_functions.R"))
```

# Load Preprocessing Bundle

```{r loadBundle}
# Load most recent preprocessing bundle
# NOTE: ATAC assay was stripped before merge in preprocessing (to avoid slow
# ChromatinAssay merge). It will be reconstructed from fragment files below.
bundleFiles <- list.files(dataOutputDir,
  pattern = "P697\\..*_preprocessingBundle\\.rds$",
  full.names = TRUE)
bundleFile <- bundleFiles[which.max(file.mtime(bundleFiles))]
cat("Loading bundle:", bundleFile, "\n")

bundle <- readRDS(bundleFile)

seuratNK <- bundle$seuratNK
tcrGeneList <- bundle$tcrGeneList
palDonorId <- bundle$palDonorId
palVisit <- bundle$palVisit

cat("NK object:", ncol(seuratNK), "cells\n")
cat("Assays:", paste(names(seuratNK@assays), collapse = ", "), "\n")

# Verify ATAC metadata survived the merge (needed for QC plots)
stopifnot("nCount_ATAC" %in% colnames(seuratNK@meta.data))
stopifnot("hasATAC" %in% colnames(seuratNK@meta.data))
nATACCells <- sum(seuratNK$hasATAC, na.rm = TRUE)
cat(sprintf("Cells with ATAC data (hasATAC): %d of %d (%.1f%%)\n",
  nATACCells, ncol(seuratNK), 100 * nATACCells / ncol(seuratNK)))
cat(sprintf("nCount_ATAC — median: %s, mean: %s\n",
  format(median(seuratNK$nCount_ATAC[seuratNK$hasATAC], na.rm = TRUE), big.mark = ","),
  format(round(mean(seuratNK$nCount_ATAC[seuratNK$hasATAC], na.rm = TRUE)), big.mark = ",")))

# Free bundle — seuratNK was extracted above, keeping bundle wastes ~5+ GB
rm(bundle)
gc(verbose = FALSE)
```

# Reconstruct ATAC Assay from Fragment Files

```{r reconstructATAC}
# The ATAC assay was stripped before merge in preprocessing to avoid extremely
# slow ChromatinAssay merging. Here we reconstruct it from the original fragment
# files using pool metadata and cell-name mapping.

# ---- Genome annotations (reused in replaceATACAssay later) ----
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86)
seqlevelsStyle(annotations) <- "UCSC"
genome(annotations) <- "hg38"

# ---- Find fragment files ----
fragDir <- file.path(baseDir, "data/outputData")
fragFiles <- list.files(fragDir,
  pattern = "pool697-.*_atac_fragments\\.5col\\.tsv\\.gz$",
  full.names = TRUE)
fragFiles <- fragFiles[!grepl("\\.tbi$", fragFiles)]
cat(sprintf("Found %d fragment files\n", length(fragFiles)))

# ---- Build per-pool Fragment objects with cell-name mapping ----
poolsInCompartment <- unique(seuratNK$pool[seuratNK$hasATAC])
cat(sprintf("Pools with ATAC cells in NK compartment: %s\n",
  paste(poolsInCompartment, collapse = ", ")))

fragList <- list()
for (poolName in poolsInCompartment) {
  # Find the fragment file for this pool
  fragFile <- fragFiles[grepl(paste0(poolName, "_atac_fragments"), fragFiles)]
  if (length(fragFile) != 1) {
    warning(sprintf("Expected 1 fragment file for %s, found %d — skipping",
      poolName, length(fragFile)))
    next
  }

  # Get merged cell names for this pool
  poolCells <- colnames(seuratNK)[seuratNK$pool == poolName & seuratNK$hasATAC]

  # Use bare barcodes (without pool prefix) to match fragment file content.
  # Pool-prefix renaming is handled later in buildPeakMatrix.
  bareBarcodes <- sub(paste0("^", poolName, "_"), "", poolCells)

  fragObj <- CreateFragmentObject(
    path = fragFile,
    cells = bareBarcodes,
    validate.fragments = FALSE
  )
  fragList <- c(fragList, fragObj)

  cat(sprintf("  %s: %d cells, fragment file: %s\n",
    poolName, length(bareBarcodes), basename(fragFile)))
}

cat(sprintf("Total Fragment objects created: %d\n", length(fragList)))

# Fragment objects and annotations are now ready as standalone variables.
# No dummy ATAC assay is created — the real ChromatinAssay will be built
# after MACS2 peak calling in the replaceATACAssay chunk.
# QC plots below use nCount_ATAC metadata from preprocessing.
atacCellNames <- colnames(seuratNK)[seuratNK$hasATAC]
cat(sprintf("ATAC cells in NK: %d\n", length(atacCellNames)))
```

# ATAC Fragment Depth QC {.tabset}

## Fragments per cell distribution

```{r atacFragmentDistribution, fig.height = 6, fig.width = 10}
# Get ATAC fragment counts per cell
# nCount_ATAC carried forward from preprocessing metadata (computed from Cell
# Ranger Arc peaks). No ATAC assay exists yet — it will be created after MACS2
# peak calling. QC plots use the preserved metadata columns.

atacMeta <- seuratNK@meta.data
if (!"barcode" %in% colnames(atacMeta)) {
  atacMeta$barcode <- rownames(atacMeta)
}

# nCount_ATAC must be present from preprocessing
stopifnot("nCount_ATAC" %in% colnames(atacMeta))
cat("Using nCount_ATAC from preprocessing metadata\n")

# Also check for CR ATAC metrics (total fragments from Cell Ranger Arc)
hasCRFragments <- "cr_arc_atac_fragments" %in% colnames(atacMeta)

# Summary statistics
cat("\n=== ATAC Fragment Depth Summary (NK) ===\n")
cat(sprintf("Total cells with ATAC: %d\n", sum(!is.na(atacMeta$nCount_ATAC))))
cat(sprintf("nCount_ATAC (peaks):\n"))
cat(sprintf("  Min:    %s\n", format(min(atacMeta$nCount_ATAC, na.rm = TRUE), big.mark = ",")))
cat(sprintf("  Q1:     %s\n", format(quantile(atacMeta$nCount_ATAC, 0.25, na.rm = TRUE), big.mark = ",")))
cat(sprintf("  Median: %s\n", format(median(atacMeta$nCount_ATAC, na.rm = TRUE), big.mark = ",")))
cat(sprintf("  Mean:   %s\n", format(round(mean(atacMeta$nCount_ATAC, na.rm = TRUE)), big.mark = ",")))
cat(sprintf("  Q3:     %s\n", format(quantile(atacMeta$nCount_ATAC, 0.75, na.rm = TRUE), big.mark = ",")))
cat(sprintf("  Max:    %s\n", format(max(atacMeta$nCount_ATAC, na.rm = TRUE), big.mark = ",")))

if (hasCRFragments) {
  cat(sprintf("\ncr_arc_atac_fragments (total ATAC fragments from Cell Ranger Arc):\n"))
  cat(sprintf("  Min:    %s\n", format(min(atacMeta$cr_arc_atac_fragments, na.rm = TRUE), big.mark = ",")))
  cat(sprintf("  Median: %s\n", format(median(atacMeta$cr_arc_atac_fragments, na.rm = TRUE), big.mark = ",")))
  cat(sprintf("  Mean:   %s\n", format(round(mean(atacMeta$cr_arc_atac_fragments, na.rm = TRUE)), big.mark = ",")))
  cat(sprintf("  Max:    %s\n", format(max(atacMeta$cr_arc_atac_fragments, na.rm = TRUE), big.mark = ",")))
}

# Cells below candidate thresholds
for (thresh in c(500, 1000, 2000, 5000)) {
  nBelow <- sum(atacMeta$nCount_ATAC < thresh, na.rm = TRUE)
  cat(sprintf("  Cells with nCount_ATAC < %s: %d (%.1f%%)\n",
    format(thresh, big.mark = ","), nBelow,
    100 * nBelow / sum(!is.na(atacMeta$nCount_ATAC))))
}

# ---- Histogram of nCount_ATAC ----
pHist <- ggplot(atacMeta %>% dplyr::filter(!is.na(nCount_ATAC)),
  aes(x = nCount_ATAC)) +
  geom_histogram(bins = 100, fill = "darkorange3", color = "black", linewidth = 0.2) +
  scale_x_log10(labels = scales::comma) +
  geom_vline(xintercept = 1000, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_vline(xintercept = 500, linetype = "dotted", color = "orange", linewidth = 0.8) +
  annotate("text", x = 1000, y = Inf, label = "1000", vjust = 2, hjust = -0.1,
    color = "red", size = 3.5) +
  annotate("text", x = 500, y = Inf, label = "500", vjust = 3.5, hjust = -0.1,
    color = "orange", size = 3.5) +
  labs(x = "ATAC fragment count per cell (log10)",
    y = "Number of cells",
    title = "NK: ATAC fragments per cell") +
  theme(plot.title = element_text(size = 14))

# ---- Density plot ----
pDens <- ggplot(atacMeta %>% dplyr::filter(!is.na(nCount_ATAC)),
  aes(x = nCount_ATAC)) +
  geom_density(fill = "darkorange3", alpha = 0.4) +
  scale_x_log10(labels = scales::comma) +
  geom_vline(xintercept = 1000, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_vline(xintercept = 500, linetype = "dotted", color = "orange", linewidth = 0.8) +
  labs(x = "ATAC fragment count per cell (log10)",
    y = "Density",
    title = "NK: Fragment depth density") +
  theme(plot.title = element_text(size = 14))

pHist / pDens

savePlot(
  plot = pHist / pDens,
  plotDir = plotDir,
  filename = "NK_ATAC_fragmentDepth_histogram",
  height = 8, width = 10
)
```

## Fragments by donor and visit

```{r atacFragmentsByDonorVisit, fig.height = 8, fig.width = 12}
# Faceted violin plots of fragment depth by donor and visit

if (all(c("donorID", "visitNum") %in% colnames(atacMeta))) {
  atacMeta$donorID <- factor(atacMeta$donorID)
  atacMeta$visitNum <- factor(atacMeta$visitNum)

  # ---- By donor ----
  pByDonor <- ggplot(atacMeta %>% dplyr::filter(!is.na(nCount_ATAC)),
    aes(x = donorID, y = nCount_ATAC, fill = donorID)) +
    geom_violin(scale = "width") +
    geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.3) +
    scale_y_log10(labels = scales::comma) +
    geom_hline(yintercept = 1000, linetype = "dashed", color = "red") +
    scale_fill_manual(values = palDonorId) +
    labs(x = "Donor", y = "ATAC fragments per cell (log10)",
      title = "NK: Fragment depth by donor") +
    theme(legend.position = "none")

  # ---- By visit ----
  pByVisit <- ggplot(atacMeta %>% dplyr::filter(!is.na(nCount_ATAC)),
    aes(x = visitNum, y = nCount_ATAC, fill = visitNum)) +
    geom_violin(scale = "width") +
    geom_boxplot(width = 0.15, fill = "white", outlier.size = 0.3) +
    scale_y_log10(labels = scales::comma) +
    geom_hline(yintercept = 1000, linetype = "dashed", color = "red") +
    scale_fill_manual(values = palVisit) +
    labs(x = "Visit", y = "ATAC fragments per cell (log10)",
      title = "NK: Fragment depth by visit") +
    theme(legend.position = "none")

  # ---- Faceted by donor x visit ----
  pFaceted <- ggplot(atacMeta %>% dplyr::filter(!is.na(nCount_ATAC)),
    aes(x = visitNum, y = nCount_ATAC, fill = visitNum)) +
    geom_violin(scale = "width") +
    geom_boxplot(width = 0.2, fill = "white", outlier.size = 0.3) +
    scale_y_log10(labels = scales::comma) +
    geom_hline(yintercept = 1000, linetype = "dashed", color = "red") +
    scale_fill_manual(values = palVisit) +
    facet_wrap(~donorID, nrow = 2) +
    labs(x = "Visit", y = "ATAC fragments (log10)",
      title = "NK: Fragment depth by donor x visit") +
    theme(legend.position = "none",
      axis.text.x = element_text(size = 9))

  (pByDonor | pByVisit) / pFaceted +
    plot_layout(heights = c(1, 1.5))

  savePlot(
    plot = (pByDonor | pByVisit) / pFaceted + plot_layout(heights = c(1, 1.5)),
    plotDir = plotDir,
    filename = "NK_ATAC_fragmentDepth_byDonorVisit",
    height = 10, width = 14
  )
}
```

## nFeature_ATAC vs nCount_ATAC

```{r atacFeatureVsCount, fig.height = 6, fig.width = 7}
if (all(c("nCount_ATAC", "nFeature_ATAC") %in% colnames(seuratNK@meta.data))) {
  pScatter <- ggplot(seuratNK@meta.data %>% dplyr::filter(!is.na(nCount_ATAC)),
    aes(x = nCount_ATAC, y = nFeature_ATAC)) +
    geom_pointdensity(size = 0.3) +
    scale_color_viridis() +
    scale_x_log10(labels = scales::comma) +
    scale_y_log10(labels = scales::comma) +
    geom_vline(xintercept = 1000, linetype = "dashed", color = "red") +
    labs(x = "nCount_ATAC (total fragments in peaks)",
      y = "nFeature_ATAC (peaks detected)",
      title = "NK: ATAC complexity") +
    theme(legend.position = "right")

  print(pScatter)

  savePlot(
    plot = pScatter,
    plotDir = plotDir,
    filename = "NK_ATAC_nFeature_vs_nCount",
    height = 6, width = 8
  )
}
```

# ATAC Depth Filtering

```{r atacDepthFilter}
# ---- Set ATAC depth threshold ----
# Review the plots above before choosing. 1000 is a common starting point.
# Adjust this value based on the fragment distribution observed above.
atacMinFragments <- 1000

nBefore <- ncol(seuratNK)
nATACBefore <- sum(!is.na(seuratNK$nCount_ATAC))

# Identify cells passing filter
cellsPassATAC <- colnames(seuratNK)[
  !is.na(seuratNK$nCount_ATAC) & seuratNK$nCount_ATAC >= atacMinFragments]
cellsFailATAC <- colnames(seuratNK)[
  !is.na(seuratNK$nCount_ATAC) & seuratNK$nCount_ATAC < atacMinFragments]

seuratNK$atacPassDepthFilter <- colnames(seuratNK) %in% cellsPassATAC

cat(sprintf("\n=== ATAC Depth Filter (threshold: %s fragments) ===\n",
  format(atacMinFragments, big.mark = ",")))
cat(sprintf("Cells with ATAC data: %d\n", nATACBefore))
cat(sprintf("Pass filter: %d (%.1f%%)\n",
  length(cellsPassATAC), 100 * length(cellsPassATAC) / nATACBefore))
cat(sprintf("Fail filter: %d (%.1f%%)\n",
  length(cellsFailATAC), 100 * length(cellsFailATAC) / nATACBefore))
cat(sprintf("Total cells retained (all assays): %d\n", ncol(seuratNK)))
```

# MACS2 Peak Calling

```{r callPeaksMACS2}
# Call peaks using MACS2 directly on filtered fragment files.
# Signac::CallPeaks doesn't support a list of Fragment objects, so we merge
# per-pool fragment BED files (filtered to depth-passing NK cells) and
# invoke MACS2 via system2.

cat("Calling peaks with MACS2 on NK compartment...\n")
cat(sprintf("Using %d cells passing ATAC depth filter\n", length(cellsPassATAC)))

macs2Path <- "/Users/tedwards/miniconda3/envs/macs2_env/bin/macs2"
stopifnot("MACS2 not found" = file.exists(macs2Path))

peakOutDir <- file.path(dataOutputDir, "macs2_peaks")
tmpDir <- file.path(dataOutputDir, "macs2_tmp")
dir.create(peakOutDir, showWarnings = FALSE, recursive = TRUE)
dir.create(tmpDir, showWarnings = FALSE, recursive = TRUE)

# ---- Merge per-pool fragment files, filtered to depth-passing cells ----
tmpBed <- file.path(tmpDir, "NK_filtered_fragments.bed")
if (file.exists(tmpBed)) unlink(tmpBed)
file.create(tmpBed)  # pre-create so >> always has a target

# Verify prerequisites
stopifnot("poolsInCompartment not found" = exists("poolsInCompartment"))
stopifnot("fragFiles not found" = exists("fragFiles"))
stopifnot("tmpDir does not exist" = dir.exists(tmpDir))
cat(sprintf("Pools to process: %s\n", paste(poolsInCompartment, collapse = ", ")))
cat(sprintf("Fragment files available: %d\n", length(fragFiles)))

# Diagnostic: verify cellsPassATAC content
cat(sprintf("cellsPassATAC: %d cells, e.g. %s\n",
  length(cellsPassATAC), paste(head(cellsPassATAC, 3), collapse = ", ")))

cat("Filtering and merging fragments per pool...\n")
for (poolName in poolsInCompartment) {
  # Match cells by name prefix (poolName_) — more robust than metadata query
  poolPrefix <- paste0(poolName, "_")
  poolCellsPass <- cellsPassATAC[startsWith(cellsPassATAC, poolPrefix)]
  if (length(poolCellsPass) == 0) {
    cat(sprintf("  %s: 0 cells passing filter — skipping\n", poolName))
    next
  }

  # Original barcodes (strip pool prefix added by merge)
  originalBarcodes <- sub(paste0("^", poolName, "_"), "", poolCellsPass)

  # Write barcodes for awk filtering
  bcFile <- tempfile(fileext = ".txt")
  writeLines(originalBarcodes, bcFile)

  # Find fragment file for this pool
  poolFragFile <- fragFiles[grepl(paste0(poolName, "_atac_fragments"), fragFiles)]
  if (length(poolFragFile) != 1) {
    warning(sprintf("Expected 1 fragment file for %s, found %d — skipping",
      poolName, length(poolFragFile)))
    unlink(bcFile)
    next
  }

  # Filter fragments by barcode (col 4) and append to merged BED
  # Use gunzip -c (not zcat) for macOS compatibility
  cmd <- sprintf("gunzip -c %s | awk 'NR==FNR{a[$1];next} $4 in a' %s - >> %s",
    shQuote(poolFragFile), shQuote(bcFile), shQuote(tmpBed))
  exitStatus <- system(cmd)

  unlink(bcFile)
  cat(sprintf("  %s: %d cells (exit %d)\n", poolName, length(poolCellsPass), exitStatus))
}

nFragLines <- as.integer(trimws(system(sprintf("wc -l < %s", shQuote(tmpBed)), intern = TRUE)))
cat(sprintf("Total fragments in merged BED: %s\n", format(nFragLines, big.mark = ",")))
stopifnot("No fragments written — check fragment files and cell barcodes" = nFragLines > 0)

# ---- Run MACS2 ----
macs2Prefix <- "NK_MACS2"
cat("Running MACS2 callpeak...\n")
macs2_result <- system2(macs2Path, args = c(
  "callpeak",
  "-t", tmpBed,
  "-f", "BED",
  "-g", "2.7e9",
  "--nomodel",
  "--extsize", "200",
  "--shift", "-100",
  "--outdir", peakOutDir,
  "-n", macs2Prefix,
  "--keep-dup", "all"
), stdout = TRUE, stderr = TRUE)
cat(paste(tail(macs2_result, 15), collapse = "\n"), "\n")

# ---- Read MACS2 peaks as GRanges ----
peakFile <- file.path(peakOutDir, paste0(macs2Prefix, "_peaks.narrowPeak"))
stopifnot("MACS2 narrowPeak output not found" = file.exists(peakFile))

macs2Peaks <- makeGRangesFromDataFrame(
  read.table(peakFile, col.names = c("chr", "start", "end", "name", "score",
    "strand", "signalValue", "pValue", "qValue", "peak")),
  keep.extra.columns = TRUE,
  starts.in.df.are.0based = TRUE
)

# Clean up temp BED
unlink(tmpBed)

cat(sprintf("\nMACS2 called %d peaks\n", length(macs2Peaks)))
cat(sprintf("Peak width: min %d, median %d, max %d bp\n",
  min(width(macs2Peaks)), median(width(macs2Peaks)), max(width(macs2Peaks))))

gc(verbose = FALSE)
```

```{r removeBlacklistRegions}
# Remove peaks overlapping ENCODE blacklist regions
nPeaksBefore <- length(macs2Peaks)

macs2Peaks <- keepStandardChromosomes(macs2Peaks, pruning.mode = "coarse")
macs2Peaks <- subsetByOverlaps(macs2Peaks, blacklist_hg38_unified, invert = TRUE)

nPeaksAfter <- length(macs2Peaks)
cat(sprintf("Peaks before blacklist filter: %d\n", nPeaksBefore))
cat(sprintf("Peaks after blacklist filter: %d (%d removed)\n",
  nPeaksAfter, nPeaksBefore - nPeaksAfter))
```

# Build New Peak-Count Matrix

```{r subsetToATACCells}
# Subset seuratNK to ATAC depth-passing cells BEFORE building the peak matrix.
# This frees ~2+ GB of RNA/HT data for non-ATAC cells, which is critical on
# a 24 GB machine where CreateChromatinAssay would otherwise OOM.

# Clear nearest-neighbor graphs from preprocessing — they cause
# "Please provide rownames to the matrix" error during Seurat v5 subsetting.
for (g in names(seuratNK@graphs)) seuratNK@graphs[[g]] <- NULL

nCellsBefore <- ncol(seuratNK)
seuratNK <- seuratNK[, cellsPassATAC]
cat(sprintf("Subset seuratNK to depth-passing ATAC cells: %d -> %d cells\n",
  nCellsBefore, ncol(seuratNK)))

# Free objects no longer needed
rm(atacMeta, cellsFailATAC)
gc(verbose = FALSE)
```

```{r buildPeakMatrix}
cat("Building peak-count matrix from MACS2 peaks...\n")
cat(sprintf("Total cells in seuratNK: %d\n", ncol(seuratNK)))

# Process one pool at a time to avoid barcode collisions across pools
# (same 10x barcodes reused in each pool) and reduce peak memory.
peakCountsList <- list()
for (i in seq_along(fragList)) {
  poolName <- poolsInCompartment[i]
  poolCells <- Cells(fragList[[i]])
  cat(sprintf("  %s: %d cells ... ", poolName, length(poolCells)))

  pc <- Signac::FeatureMatrix(
    fragments = fragList[i],
    features = macs2Peaks,
    cells = poolCells,
    verbose = FALSE
  )

  # Fix peak name format (chr1:100-200 -> chr1-100-200)
  rownames(pc) <- gsub(":", "-", rownames(pc))

  # Rename columns from bare barcodes to pool-prefixed merged names
  mergedNames <- paste0(poolName, "_", colnames(pc))
  # Only keep cells that exist in seuratNK
  keep <- mergedNames %in% colnames(seuratNK)
  pc <- pc[, keep, drop = FALSE]
  colnames(pc) <- mergedNames[keep]

  cat(sprintf("%d cells matched\n", ncol(pc)))
  peakCountsList[[i]] <- pc
  rm(pc); gc(verbose = FALSE)
}

peakCounts <- do.call(cbind, peakCountsList)
rm(peakCountsList); gc(verbose = FALSE)

# FeatureMatrix can return ngCMatrix (binary pattern matrix) — convert to
# numeric dgCMatrix so CreateChromatinAssay stores actual count values.
if (is(peakCounts, "ngCMatrix") || is(peakCounts, "lgCMatrix")) {
  peakCounts <- as(peakCounts, "dgCMatrix")
}

stopifnot("No cells in peak matrix" = ncol(peakCounts) > 0)
cat(sprintf("Peak matrix: %d peaks x %d cells\n",
  nrow(peakCounts), ncol(peakCounts)))
```

```{r replaceATACAssay}
# Create new ChromatinAssay with MACS2 peaks.
# Pass pre-built GRanges via ranges= to bypass tidyr::separate() which hits
# R 4.3 long vector limits on large matrices.
peakNames <- rownames(peakCounts)
parts <- do.call(rbind, strsplit(peakNames, "-"))
peakRanges <- GRanges(
  seqnames = parts[, 1],
  ranges = IRanges(start = as.integer(parts[, 2]), end = as.integer(parts[, 3]))
)

newATAC <- CreateChromatinAssay(
  counts = peakCounts,
  ranges = peakRanges,
  min.cells = 0,
  min.features = 0
)

# Add annotation (fragments not attached here — Signac's Cells() returns bare
# barcodes from Fragment objects regardless of cellMap, causing AddFragments to
# fail. Fragments will be re-attached when needed for coverage/footprinting.)
Annotation(newATAC) <- annotations

# Replace the ATAC assay
seuratNK[["ATAC"]] <- newATAC
DefaultAssay(seuratNK) <- "ATAC"

# Free intermediates no longer needed
rm(newATAC, peakCounts)
gc(verbose = FALSE)

cat("Replaced ATAC assay with MACS2-called peaks\n")
cat(sprintf("New ATAC assay: %d peaks x %d cells\n",
  nrow(seuratNK[["ATAC"]]), ncol(seuratNK[["ATAC"]])))
```

# ATAC Normalization (LSI Pipeline)

```{r lsiNormalization}
DefaultAssay(seuratNK) <- "ATAC"

seuratNK <- FindTopFeatures(seuratNK, min.cutoff = "q5")
cat(sprintf("Variable features after q5 cutoff: %d of %d peaks\n",
  length(VariableFeatures(seuratNK)), nrow(seuratNK[["ATAC"]])))

seuratNK <- RunTFIDF(seuratNK)

seuratNK <- RunSVD(seuratNK, n = 50)

cat("LSI normalization complete (50 components)\n")
```

```{r harmonyIntegration}
# Integrate across donors to remove donor-level technical variation
# (library complexity, fragment depth) that can dominate the LSI embedding.
seuratNK <- RunHarmony(
  seuratNK,
  group.by.vars = "donorID",
  reduction = "lsi",
  reduction.save = "harmony_lsi",
  dims.use = 2:50,
  project.dim = FALSE
)
cat("Harmony integration complete on LSI embedding\n")
```

```{r depthCorrelation, fig.height = 5, fig.width = 8}
# Check correlation of LSI components with sequencing depth
depthCorPlot <- DepthCor(seuratNK, n = 50)
print(depthCorPlot)

savePlot(
  plot = depthCorPlot,
  plotDir = plotDir,
  filename = "NK_ATAC_LSI_DepthCor",
  height = 5, width = 8
)

# Report correlation values for first few components
lsiEmbeddings <- Embeddings(seuratNK, reduction = "lsi")
atacCounts <- seuratNK$nCount_ATAC
corValues <- sapply(1:min(10, ncol(lsiEmbeddings)), function(i) {
  cor(lsiEmbeddings[, i], atacCounts[rownames(lsiEmbeddings)], use = "complete.obs")
})
cat("\nLSI component correlations with nCount_ATAC:\n")
for (i in seq_along(corValues)) {
  flag <- ifelse(abs(corValues[i]) > 0.7, " ** EXCLUDE **", "")
  cat(sprintf("  Component %d: r = %.3f%s\n", i, corValues[i], flag))
}
```

# ATAC UMAP

```{r atacUMAP, fig.height = 6, fig.width = 8}
lsiDimsToUse <- 2:30
cat(sprintf("Running ATAC UMAP with LSI components %d-%d\n",
  min(lsiDimsToUse), max(lsiDimsToUse)))

set.seed(6022)
seuratNK <- RunUMAP(
  seuratNK,
  reduction = "harmony_lsi",
  dims = 1:29,
  reduction.name = "umap.atac",
  reduction.key = "atacUMAP_"
)

# Plot ATAC UMAP colored by various metadata
pCluster <- DimPlot(seuratNK, reduction = "umap.atac",
  group.by = "seurat_clusters", label = TRUE) +
  ggtitle("NK ATAC UMAP: RNA clusters")

pDonor <- DimPlot(seuratNK, reduction = "umap.atac",
  group.by = "donorID", cols = palDonorId) +
  ggtitle("NK ATAC UMAP: Donor")

pVisit <- DimPlot(seuratNK, reduction = "umap.atac",
  group.by = "visitNum", cols = palVisit) +
  ggtitle("NK ATAC UMAP: Visit")

(pCluster | pDonor) / (pVisit | plot_spacer())

savePlot(
  plot = (pCluster | pDonor) / (pVisit | plot_spacer()),
  plotDir = plotDir,
  filename = "NK_ATAC_UMAP_overview",
  height = 10, width = 14
)
```

# QC: New Peaks vs Cell Ranger Peaks

```{r peakComparisonQC}
cat("\n=== Peak Calling Comparison ===\n")
cat(sprintf("MACS2 peaks (NK): %d\n", length(macs2Peaks)))
cat(sprintf("  Total width: %s bp\n",
  format(sum(width(macs2Peaks)), big.mark = ",")))
cat(sprintf("  Mean width: %.0f bp\n", mean(width(macs2Peaks))))
cat(sprintf("  Median width: %.0f bp\n", median(width(macs2Peaks))))

# FRiP with new peaks
newPeakCounts <- colSums(GetAssayData(seuratNK, assay = "ATAC", layer = "counts"))
if ("cr_arc_atac_fragments" %in% colnames(seuratNK@meta.data)) {
  totalFrags <- seuratNK$cr_arc_atac_fragments
  cellsWithBoth <- !is.na(totalFrags) & !is.na(newPeakCounts) & totalFrags > 0
  newFRiP <- newPeakCounts[cellsWithBoth] / totalFrags[cellsWithBoth]
  cat(sprintf("\nFRiP with MACS2 peaks:\n"))
  cat(sprintf("  Median: %.1f%%\n", 100 * median(newFRiP)))
  cat(sprintf("  Mean: %.1f%%\n", 100 * mean(newFRiP)))
}
```

# Save Peak Calling Bundle

```{r saveBundle}
peakCallingBundle_path <- file.path(dataOutputDir,
  paste0(filenameSuffix, "_NK_peakCallingBundle.rds"))

peakCallingBundle <- list(
  seuratNK = seuratNK,
  macs2Peaks = macs2Peaks,
  atacMinFragments = atacMinFragments,
  lsiDimsToUse = lsiDimsToUse,

  # Carry forward from preprocessing
  tcrGeneList = tcrGeneList,
  palDonorId = palDonorId,
  palVisit = palVisit,

  # Metadata
  dataDate = dataDate,
  bundle_created = Sys.time(),
  source_script = "P697_NKPeakCalling.rmd"
)

saveRDS(peakCallingBundle, file = peakCallingBundle_path)
cat("NK peak calling bundle saved to:", peakCallingBundle_path, "\n")
cat("Bundle contains:", paste(names(peakCallingBundle), collapse = ", "), "\n")
```

# Session Info

```{r sessionInfo, results = "markup"}
sessionInfo()
```
