---
title: "P697: WNN Multimodal Analysis (RNA + ATAC)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
setwd("/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

library(knitr)
library(dplyr)
library(tibble)
library(ggplot2)
library(viridis)
library(stringr)
library(patchwork)
library(Seurat)
library(Signac)
library(ggpointdensity)
library(RColorBrewer)
library(cowplot)
library(cluster)
library(ComplexHeatmap)
library(circlize)

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      axis.text = element_text(colour = "black"),
      axis.ticks = element_line(colour = "black"),
      legend.key = element_blank(),
      strip.text.x = element_text(size = 14, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(fill = "white", colour = "black")))

opts_chunk$set(fig.width = 6, fig.height = 4.0, cache = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE, cache.lazy = FALSE, results = "hide")
opts_knit$set(root.dir = "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

options(stringsAsFactors = FALSE)
knitr::opts_chunk$set(error = FALSE)
options(future.globals.maxSize = 1591289600)
```

# Setup

```{r setUpDirectories}
projNumber <- "P697"
baseDir <- "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2"
dataInputDir <- file.path(baseDir, "data/inputData")
plotDir <- file.path(baseDir, "figures")
dataOutputDir <- file.path(baseDir, "data/outputData")
tableDir <- file.path(baseDir, "tables")
if (!dir.exists(tableDir)) dir.create(tableDir, recursive = TRUE)
dataDate <- Sys.Date()
filenameSuffix <- paste0(projNumber, ".", dataDate)
```

```{r setupFunctions}
suppressPackageStartupMessages(library(theTools))
# theTools's savePlot takes its figure-name prefix from this option rather
# than reading a `filenameSuffix` global out of the calling frame, which is what
# the retired local copy did. Without it every figure silently loses its
# "<proj>.<date>_" prefix, writes to a new name, and stops overwriting the
# previous run's output.
options(theTools.filenameSuffix = filenameSuffix)
```

```{r analysisHelpers}
phaseLevels <- c("Baseline", "Post-IL2", "Post-RAPA", "Followup")
primaryContrastDefs <- list(
  PostIL2_vs_BL = c("Post-IL2", "Baseline"),
  PostRAPA_vs_BL = c("Post-RAPA", "Baseline"),
  Followup_vs_BL = c("Followup", "Baseline"),
  Followup_vs_PostRAPA = c("Followup", "Post-RAPA")
)

integrationSettings <- list(
  rnaReduction = "harmony",
  rnaDims = 1:18,
  atacReduction = "harmony_lsi",
  atacDims = 1:29,
  wnnReductions = c("harmony", "harmony_lsi")
)

saveComplexHeatmap <- function(ht, plotDir, filename, width, height) {
  pdf(file.path(plotDir, paste0(filenameSuffix, "_", filename, ".pdf")),
    width = width, height = height)
  ComplexHeatmap::draw(ht)
  dev.off()

  png(file.path(plotDir, paste0(filenameSuffix, "_", filename, ".png")),
    width = width, height = height, units = "in", res = 600)
  ComplexHeatmap::draw(ht)
  dev.off()
}

aggregateCountsByGroups <- function(countMat, metadata, groupCols) {
  groupMeta <- metadata[, groupCols, drop = FALSE]
  groupMeta[] <- lapply(groupMeta, as.character)
  groupMeta$groupID <- apply(groupMeta[, groupCols, drop = FALSE], 1, paste,
    collapse = "__")

  groups <- unique(groupMeta$groupID)
  aggMat <- matrix(0, nrow = nrow(countMat), ncol = length(groups),
    dimnames = list(rownames(countMat), groups))

  for (grp in groups) {
    idx <- which(groupMeta$groupID == grp)
    if (length(idx) == 1) {
      aggMat[, grp] <- as.numeric(countMat[, idx])
    } else {
      aggMat[, grp] <- Matrix::rowSums(countMat[, idx, drop = FALSE])
    }
  }

  groupSummary <- tibble::as_tibble(groupMeta) %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c(groupCols, "groupID")))) %>%
    dplyr::summarise(nCells = dplyr::n(), .groups = "drop")
  groupSummary <- groupSummary[match(colnames(aggMat), groupSummary$groupID), , drop = FALSE]

  list(counts = aggMat, groupMeta = groupSummary)
}

mostFrequentValue <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  names(sort(table(x), decreasing = TRUE))[1]
}

buildClusterCoverageTable <- function(seuratObj, clusterCol = "wnn_clusters") {
  meta <- seuratObj@meta.data
  donorLevels <- sort(unique(as.character(meta$donorID)))
  clusterLevels <- gtools::mixedsort(unique(as.character(meta[[clusterCol]])))

  coverageLong <- meta %>%
    dplyr::transmute(
      donorID = factor(as.character(donorID), levels = donorLevels),
      treatmentPhase = factor(as.character(treatmentPhase), levels = phaseLevels),
      cluster = factor(as.character(.data[[clusterCol]]), levels = clusterLevels)
    ) %>%
    dplyr::count(cluster, treatmentPhase, donorID, name = "nCells") %>%
    tidyr::complete(
      cluster = factor(clusterLevels, levels = clusterLevels),
      treatmentPhase = factor(phaseLevels, levels = phaseLevels),
      donorID = factor(donorLevels, levels = donorLevels),
      fill = list(nCells = 0)
    )

  coverageSummary <- coverageLong %>%
    dplyr::group_by(cluster, treatmentPhase) %>%
    dplyr::summarise(
      nCells = sum(nCells),
      nDonors = sum(nCells > 0),
      medianCellsPerDonor = ifelse(any(nCells > 0), median(nCells[nCells > 0]), NA_real_),
      minCellsPerDonor = ifelse(any(nCells > 0), min(nCells[nCells > 0]), NA_real_),
      maxCellsPerDonor = ifelse(any(nCells > 0), max(nCells[nCells > 0]), NA_real_),
      .groups = "drop"
    )

  list(long = coverageLong, summary = coverageSummary)
}

summarizePrimaryContrastReadiness <- function(coverageLong, contrastDefs) {
  bind_rows(lapply(names(contrastDefs), function(nm) {
    phase1 <- contrastDefs[[nm]][1]
    phase2 <- contrastDefs[[nm]][2]

    coverageLong %>%
      dplyr::filter(treatmentPhase %in% c(phase1, phase2)) %>%
      dplyr::mutate(treatmentPhase = as.character(treatmentPhase)) %>%
      tidyr::pivot_wider(names_from = treatmentPhase, values_from = nCells,
        values_fill = 0) %>%
      dplyr::group_by(cluster) %>%
      dplyr::summarise(
        contrast = nm,
        phase1 = phase1,
        phase2 = phase2,
        nDonorsPhase1 = sum(.data[[phase1]] > 0),
        nDonorsPhase2 = sum(.data[[phase2]] > 0),
        nDonorPairsWithCells = sum(.data[[phase1]] > 0 & .data[[phase2]] > 0),
        totalCellsPhase1 = sum(.data[[phase1]]),
        totalCellsPhase2 = sum(.data[[phase2]]),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        readyForPairedDA = nDonorPairsWithCells >= 2,
        limitingPhase = dplyr::case_when(
          nDonorsPhase1 < nDonorsPhase2 ~ phase1,
          nDonorsPhase2 < nDonorsPhase1 ~ phase2,
          TRUE ~ "Balanced"
        )
      )
  }))
}

plotClusterCoverageHeatmap <- function(coverageSummary, title) {
  ggplot(coverageSummary, aes(x = treatmentPhase, y = cluster, fill = nDonors)) +
    geom_tile(color = "white") +
    geom_text(aes(label = paste0(nDonors, "D / ", nCells, "C")), size = 3) +
    scale_fill_viridis_c(option = "C", name = "# Donors") +
    labs(
      x = "Treatment Phase",
      y = "WNN Cluster",
      title = title,
      subtitle = "Tile labels show donors represented and total cells"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

plotPrimaryContrastReadiness <- function(readinessDf, title) {
  ggplot(readinessDf,
    aes(x = contrast, y = cluster, fill = nDonorPairsWithCells)) +
    geom_tile(color = "white") +
    geom_text(aes(label = ifelse(readyForPairedDA,
      paste0(nDonorPairsWithCells, " pairs"),
      paste0(nDonorPairsWithCells, " pairs*"))), size = 3) +
    scale_fill_viridis_c(option = "B", name = "# Donor pairs") +
    labs(
      x = "Primary Contrast",
      y = "WNN Cluster",
      title = title,
      subtitle = "* fewer than 2 paired donors with cells in both phases"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

buildClusterAbundanceTable <- function(seuratObj, clusterCol = "wnn_clusters") {
  meta <- seuratObj@meta.data
  donorLevels <- sort(unique(as.character(meta$donorID)))
  clusterLevels <- gtools::mixedsort(unique(as.character(meta[[clusterCol]])))

  meta %>%
    dplyr::transmute(
      donorID = factor(as.character(donorID), levels = donorLevels),
      treatmentPhase = factor(as.character(treatmentPhase), levels = phaseLevels),
      cluster = factor(as.character(.data[[clusterCol]]), levels = clusterLevels)
    ) %>%
    dplyr::count(donorID, treatmentPhase, cluster, name = "nCells") %>%
    tidyr::complete(
      donorID = factor(donorLevels, levels = donorLevels),
      treatmentPhase = factor(phaseLevels, levels = phaseLevels),
      cluster = factor(clusterLevels, levels = clusterLevels),
      fill = list(nCells = 0)
    ) %>%
    dplyr::group_by(donorID, treatmentPhase) %>%
    dplyr::mutate(
      totalCells = sum(nCells),
      proportion = ifelse(totalCells > 0, nCells / totalCells, NA_real_)
    ) %>%
    dplyr::ungroup()
}

plotClusterAbundanceTrajectories <- function(abundanceDf, title) {
  validDf <- abundanceDf %>%
    dplyr::filter(!is.na(proportion), totalCells > 0)

  summaryDf <- validDf %>%
    dplyr::group_by(cluster, treatmentPhase) %>%
    dplyr::summarise(
      medianProportion = median(proportion),
      nDonors = dplyr::n_distinct(donorID),
      .groups = "drop"
    )

  ggplot(validDf, aes(x = treatmentPhase, y = proportion, group = donorID)) +
    geom_line(color = "grey75", linewidth = 0.4, alpha = 0.8) +
    geom_point(color = "grey45", size = 0.8, alpha = 0.8) +
    geom_line(data = summaryDf,
      aes(x = treatmentPhase, y = medianProportion, group = 1),
      inherit.aes = FALSE, color = "#8B0000", linewidth = 0.9) +
    geom_point(data = summaryDf,
      aes(x = treatmentPhase, y = medianProportion),
      inherit.aes = FALSE, color = "#8B0000", size = 1.6) +
    facet_wrap(~cluster, scales = "free_y") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      x = "Treatment Phase",
      y = "Within-donor cluster proportion",
      title = title,
      subtitle = "Grey = individual donors, red = median across donors"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

buildClusterMappingTables <- function(metadata, referenceCol, referenceName,
                                      wnnCol = "wnn_clusters") {
  ct <- table(
    referenceCluster = metadata[[referenceCol]],
    wnnCluster = metadata[[wnnCol]]
  )

  counts <- as.data.frame(ct)
  colnames(counts) <- c("referenceCluster", "wnnCluster", "Count")
  counts$referenceModality <- referenceName

  proportions <- as.data.frame(prop.table(ct, margin = 1))
  colnames(proportions) <- c("referenceCluster", "wnnCluster", "Proportion")
  proportions$referenceModality <- referenceName

  list(counts = counts, proportions = proportions)
}

plotClusterMappingHeatmap <- function(proportionDf, title, yLabel, highColor) {
  ggplot(proportionDf, aes(x = wnnCluster, y = referenceCluster, fill = Proportion)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.0f%%", 100 * Proportion)),
      size = 3, color = "black") +
    scale_fill_gradient(low = "white", high = highColor, limits = c(0, 1)) +
    labs(x = "WNN Cluster", y = yLabel, title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

buildPseudobulkHeatmapData <- function(seuratObj, genes, groupCols) {
  counts <- GetAssayData(seuratObj, layer = "counts")
  genesPresent <- intersect(genes, rownames(counts))
  if (length(genesPresent) == 0) return(NULL)

  agg <- aggregateCountsByGroups(
    countMat = counts[genesPresent, , drop = FALSE],
    metadata = seuratObj@meta.data,
    groupCols = groupCols
  )

  ## Drop pseudobulk samples with zero total counts (empty group combos)
  libSizes <- colSums(agg$counts)
  validCols <- is.finite(libSizes) & libSizes > 0
  if (sum(validCols) == 0) return(NULL)
  agg$counts <- agg$counts[, validCols, drop = FALSE]
  agg$groupMeta <- agg$groupMeta[validCols, , drop = FALSE]

  logCpm <- edgeR::cpm(agg$counts, log = TRUE, prior.count = 1)
  matScaled <- t(scale(t(logCpm)))
  matScaled[is.na(matScaled)] <- 0

  list(matrix = matScaled, annotation = agg$groupMeta)
}

buildNamedAnnotationPalette <- function(values, paletteValues = NULL) {
  valueLevels <- if (is.factor(values)) {
    levels(values)
  } else {
    unique(as.character(values))
  }
  valueLevels <- valueLevels[!is.na(valueLevels)]

  if (length(valueLevels) == 0) return(character(0))

  fallback <- setNames(scales::hue_pal()(length(valueLevels)), valueLevels)

  if (is.null(paletteValues) || length(paletteValues) == 0) {
    return(fallback)
  }

  paletteValues <- paletteValues[!is.na(paletteValues)]
  paletteNames <- names(paletteValues)

  if (is.null(paletteNames) || any(paletteNames == "")) {
    fallback[valueLevels[seq_len(min(length(paletteValues), length(valueLevels)))]] <-
      unname(paletteValues)[seq_len(min(length(paletteValues), length(valueLevels)))]
    return(fallback)
  }

  matchingLevels <- intersect(valueLevels, paletteNames)
  if (length(matchingLevels) > 0) {
    fallback[matchingLevels] <- paletteValues[matchingLevels]
  }

  fallback
}

normalizeAnnotationColors <- function(annotationDf, annotationColors) {
  normalized <- lapply(names(annotationColors), function(colName) {
    if (!colName %in% colnames(annotationDf)) return(character(0))
    buildNamedAnnotationPalette(annotationDf[[colName]], annotationColors[[colName]])
  })
  names(normalized) <- names(annotationColors)
  normalized[vapply(normalized, length, integer(1)) > 0]
}

drawPseudobulkHeatmap <- function(matScaled, annotationDf, annotationColors,
                                  splitCol, title, plotDir, filename,
                                  width, height) {
  if (is.null(matScaled) || ncol(matScaled) == 0) return(invisible(NULL))

  matScaled[matScaled > 2.5] <- 2.5
  matScaled[matScaled < -2.5] <- -2.5
  annotationColors <- normalizeAnnotationColors(annotationDf, annotationColors)

  htAnnot <- ComplexHeatmap::HeatmapAnnotation(
    df = annotationDf,
    col = annotationColors,
    show_annotation_name = TRUE,
    annotation_name_side = "left"
  )

  ht <- ComplexHeatmap::Heatmap(matScaled,
    name = "z-score",
    top_annotation = htAnnot,
    column_split = annotationDf[[splitCol]],
    cluster_columns = FALSE,
    show_column_names = FALSE,
    row_names_gp = grid::gpar(fontsize = 8),
    column_title_gp = grid::gpar(fontsize = 10),
    column_title_rot = 0,
    col = colorRamp2(c(-2.5, 0, 2.5), c("blue", "white", "red")),
    column_title = title,
    heatmap_legend_param = list(title = "z-score"),
    use_raster = TRUE,
    raster_quality = 3
  )

  ComplexHeatmap::draw(ht)
  saveComplexHeatmap(ht, plotDir, filename, width, height)
}

buildClusterPhenotypeSummary <- function(seuratObj, markers,
                                         clusterCol = "wnn_clusters") {
  meta <- seuratObj@meta.data
  clusterLevels <- gtools::mixedsort(unique(as.character(meta[[clusterCol]])))

  bind_rows(lapply(clusterLevels, function(cl) {
    clMeta <- meta[as.character(meta[[clusterCol]]) == cl, , drop = FALSE]
    topMarkers <- markers %>%
      dplyr::filter(cluster == cl, p_val_adj < 0.05) %>%
      dplyr::arrange(dplyr::desc(avg_log2FC), p_val_adj) %>%
      dplyr::slice_head(n = 10)

    data.frame(
      cluster = cl,
      nCells = nrow(clMeta),
      nDonors = dplyr::n_distinct(clMeta$donorID),
      dominantCellTypistL2 = if ("cellTypistCallingL2" %in% colnames(clMeta)) {
        mostFrequentValue(clMeta$cellTypistCallingL2)
      } else {
        NA_character_
      },
      dominantCellTypistL3 = if ("cellTypistCallingL3" %in% colnames(clMeta)) {
        mostFrequentValue(clMeta$cellTypistCallingL3)
      } else {
        NA_character_
      },
      topRNAMarkers = if (nrow(topMarkers) > 0) {
        paste(unique(topMarkers$gene), collapse = "; ")
      } else {
        NA_character_
      },
      medianRNAWeight = if ("RNA.weight" %in% colnames(clMeta)) {
        median(clMeta$RNA.weight, na.rm = TRUE)
      } else {
        NA_real_
      },
      medianATACWeight = if ("ATAC.weight" %in% colnames(clMeta)) {
        median(clMeta$ATAC.weight, na.rm = TRUE)
      } else {
        NA_real_
      },
      stringsAsFactors = FALSE
    )
  }))
}

buildIl2Stat5Summary <- function(seuratObj, clusterCol = "wnn_clusters",
                                 scoreCol = "IL2_STAT5_score1") {
  meta <- seuratObj@meta.data %>%
    dplyr::transmute(
      donorID = factor(as.character(donorID), levels = sort(unique(as.character(donorID)))),
      treatmentPhase = factor(as.character(treatmentPhase), levels = phaseLevels),
      cluster = factor(as.character(.data[[clusterCol]]),
        levels = gtools::mixedsort(unique(as.character(.data[[clusterCol]])))),
      score = .data[[scoreCol]]
    ) %>%
    dplyr::filter(!is.na(score))

  donorPhaseCluster <- meta %>%
    dplyr::group_by(cluster, donorID, treatmentPhase) %>%
    dplyr::summarise(
      nCells = dplyr::n(),
      meanScore = mean(score),
      medianScore = median(score),
      .groups = "drop"
    )

  phaseSummary <- donorPhaseCluster %>%
    dplyr::group_by(cluster, treatmentPhase) %>%
    dplyr::summarise(
      nDonors = dplyr::n_distinct(donorID),
      medianDonorScore = median(meanScore),
      meanDonorScore = mean(meanScore),
      .groups = "drop"
    )

  list(donorPhaseCluster = donorPhaseCluster, phaseSummary = phaseSummary)
}

plotIl2Stat5Heatmap <- function(phaseSummary, title) {
  ggplot(phaseSummary,
    aes(x = treatmentPhase, y = cluster, fill = medianDonorScore)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f\nn=%d", medianDonorScore, nDonors)),
      size = 3) +
    scale_fill_viridis_c(option = "D", name = "Median donor score",
      na.value = "grey95") +
    labs(
      x = "Treatment Phase",
      y = "WNN Cluster",
      title = title,
      subtitle = "Tile labels show median donor score and donors represented"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

plotIl2Stat5Trajectories <- function(donorSummary, title) {
  summaryDf <- donorSummary %>%
    dplyr::group_by(cluster, treatmentPhase) %>%
    dplyr::summarise(medianMeanScore = median(meanScore), .groups = "drop")

  ggplot(donorSummary, aes(x = treatmentPhase, y = meanScore, group = donorID)) +
    geom_line(color = "grey75", linewidth = 0.4, alpha = 0.8) +
    geom_point(color = "grey45", size = 0.8, alpha = 0.8) +
    geom_line(data = summaryDf,
      aes(x = treatmentPhase, y = medianMeanScore, group = 1),
      inherit.aes = FALSE, color = "#8B0000", linewidth = 0.9) +
    geom_point(data = summaryDf,
      aes(x = treatmentPhase, y = medianMeanScore),
      inherit.aes = FALSE, color = "#8B0000", size = 1.6) +
    facet_wrap(~cluster, scales = "free_y") +
    labs(
      x = "Treatment Phase",
      y = "Mean IL2-STAT5 score",
      title = title,
      subtitle = "Grey = individual donors, red = median across donors"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

addDerivedQCMetrics <- function(seuratObj) {
  meta <- seuratObj@meta.data

  if (!"complexity" %in% colnames(meta) &&
    all(c("nFeature_RNA", "nCount_RNA") %in% colnames(meta))) {
    seuratObj[["complexity"]] <- log10(seuratObj$nFeature_RNA) / log10(seuratObj$nCount_RNA)
  }

  if (!"tssEnrichment" %in% colnames(meta) &&
    all(c("cr_arc_atac_TSS_fragments", "cr_arc_atac_fragments") %in% colnames(meta))) {
    seuratObj[["tssEnrichment"]] <- seuratObj$cr_arc_atac_TSS_fragments /
      (seuratObj$cr_arc_atac_fragments + 1) * 100
  }

  if (!"frip" %in% colnames(meta) &&
    all(c("cr_arc_atac_peak_region_fragments", "cr_arc_atac_fragments") %in% colnames(meta))) {
    seuratObj[["frip"]] <- seuratObj$cr_arc_atac_peak_region_fragments /
      (seuratObj$cr_arc_atac_fragments + 1) * 100
  }

  if (!"atacMitoPct" %in% colnames(meta) &&
    all(c("cr_arc_atac_mitochondrial_reads", "cr_arc_atac_fragments") %in% colnames(meta))) {
    seuratObj[["atacMitoPct"]] <- seuratObj$cr_arc_atac_mitochondrial_reads /
      (seuratObj$cr_arc_atac_fragments + 1) * 100
  }

  seuratObj
}

qcMetricSpecs <- list(
  list(column = "nCount_RNA", label = "RNA UMI count", transform = "log10p"),
  list(column = "nFeature_RNA", label = "RNA genes detected", transform = "log10p"),
  list(column = "percentMt", label = "% mitochondrial RNA", transform = "identity"),
  list(column = "percentRibo", label = "% ribosomal RNA", transform = "identity"),
  list(column = "complexity", label = "RNA complexity", transform = "identity"),
  list(column = "pctCountsInTop20Genes", label = "% counts in top 20 genes", transform = "identity"),
  list(column = "nCount_ATAC", label = "ATAC peak counts", transform = "log10p"),
  list(column = "cr_arc_atac_fragments", label = "ATAC fragments", transform = "log10p"),
  list(column = "tssEnrichment", label = "TSS enrichment (%)", transform = "identity"),
  list(column = "frip", label = "FRiP (%)", transform = "identity"),
  list(column = "atacMitoPct", label = "% ATAC mitochondrial reads", transform = "identity"),
  list(column = "nucleosomeSignal", label = "Nucleosome signal", transform = "log10p"),
  list(column = "nfrFragments", label = "NFR fragments", transform = "log10p"),
  list(column = "monoFragments", label = "Mononucleosomal fragments", transform = "log10p")
)

buildQCMetricPlot <- function(seuratObj, metricSpec, compartment) {
  metricCol <- metricSpec$column
  metricVals <- seuratObj@meta.data[[metricCol]]
  if (is.null(metricVals) || sum(!is.na(metricVals)) < 10) return(NULL)

  plotVals <- if (identical(metricSpec$transform, "log10p")) {
    log10(metricVals + 1)
  } else {
    metricVals
  }

  finiteVals <- plotVals[is.finite(plotVals)]
  if (length(finiteVals) < 10) return(NULL)

  limits <- stats::quantile(finiteVals, probs = c(0.01, 0.99), na.rm = TRUE)
  if (!all(is.finite(limits)) || diff(limits) <= 0) {
    limits <- range(finiteVals, na.rm = TRUE)
  }
  if (!all(is.finite(limits)) || diff(limits) <= 0) return(NULL)

  reductionInfo <- list(
    list(name = "umap.rna", title = "RNA UMAP"),
    list(name = "umap.atac", title = "ATAC UMAP"),
    list(name = "umap.wnn", title = "WNN UMAP")
  )

  legendTitle <- if (identical(metricSpec$transform, "log10p")) {
    paste0("log10(", metricSpec$label, " + 1)")
  } else {
    metricSpec$label
  }

  plotList <- lapply(reductionInfo, function(red) {
    if (!red$name %in% names(seuratObj@reductions)) return(NULL)

    emb <- Embeddings(seuratObj, red$name)[, 1:2, drop = FALSE]
    plotDf <- data.frame(
      dim1 = emb[, 1],
      dim2 = emb[, 2],
      metricValue = plotVals,
      stringsAsFactors = FALSE
    )

    ggplot(plotDf, aes(x = dim1, y = dim2, color = metricValue)) +
      geom_point(size = 0.25, alpha = 0.8) +
      scale_color_viridis_c(
        option = "D",
        limits = limits,
        oob = scales::squish,
        na.value = "grey85",
        name = legendTitle
      ) +
      labs(
        x = NULL,
        y = NULL,
        title = paste(compartment, "—", red$title),
        subtitle = metricSpec$label
      ) +
      theme(
        axis.text = element_blank(),
        axis.ticks = element_blank()
      ) +
      coord_equal()
  })

  plotList <- Filter(Negate(is.null), plotList)
  if (length(plotList) == 0) return(NULL)

  wrap_plots(plotList, ncol = length(plotList))
}
```

# Load Peak Calling Bundles

```{r loadBundles}
compartments <- c("NK", "Treg")
seuratList <- list()
macs2PeaksList <- list()
sourcePeakCallingBundles <- list()

for (comp in compartments) {
  pattern <- sprintf("P697\\..*_%s_peakCallingBundle\\.rds$", comp)
  bundleFiles <- list.files(dataOutputDir, pattern = pattern, full.names = TRUE)
  if (length(bundleFiles) == 0) {
    stop(sprintf("No %s peak-calling bundle found in data/outputData", comp))
  }
  bundleFile <- bundleFiles[which.max(file.mtime(bundleFiles))]
  cat(sprintf("Loading %s bundle: %s\n", comp, basename(bundleFile)))
  sourcePeakCallingBundles[[comp]] <- basename(bundleFile)

  bundle <- readRDS(bundleFile)

  objName <- paste0("seurat", comp)
  seuratList[[comp]] <- bundle[[objName]]
  macs2PeaksList[[comp]] <- bundle$macs2Peaks

  if (comp == compartments[1]) {
    tcrGeneList <- bundle$tcrGeneList
    palDonorId <- bundle$palDonorId
    palVisit <- bundle$palVisit
  }

  cat(sprintf("  %s: %d cells, assays: %s\n", comp,
    ncol(seuratList[[comp]]),
    paste(names(seuratList[[comp]]@assays), collapse = ", ")))
  cat(sprintf("  Reductions: %s\n",
    paste(names(seuratList[[comp]]@reductions), collapse = ", ")))

  rm(bundle)
  gc(verbose = FALSE)
}
```

# Create Treatment Phase Variable

```{r createTreatmentPhase}
# Map visitNum to treatment phase based on trial design:
#   LD IL-2 treatment → stopped → RAPA treatment → stopped → Followup
#   V0:  Baseline (pre-treatment)
#   V4/5: End of IL-2 phase (Post-IL2)
#   V8/10: End of RAPA phase (Post-RAPA)
#   V12: ~6 months after all treatment cessation (Followup)
treatmentPhaseMap <- c(
  "0" = "Baseline",
  "4" = "Post-IL2", "5" = "Post-IL2",
  "8" = "Post-RAPA", "10" = "Post-RAPA",
  "12" = "Followup"
)

palTreatmentPhase <- c(
  "Baseline" = "#4DAF4A",
  "Post-IL2" = "#377EB8",
  "Post-RAPA" = "#FF7F00",
  "Followup" = "#984EA3"
)

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  seuratObj$treatmentPhase <- factor(
    unname(treatmentPhaseMap[as.character(seuratObj$visitNum)]),
    levels = c("Baseline", "Post-IL2", "Post-RAPA", "Followup")
  )

  cat(sprintf("\n%s treatmentPhase distribution:\n", comp))
  print(table(seuratObj$treatmentPhase, seuratObj$donorID))

  nNA <- sum(is.na(seuratObj$treatmentPhase))
  if (nNA > 0) cat(sprintf("  WARNING: %d cells with unmapped visitNum\n", nNA))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
```

# Recompute RNA UMAP on Subset

```{r recomputeRNAUMAP}
# The RNA UMAP from preprocessing was computed on all cells (including
# non-ATAC-passing). After subsetting in peak calling, coordinates are stale.
# Recompute on the harmony embedding for fair comparison with ATAC and WNN UMAPs.
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  set.seed(6022)
  seuratObj <- RunUMAP(
    seuratObj,
    reduction = integrationSettings$rnaReduction,
    dims = integrationSettings$rnaDims,
    reduction.name = "umap.rna",
    reduction.key = "rnaUMAP_",
    verbose = FALSE
  )

  cat(sprintf("%s: RNA UMAP recomputed (%d cells)\n", comp, ncol(seuratObj)))
  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

# Recompute ATAC UMAP on Integrated ATAC Embedding

```{r recomputeATACUMAP, cache = FALSE}
# Recompute the ATAC UMAP from harmony_lsi here so all three displayed embeddings
# in this script are derived from the integrated reductions used for clustering.
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  set.seed(6022)
  seuratObj <- RunUMAP(
    seuratObj,
    reduction = integrationSettings$atacReduction,
    dims = integrationSettings$atacDims,
    reduction.name = "umap.atac",
    reduction.key = "atacUMAP_",
    verbose = FALSE
  )

  cat(sprintf("%s: ATAC UMAP recomputed from %s (%d cells)\n",
    comp, integrationSettings$atacReduction, ncol(seuratObj)))
  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

# WNN Integration

```{r wnnIntegration}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  cat(sprintf("\n%s: Running FindMultiModalNeighbors...\n", comp))
  cat(sprintf("  RNA: %s (dims %s)\n",
    integrationSettings$rnaReduction,
    paste(range(integrationSettings$rnaDims), collapse = ":")))
  cat(sprintf("  ATAC: %s (dims %s)\n",
    integrationSettings$atacReduction,
    paste(range(integrationSettings$atacDims), collapse = ":")))

  seuratObj <- FindMultiModalNeighbors(
    seuratObj,
    reduction.list = as.list(integrationSettings$wnnReductions),
    dims.list = list(integrationSettings$rnaDims, integrationSettings$atacDims),
    modality.weight.name = c("RNA.weight", "ATAC.weight"),
    verbose = FALSE
  )

  cat(sprintf("  Median RNA weight: %.3f\n", median(seuratObj$RNA.weight)))
  cat(sprintf("  Median ATAC weight: %.3f\n", median(seuratObj$ATAC.weight)))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

# WNN UMAP

```{r wnnUMAP}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  set.seed(6022)
  seuratObj <- RunUMAP(
    seuratObj,
    nn.name = "weighted.nn",
    reduction.name = "umap.wnn",
    reduction.key = "wnnUMAP_",
    verbose = FALSE
  )

  cat(sprintf("%s: WNN UMAP computed\n", comp))
  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

# WNN Clustering Evaluation

```{r evaluateResolutions}
resolutions <- seq(0.05, 0.5, by = 0.05)
evalResults <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  # Build combined scaled embedding for silhouette computation
  rnaEmbed <- scale(Embeddings(seuratObj, "harmony")[, 1:18])
  atacEmbed <- scale(Embeddings(seuratObj, "harmony_lsi")[, 1:29])
  combinedEmbed <- cbind(rnaEmbed, atacEmbed)
  rm(rnaEmbed, atacEmbed)

  # Subsample for efficiency
  set.seed(42)
  nSample <- min(3000, nrow(combinedEmbed))
  sampleIdx <- sample(nrow(combinedEmbed), nSample)
  distSub <- dist(combinedEmbed[sampleIdx, ])
  rm(combinedEmbed)

  metrics <- data.frame(
    resolution = resolutions,
    nClusters = NA_integer_,
    meanSilhouette = NA_real_
  )

  cat(sprintf("\n%s: evaluating %d resolutions...\n", comp, length(resolutions)))
  for (j in seq_along(resolutions)) {
    res <- resolutions[j]
    seuratObj <- FindClusters(seuratObj, graph.name = "wsnn",
      resolution = res, verbose = FALSE)
    clusters <- as.integer(seuratObj$seurat_clusters)
    clustersSub <- clusters[sampleIdx]
    nUnique <- length(unique(clustersSub))

    metrics$nClusters[j] <- length(unique(clusters))

    if (nUnique > 1 && nUnique < nSample) {
      sil <- cluster::silhouette(clustersSub, distSub)
      metrics$meanSilhouette[j] <- mean(sil[, 3])
    }

    cat(sprintf("  res %.2f: %d clusters, sil %.3f\n",
      res, metrics$nClusters[j],
      ifelse(is.na(metrics$meanSilhouette[j]), NA, metrics$meanSilhouette[j])))
  }

  evalResults[[comp]] <- metrics
  seuratList[[comp]] <- seuratObj
  rm(distSub)
}
rm(seuratObj)
gc(verbose = FALSE)
```

```{r clusterMetricsPlot, fig.height = 5, fig.width = 12}
allMetrics <- bind_rows(
  lapply(compartments, function(comp) {
    evalResults[[comp]] %>% mutate(compartment = comp)
  })
)

pSil <- ggplot(allMetrics, aes(x = resolution, y = meanSilhouette, color = compartment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("NK" = "#E41A1C", "Treg" = "#377EB8")) +
  labs(x = "Resolution", y = "Mean Silhouette Width",
    title = "WNN Clustering: Silhouette") +
  theme(legend.position = "bottom")

pNCl <- ggplot(allMetrics, aes(x = resolution, y = nClusters, color = compartment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("NK" = "#E41A1C", "Treg" = "#377EB8")) +
  labs(x = "Resolution", y = "Number of Clusters",
    title = "WNN Clustering: Cluster Count") +
  theme(legend.position = "bottom")

pSil | pNCl

savePlot(plot = pSil | pNCl, plotDir = plotDir,
  filename = "WNN_clusterMetrics", height = 5, width = 12)

for (comp in compartments) {
  bestIdx <- which.max(evalResults[[comp]]$meanSilhouette)
  cat(sprintf("%s best silhouette: res %.2f → %d clusters (sil = %.3f)\n",
    comp, evalResults[[comp]]$resolution[bestIdx],
    evalResults[[comp]]$nClusters[bestIdx],
    evalResults[[comp]]$meanSilhouette[bestIdx]))
}
```

# Select Final Resolution

```{r selectResolution, cache = FALSE}
# Default: resolution with highest mean silhouette.
# Override by uncommenting the manual lines below after inspecting the metrics.
# The counts printed here refer only to WNN clustering on the wsnn graph.
# RNA-only and ATAC-only clustering below reuse the same numeric resolution for
# comparability, but they operate on different graphs and are not expected to
# return the same number of clusters.
selectedRes <- list()
for (comp in compartments) {
  bestIdx <- which.max(evalResults[[comp]]$meanSilhouette)
  selectedRes[[comp]] <- evalResults[[comp]]$resolution[bestIdx]
}

# Manual override (uncomment to use):
# selectedRes[["NK"]] <- 0.05
# selectedRes[["Treg"]] <- 0.05 # 0.05

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  seuratObj <- FindClusters(seuratObj, graph.name = "wsnn",
    resolution = selectedRes[[comp]], verbose = FALSE)

  seuratObj$wnn_clusters <- seuratObj$seurat_clusters

  cat(sprintf("\n%s: %d WNN clusters at resolution %.2f\n",
    comp, length(unique(seuratObj$wnn_clusters)), selectedRes[[comp]]))
  print(table(seuratObj$wnn_clusters))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

# WNN Cluster Coverage and Primary Contrast Readiness

```{r clusterCoverageReadiness, fig.height = 10, fig.width = 10, cache = FALSE}
clusterCoverage <- list()
clusterReadiness <- list()

for (comp in compartments) {
  coverageObj <- buildClusterCoverageTable(seuratList[[comp]])
  readinessDf <- summarizePrimaryContrastReadiness(
    coverageObj$long,
    primaryContrastDefs
  )

  clusterCoverage[[comp]] <- coverageObj
  clusterReadiness[[comp]] <- readinessDf

  write.csv(
    coverageObj$long,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_WNN_cluster_donor_phase_cells.csv")),
    row.names = FALSE
  )
  write.csv(
    coverageObj$summary,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_WNN_cluster_phase_summary.csv")),
    row.names = FALSE
  )
  write.csv(
    readinessDf,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_WNN_primary_contrast_readiness.csv")),
    row.names = FALSE
  )

  pCoverage <- plotClusterCoverageHeatmap(
    coverageObj$summary,
    paste(comp, "— WNN Cluster Coverage by Treatment Phase")
  )
  pReadiness <- plotPrimaryContrastReadiness(
    readinessDf,
    paste(comp, "— Primary Contrast Readiness by WNN Cluster")
  )

  combined <- pCoverage / pReadiness
  print(combined)
  savePlot(
    plot = combined,
    plotDir = plotDir,
    filename = paste0(comp, "_WNN_cluster_readiness"),
    height = 10,
    width = 10
  )
}
```

# RNA-Only and ATAC-Only Clustering

```{r rnaClustering, cache = FALSE}
# RNA-only clustering using the same numeric resolution as WNN for comparability.
# The cluster count can differ from WNN because the RNA_snn graph is built from
# the harmony RNA reduction alone, not the wsnn graph.
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  seuratObj <- FindNeighbors(seuratObj,
    reduction = integrationSettings$rnaReduction,
    dims = integrationSettings$rnaDims,
    graph.name = c("RNA_nn", "RNA_snn"), verbose = FALSE)
  seuratObj <- FindClusters(seuratObj, graph.name = "RNA_snn",
    resolution = selectedRes[[comp]], verbose = FALSE)
  seuratObj$rna_clusters <- seuratObj$seurat_clusters

  cat(sprintf("%s: %d RNA clusters at resolution %.2f\n",
    comp, length(unique(seuratObj$rna_clusters)), selectedRes[[comp]]))
  print(table(seuratObj$rna_clusters))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

```{r atacClustering, cache = FALSE}
# ATAC-only clustering using the same numeric resolution as WNN for comparability.
# The cluster count can differ from WNN because the ATAC_snn graph is built from
# harmony_lsi alone, not the wsnn graph.
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  seuratObj <- FindNeighbors(seuratObj,
    reduction = integrationSettings$atacReduction,
    dims = integrationSettings$atacDims,
    graph.name = c("ATAC_nn", "ATAC_snn"), verbose = FALSE)
  seuratObj <- FindClusters(seuratObj, graph.name = "ATAC_snn",
    resolution = selectedRes[[comp]], verbose = FALSE)
  seuratObj$atac_clusters <- seuratObj$seurat_clusters

  cat(sprintf("%s: %d ATAC clusters at resolution %.2f\n",
    comp, length(unique(seuratObj$atac_clusters)), selectedRes[[comp]]))
  print(table(seuratObj$atac_clusters))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
gc(verbose = FALSE)
```

# Cluster Count Summary at Selected Resolution

```{r clusterCountSummary, fig.height = 4, fig.width = 9, cache = FALSE}
clusterCountSummary <- list()

for (comp in compartments) {
  summaryDf <- data.frame(
    compartment = comp,
    clustering = c("WNN", "RNA", "ATAC"),
    graph = c("wsnn", "RNA_snn", "ATAC_snn"),
    reduction = c(
      paste(integrationSettings$wnnReductions, collapse = " + "),
      integrationSettings$rnaReduction,
      integrationSettings$atacReduction
    ),
    resolution = selectedRes[[comp]],
    nClusters = c(
      length(unique(seuratList[[comp]]$wnn_clusters)),
      length(unique(seuratList[[comp]]$rna_clusters)),
      length(unique(seuratList[[comp]]$atac_clusters))
    ),
    stringsAsFactors = FALSE
  )
  clusterCountSummary[[comp]] <- summaryDf

  write.csv(
    summaryDf,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_cluster_count_summary.csv")),
    row.names = FALSE
  )

  cat(sprintf("\n%s clustering summary at resolution %.2f:\n",
    comp, selectedRes[[comp]]))
  print(summaryDf[, c("clustering", "graph", "reduction", "nClusters")])
}

clusterCountSummaryDf <- bind_rows(clusterCountSummary)

pCountSummary <- ggplot(clusterCountSummaryDf,
  aes(x = clustering, y = nClusters, fill = clustering)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = nClusters), vjust = -0.3, size = 4) +
  facet_wrap(~compartment, scales = "free_y") +
  scale_fill_manual(values = c("WNN" = "#4DAF4A", "RNA" = "#377EB8", "ATAC" = "#FF7F00")) +
  labs(
    x = "Clustering scheme",
    y = "# Clusters",
    title = "Cluster Counts at the Selected Resolution",
    subtitle = "Same numeric resolution across different graphs does not imply the same number of clusters"
  ) +
  theme(legend.position = "none")

print(pCountSummary)
savePlot(
  plot = pCountSummary,
  plotDir = plotDir,
  filename = "cluster_count_summary",
  height = 4,
  width = 9
)
```

# QC Metrics on Integrated UMAPs

```{r qcMetricsOnUMAPs, fig.height = 5, fig.width = 18, cache = FALSE}
qcPlotDir <- file.path(plotDir, "QC_UMAPs")
if (!dir.exists(qcPlotDir)) dir.create(qcPlotDir, recursive = TRUE)

qcMetricAvailability <- list()

for (comp in compartments) {
  seuratObj <- addDerivedQCMetrics(seuratList[[comp]])
  seuratList[[comp]] <- seuratObj

  availableSpecs <- Filter(function(spec) {
    spec$column %in% colnames(seuratObj@meta.data) &&
      sum(!is.na(seuratObj@meta.data[[spec$column]])) >= 10
  }, qcMetricSpecs)

  qcMetricAvailability[[comp]] <- data.frame(
    compartment = comp,
    metric = vapply(availableSpecs, `[[`, character(1), "column"),
    label = vapply(availableSpecs, `[[`, character(1), "label"),
    nNonMissing = vapply(availableSpecs, function(spec) {
      sum(!is.na(seuratObj@meta.data[[spec$column]]))
    }, numeric(1)),
    stringsAsFactors = FALSE
  )

  cat(sprintf("\n%s: plotting %d QC metrics on integrated UMAPs\n",
    comp, length(availableSpecs)))

  for (spec in availableSpecs) {
    pQC <- buildQCMetricPlot(seuratObj, spec, comp)
    if (is.null(pQC)) next

    print(pQC)
    savePlot(
      plot = pQC,
      plotDir = qcPlotDir,
      filename = paste0(comp, "_QC_UMAP_", spec$column),
      height = 5,
      width = 18
    )
  }
}

qcMetricAvailabilityDf <- bind_rows(qcMetricAvailability)
write.csv(
  qcMetricAvailabilityDf,
  file.path(tableDir, paste0(filenameSuffix, "_QC_metric_availability_for_UMAPs.csv")),
  row.names = FALSE
)
```

# WNN Visualization {.tabset}

## Overview UMAPs

```{r overviewUMAPs, fig.height = 5, fig.width = 18}
# Side-by-side RNA, ATAC, and WNN UMAPs colored by WNN clusters
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pRNA <- DimPlot(seuratObj, reduction = "umap.rna",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle(paste(comp, "— RNA UMAP")) + NoLegend()

  pATAC <- DimPlot(seuratObj, reduction = "umap.atac",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle(paste(comp, "— ATAC UMAP")) + NoLegend()

  pWNN <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle(paste(comp, "— WNN UMAP"))

  print(pRNA | pATAC | pWNN)

  savePlot(plot = pRNA | pATAC | pWNN, plotDir = plotDir,
    filename = paste0(comp, "_WNN_UMAP_overview"),
    height = 5, width = 18)
}
```

## WNN UMAP by Metadata

```{r wnnMetadataUMAPs, fig.height = 10, fig.width = 14}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pClust <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle("WNN Clusters")

  pDonor <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "donorID", cols = palDonorId) +
    ggtitle("Donor")

  pPhase <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "treatmentPhase", cols = palTreatmentPhase) +
    ggtitle("Treatment Phase")

  pVisit <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "visitNum", cols = palVisit) +
    ggtitle("Visit")

  combined <- (pClust | pDonor) / (pPhase | pVisit)
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_WNN_UMAP_metadata"),
    height = 10, width = 14)
}
```

## WNN UMAP Faceted by Phase

```{r wnnUMAPByPhase, fig.height = 5, fig.width = 20}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pFaceted <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "wnn_clusters", split.by = "treatmentPhase",
    label = TRUE, repel = TRUE, ncol = 4) +
    ggtitle(paste(comp, "— WNN Clusters by Treatment Phase"))

  print(pFaceted)

  savePlot(plot = pFaceted, plotDir = plotDir,
    filename = paste0(comp, "_WNN_UMAP_byPhase"),
    height = 5, width = 20)
}
```

## Modality Weights

```{r modalityWeights, fig.height = 8, fig.width = 10}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pDens <- ggplot(seuratObj@meta.data, aes(x = RNA.weight)) +
    geom_density(fill = "#377EB8", alpha = 0.5) +
    geom_vline(xintercept = 0.5, linetype = "dashed") +
    labs(x = "RNA Weight (ATAC Weight = 1 - RNA Weight)", y = "Density",
      title = paste(comp, "— Modality Weight Distribution")) +
    xlim(0, 1)

  pViolin <- VlnPlot(seuratObj, features = "RNA.weight",
    group.by = "wnn_clusters", pt.size = 0) +
    geom_hline(yintercept = 0.5, linetype = "dashed") +
    ggtitle(paste(comp, "— RNA Weight per WNN Cluster")) +
    NoLegend()

  print(pDens / pViolin)

  savePlot(plot = pDens / pViolin, plotDir = plotDir,
    filename = paste0(comp, "_WNN_modalityWeights"),
    height = 8, width = 10)
}
```

# WNN vs Modality-Specific Cluster Comparison

```{r clusterComparison, fig.height = 6, fig.width = 8, cache = FALSE}
clusterMappings <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  meta <- seuratObj@meta.data
  clusterMappings[[comp]] <- list()
  mappingPlots <- list()

  if ("rna_clusters" %in% colnames(meta)) {
    cat(sprintf("%s: comparing WNN clusters to RNA-only clusters\n", comp))
    mapping <- buildClusterMappingTables(meta, "rna_clusters", "RNA")
    clusterMappings[[comp]]$RNA <- mapping

    write.csv(
      mapping$counts,
      file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_WNN_vs_RNA_cluster_mapping_counts.csv")),
      row.names = FALSE
    )
    write.csv(
      mapping$proportions,
      file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_WNN_vs_RNA_cluster_mapping_proportions.csv")),
      row.names = FALSE
    )

    mappingPlots[["RNA"]] <- plotClusterMappingHeatmap(
      mapping$proportions,
      title = paste(comp, "— RNA-only → WNN Cluster Mapping"),
      yLabel = "RNA Cluster",
      highColor = "#377EB8"
    )
  }

  if ("atac_clusters" %in% colnames(meta)) {
    cat(sprintf("%s: comparing WNN clusters to ATAC-only clusters\n", comp))
    mapping <- buildClusterMappingTables(meta, "atac_clusters", "ATAC")
    clusterMappings[[comp]]$ATAC <- mapping

    write.csv(
      mapping$counts,
      file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_WNN_vs_ATAC_cluster_mapping_counts.csv")),
      row.names = FALSE
    )
    write.csv(
      mapping$proportions,
      file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_WNN_vs_ATAC_cluster_mapping_proportions.csv")),
      row.names = FALSE
    )

    mappingPlots[["ATAC"]] <- plotClusterMappingHeatmap(
      mapping$proportions,
      title = paste(comp, "— ATAC-only → WNN Cluster Mapping"),
      yLabel = "ATAC Cluster",
      highColor = "#FF7F00"
    )
  }

  if (length(mappingPlots) == 0) {
    cat(sprintf("%s: no RNA or ATAC cluster columns found for mapping\n", comp))
    next
  }

  combined <- wrap_plots(mappingPlots, ncol = length(mappingPlots))
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_WNN_cluster_mappings"),
    height = 6, width = 8 * length(mappingPlots))
}
```

# WNN Cluster Characterization

```{r clusterMarkers, cache = FALSE}
allMarkers <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"
  Idents(seuratObj) <- "wnn_clusters"

  cat(sprintf("\n%s: Finding markers for %d WNN clusters...\n",
    comp, length(unique(seuratObj$wnn_clusters))))

  markers <- FindAllMarkers(
    seuratObj,
    only.pos = TRUE,
    min.pct = 0.25,
    logfc.threshold = 0.25,
    test.use = "wilcox",
    verbose = FALSE
  )

  nSig <- sum(markers$p_val_adj < 0.05)
  cat(sprintf("  %d significant markers (p_val_adj < 0.05)\n", nSig))

  allMarkers[[comp]] <- markers
}
```

```{r markerDotplots, fig.height = 6, fig.width = 12, cache = FALSE}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  markers <- allMarkers[[comp]]
  DefaultAssay(seuratObj) <- "RNA"
  Idents(seuratObj) <- "wnn_clusters"

  topMarkers <- markers %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(n = 15, wt = avg_log2FC) %>%
    ungroup()

  if (nrow(topMarkers) > 0) {
    topGenes <- unique(topMarkers$gene)

    pDot <- createGeneDotPlot(
      seuratObj = seuratObj,
      genes = topGenes,
      groupBy = "wnn_clusters",
      title = paste0(comp, " — Top 15 WNN Cluster Markers (",
        length(unique(seuratObj$wnn_clusters)), " clusters)"),
      xlab = "Marker Genes",
      ylab = "WNN Cluster",
      plotDir = plotDir,
      filename = paste0(comp, "_WNN_clusterMarkers_dotplot"),
      orderGenes = TRUE,
      height = 6,
      width = 12
    )

    if (!is.null(pDot)) print(pDot)
  }
}
```

## Pseudobulk Marker Heatmap by Treatment Phase

```{r markerHeatmapByPhase, fig.height = 8, fig.width = 12, cache = FALSE}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  markers <- allMarkers[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  topMarkers <- markers %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) %>%
    ungroup()

  if (nrow(topMarkers) == 0) next
  topGenes <- unique(topMarkers$gene)

  heatmapData <- buildPseudobulkHeatmapData(
    seuratObj = seuratObj,
    genes = topGenes,
    groupCols = c("treatmentPhase", "wnn_clusters", "donorID")
  )
  if (is.null(heatmapData)) next

  annotDf <- heatmapData$annotation %>%
    dplyr::mutate(
      treatmentPhase = factor(treatmentPhase, levels = phaseLevels),
      wnn_clusters = factor(as.character(wnn_clusters),
        levels = gtools::mixedsort(unique(as.character(wnn_clusters)))),
      donorID = factor(as.character(donorID),
        levels = sort(unique(as.character(donorID))))
    ) %>%
    dplyr::arrange(treatmentPhase, wnn_clusters, donorID)

  matScaled <- heatmapData$matrix[, annotDf$groupID, drop = FALSE]
  colnames(matScaled) <- annotDf$groupID

  clusterLevels <- levels(annotDf$wnn_clusters)
  drawPseudobulkHeatmap(
    matScaled = matScaled,
    annotationDf = annotDf[, c("treatmentPhase", "wnn_clusters", "donorID"), drop = FALSE],
    annotationColors = list(
      treatmentPhase = palTreatmentPhase,
      wnn_clusters = setNames(scales::hue_pal()(length(clusterLevels)), clusterLevels),
      donorID = palDonorId[levels(annotDf$donorID)]
    ),
    splitCol = "treatmentPhase",
    title = paste0(comp, " — Pseudobulk Marker Genes by Treatment Phase"),
    plotDir = plotDir,
    filename = paste0(comp, "_WNN_markerHeatmap_byPhase_pseudobulk"),
    width = 12,
    height = 8
  )
}
```

## Pseudobulk Marker Heatmap by Visit

```{r markerHeatmapByVisit, fig.height = 8, fig.width = 14, cache = FALSE}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  markers <- allMarkers[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  topMarkers <- markers %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) %>%
    ungroup()

  if (nrow(topMarkers) == 0) next
  topGenes <- unique(topMarkers$gene)

  heatmapData <- buildPseudobulkHeatmapData(
    seuratObj = seuratObj,
    genes = topGenes,
    groupCols = c("visitNum", "treatmentPhase", "wnn_clusters", "donorID")
  )
  if (is.null(heatmapData)) next

  visitLevels <- sort(unique(as.numeric(as.character(seuratObj$visitNum))))
  annotDf <- heatmapData$annotation %>%
    dplyr::mutate(
      visitNum = factor(as.character(visitNum), levels = as.character(visitLevels)),
      treatmentPhase = factor(treatmentPhase, levels = phaseLevels),
      wnn_clusters = factor(as.character(wnn_clusters),
        levels = gtools::mixedsort(unique(as.character(wnn_clusters)))),
      donorID = factor(as.character(donorID),
        levels = sort(unique(as.character(donorID))))
    ) %>%
    dplyr::arrange(visitNum, wnn_clusters, donorID)

  matScaled <- heatmapData$matrix[, annotDf$groupID, drop = FALSE]
  colnames(matScaled) <- annotDf$groupID

  clusterLevels <- levels(annotDf$wnn_clusters)
  drawPseudobulkHeatmap(
    matScaled = matScaled,
    annotationDf = annotDf[, c("visitNum", "treatmentPhase", "wnn_clusters", "donorID"), drop = FALSE],
    annotationColors = list(
      visitNum = palVisit[levels(annotDf$visitNum)],
      treatmentPhase = palTreatmentPhase,
      wnn_clusters = setNames(scales::hue_pal()(length(clusterLevels)), clusterLevels),
      donorID = palDonorId[levels(annotDf$donorID)]
    ),
    splitCol = "visitNum",
    title = paste0(comp, " — Pseudobulk Marker Genes by Visit"),
    plotDir = plotDir,
    filename = paste0(comp, "_WNN_markerHeatmap_byVisit_pseudobulk"),
    width = 14,
    height = 8
  )
}
```

# Cluster Exploration: All Three Clusterings {.tabset}

## Cluster-Defining Markers (RNA & ATAC Clusters)

```{r rnaAtacClusterMarkers, cache = FALSE}
# FindAllMarkers for RNA-only and ATAC-only cluster identities (WNN already done above).
rnaClusterMarkers <- list()
atacClusterMarkers <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  # RNA cluster markers
  Idents(seuratObj) <- "rna_clusters"
  cat(sprintf("\n%s: Finding markers for %d RNA clusters...\n",
    comp, length(unique(seuratObj$rna_clusters))))
  rnaClusterMarkers[[comp]] <- FindAllMarkers(seuratObj, only.pos = TRUE,
    min.pct = 0.25, logfc.threshold = 0.25, test.use = "wilcox", verbose = FALSE)
  cat(sprintf("  %d significant markers (p_val_adj < 0.05)\n",
    sum(rnaClusterMarkers[[comp]]$p_val_adj < 0.05)))

  # ATAC cluster markers (still using RNA expression to define them)
  Idents(seuratObj) <- "atac_clusters"
  cat(sprintf("%s: Finding markers for %d ATAC clusters...\n",
    comp, length(unique(seuratObj$atac_clusters))))
  atacClusterMarkers[[comp]] <- FindAllMarkers(seuratObj, only.pos = TRUE,
    min.pct = 0.25, logfc.threshold = 0.25, test.use = "wilcox", verbose = FALSE)
  cat(sprintf("  %d significant markers (p_val_adj < 0.05)\n",
    sum(atacClusterMarkers[[comp]]$p_val_adj < 0.05)))
}
```

## Top 15 Marker Dotplots

```{r clusterDefiningDotplots, fig.height = 6, fig.width = 12, cache = FALSE}
clusteringSchemes <- list(
  list(name = "RNA",  clusterCol = "rna_clusters",  markerList = rnaClusterMarkers),
  list(name = "ATAC", clusterCol = "atac_clusters", markerList = atacClusterMarkers),
  list(name = "WNN",  clusterCol = "wnn_clusters",  markerList = allMarkers)
)

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  for (scheme in clusteringSchemes) {
    markers <- scheme$markerList[[comp]]
    if (is.null(markers) || nrow(markers) == 0) next

    topMarkers <- markers %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      group_by(cluster) %>%
      top_n(n = 15, wt = avg_log2FC) %>%
      ungroup()

    if (nrow(topMarkers) == 0) next
    topGenes <- unique(topMarkers$gene)

    Idents(seuratObj) <- scheme$clusterCol
    nClust <- length(unique(seuratObj@meta.data[[scheme$clusterCol]]))

    pDot <- createGeneDotPlot(
      seuratObj = seuratObj,
      genes = topGenes,
      groupBy = scheme$clusterCol,
      title = paste0(comp, " \u2014 Top 15 ", scheme$name, " Cluster Markers (",
        nClust, " clusters)"),
      xlab = "Cluster-defining genes",
      ylab = paste(scheme$name, "Cluster"),
      plotDir = plotDir,
      filename = paste0(comp, "_", scheme$name, "_clusterMarkers_top15_dotplot"),
      orderGenes = TRUE,
      height = 6,
      width = 12
    )

    if (!is.null(pDot)) print(pDot)
  }
}
```

## Cluster Composition Barplots

```{r clusterCompositionBarplots, fig.height = 14, fig.width = 18, cache = FALSE}
barTheme <- theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  meta <- seuratObj@meta.data

  # Build CellTypist L2 palette dynamically
  ctL2Levels <- sort(unique(as.character(meta$cellTypistCallingL2)))
  ctL2Levels <- ctL2Levels[!is.na(ctL2Levels)]
  nL2 <- length(ctL2Levels)
  if (nL2 > 0) {
    palCellTypistL2 <- setNames(scales::hue_pal()(nL2), ctL2Levels)
  } else {
    palCellTypistL2 <- character(0)
  }

  barPlots <- list()

  for (scheme in clusteringSchemes) {
    clCol <- scheme$clusterCol
    clName <- scheme$name

    plotMeta <- meta %>%
      tibble::rownames_to_column("cell") %>%
      dplyr::mutate(cluster = factor(.data[[clCol]],
        levels = gtools::mixedsort(unique(as.character(.data[[clCol]])))))

    countsPerCluster <- plotMeta %>%
      dplyr::count(cluster, name = "n") %>%
      dplyr::mutate(y = 1.02, label = as.character(n))

    # Donor
    pDonor <- ggplot(plotMeta, aes(x = cluster, fill = donorID)) +
      geom_bar(position = "fill") +
      scale_fill_manual(values = palDonorId, na.value = "grey80") +
      labs(title = paste(clName, "\u2014 Donor"), x = "Cluster",
        y = "Proportion", fill = "Donor") +
      barTheme +
      coord_cartesian(ylim = c(0, 1.08), clip = "off") +
      geom_text(data = countsPerCluster,
        aes(x = cluster, y = y, label = label),
        inherit.aes = FALSE, size = 3, vjust = 0)

    # Treatment Phase
    pPhase <- ggplot(plotMeta, aes(x = cluster, fill = treatmentPhase)) +
      geom_bar(position = "fill") +
      scale_fill_manual(values = palTreatmentPhase, na.value = "grey80") +
      labs(title = paste(clName, "\u2014 Treatment Phase"), x = "Cluster",
        y = "Proportion", fill = "Phase") +
      barTheme +
      coord_cartesian(ylim = c(0, 1.08), clip = "off") +
      geom_text(data = countsPerCluster,
        aes(x = cluster, y = y, label = label),
        inherit.aes = FALSE, size = 3, vjust = 0)

    # CellTypist L2
    pCT <- ggplot(plotMeta, aes(x = cluster, fill = cellTypistCallingL2)) +
      geom_bar(position = "fill") +
      scale_fill_manual(values = palCellTypistL2, na.value = "grey80") +
      labs(title = paste(clName, "\u2014 CellTypist L2"), x = "Cluster",
        y = "Proportion", fill = "CellTypist L2") +
      barTheme +
      coord_cartesian(ylim = c(0, 1.08), clip = "off") +
      geom_text(data = countsPerCluster,
        aes(x = cluster, y = y, label = label),
        inherit.aes = FALSE, size = 3, vjust = 0)

    barPlots[[paste0(clName, "_donor")]] <- pDonor
    barPlots[[paste0(clName, "_phase")]] <- pPhase
    barPlots[[paste0(clName, "_ct")]]    <- pCT
  }

  # 3x3 grid: rows = metadata variable, columns = clustering type
  combined <- (barPlots[["RNA_donor"]] | barPlots[["ATAC_donor"]] | barPlots[["WNN_donor"]]) /
    (barPlots[["RNA_phase"]] | barPlots[["ATAC_phase"]] | barPlots[["WNN_phase"]]) /
    (barPlots[["RNA_ct"]] | barPlots[["ATAC_ct"]] | barPlots[["WNN_ct"]]) +
    patchwork::plot_annotation(
      title = paste(comp, "\u2014 Cluster Composition by Metadata"),
      theme = theme(plot.title = element_text(size = 16, face = "bold")))

  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_cluster_composition_barplots"),
    height = 14, width = 18)
}
```

## Donor-Normalized WNN Cluster Abundance

```{r donorNormalizedClusterAbundance, fig.height = 8, fig.width = 14, cache = FALSE}
clusterAbundance <- list()
clusterAbundanceSummary <- list()

for (comp in compartments) {
  abundanceDf <- buildClusterAbundanceTable(seuratList[[comp]])
  clusterAbundance[[comp]] <- abundanceDf

  summaryDf <- abundanceDf %>%
    dplyr::filter(!is.na(proportion), totalCells > 0) %>%
    dplyr::group_by(cluster, treatmentPhase) %>%
    dplyr::summarise(
      nDonors = dplyr::n_distinct(donorID),
      medianProportion = median(proportion),
      meanProportion = mean(proportion),
      .groups = "drop"
    )
  clusterAbundanceSummary[[comp]] <- summaryDf

  write.csv(
    abundanceDf,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_WNN_cluster_abundance_by_donor_phase.csv")),
    row.names = FALSE
  )
  write.csv(
    summaryDf,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_WNN_cluster_abundance_summary.csv")),
    row.names = FALSE
  )

  pHeat <- ggplot(summaryDf,
    aes(x = treatmentPhase, y = cluster, fill = medianProportion)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.0f%%", 100 * medianProportion)), size = 3) +
    scale_fill_viridis_c(option = "C",
      labels = scales::percent_format(accuracy = 1),
      name = "Median donor proportion") +
    labs(
      x = "Treatment Phase",
      y = "WNN Cluster",
      title = paste(comp, "— Donor-Normalized WNN Cluster Abundance"),
      subtitle = "Median within-donor cluster proportion"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pHeat)
  savePlot(
    plot = pHeat,
    plotDir = plotDir,
    filename = paste0(comp, "_WNN_cluster_abundance_heatmap"),
    height = 6,
    width = 10
  )

  pTraj <- plotClusterAbundanceTrajectories(
    abundanceDf,
    paste(comp, "— Donor-Normalized WNN Cluster Abundance Trajectories")
  )

  print(pTraj)
  savePlot(
    plot = pTraj,
    plotDir = plotDir,
    filename = paste0(comp, "_WNN_cluster_abundance_trajectories"),
    height = max(6, 2.5 * ceiling(length(unique(abundanceDf$cluster)) / 3)),
    width = 14
  )
}
```

## WNN Cluster Phenotype Summary

```{r clusterPhenotypeSummary, cache = FALSE}
clusterPhenotypes <- list()

for (comp in compartments) {
  clusterPhenotypes[[comp]] <- buildClusterPhenotypeSummary(
    seuratObj = seuratList[[comp]],
    markers = allMarkers[[comp]]
  )

  outFile <- file.path(tableDir,
    paste0(filenameSuffix, "_", comp, "_WNN_cluster_phenotypes.csv"))
  write.csv(clusterPhenotypes[[comp]], file = outFile, row.names = FALSE)

  cat(sprintf("%s: cluster phenotype summary saved (%d clusters)\n",
    comp, nrow(clusterPhenotypes[[comp]])))
}
```

## HALLMARK IL2-STAT5 Signaling Score

```{r il2Stat5Score, cache = FALSE}
library(msigdbr)

il2Stat5Genes <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::filter(gs_name == "HALLMARK_IL2_STAT5_SIGNALING") %>%
  dplyr::pull(gene_symbol) %>%
  unique()

cat(sprintf("HALLMARK_IL2_STAT5_SIGNALING: %d genes\n", length(il2Stat5Genes)))

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  genesPresent <- il2Stat5Genes[il2Stat5Genes %in% rownames(seuratObj)]
  cat(sprintf("%s: %d / %d IL2-STAT5 genes present\n",
    comp, length(genesPresent), length(il2Stat5Genes)))

  seuratObj <- AddModuleScore(seuratObj, features = list(genesPresent),
    name = "IL2_STAT5_score", seed = 42)

  seuratList[[comp]] <- seuratObj
}
```

## IL2-STAT5 Donor-Phase Summaries

```{r il2Stat5Summaries, cache = FALSE}
il2Stat5Summaries <- list()

for (comp in compartments) {
  il2Stat5Summaries[[comp]] <- buildIl2Stat5Summary(seuratList[[comp]])

  write.csv(
    il2Stat5Summaries[[comp]]$donorPhaseCluster,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_IL2STAT5_scores_by_donor_phase_cluster.csv")),
    row.names = FALSE
  )
  write.csv(
    il2Stat5Summaries[[comp]]$phaseSummary,
    file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_IL2STAT5_scores_phase_summary.csv")),
    row.names = FALSE
  )

  cat(sprintf("%s: IL2-STAT5 donor-phase summaries saved (%d donor-phase-cluster groups)\n",
    comp, nrow(il2Stat5Summaries[[comp]]$donorPhaseCluster)))
}
```

## IL2-STAT5 UMAPs

```{r il2Stat5UMAPs, fig.height = 5, fig.width = 18, cache = FALSE}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pRNA <- FeaturePlot(seuratObj, features = "IL2_STAT5_score1",
    reduction = "umap.rna") +
    scale_color_viridis_c() +
    ggtitle(paste(comp, "\u2014 RNA UMAP"))

  pATAC <- FeaturePlot(seuratObj, features = "IL2_STAT5_score1",
    reduction = "umap.atac") +
    scale_color_viridis_c() +
    ggtitle(paste(comp, "\u2014 ATAC UMAP"))

  pWNN <- FeaturePlot(seuratObj, features = "IL2_STAT5_score1",
    reduction = "umap.wnn") +
    scale_color_viridis_c() +
    ggtitle(paste(comp, "\u2014 WNN UMAP"))

  combined <- pRNA | pATAC | pWNN
  combined <- combined + patchwork::plot_annotation(
    title = paste(comp, "\u2014 HALLMARK IL2-STAT5 Signaling Score"),
    theme = theme(plot.title = element_text(size = 16, face = "bold")))

  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_IL2_STAT5_UMAPs"),
    height = 5, width = 18)
}
```

## IL2-STAT5 Violins by Cluster

```{r il2Stat5Violins, fig.height = 6, fig.width = 14, cache = FALSE}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  meta <- seuratObj@meta.data

  vlnPlots <- list()
  for (scheme in clusteringSchemes) {
    clCol <- scheme$clusterCol
    clName <- scheme$name

    clLevels <- gtools::mixedsort(unique(as.character(meta[[clCol]])))
    nClust <- length(clLevels)
    clPal <- setNames(scales::hue_pal()(nClust), clLevels)

    pVln <- ggplot(meta, aes(x = factor(.data[[clCol]], levels = clLevels),
      y = IL2_STAT5_score1, fill = factor(.data[[clCol]], levels = clLevels))) +
      geom_violin(scale = "width", linewidth = 0.3) +
      geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white",
        alpha = 0.6, linewidth = 0.4) +
      scale_fill_manual(values = clPal) +
      labs(title = paste(clName, "Clusters"), x = "Cluster",
        y = "IL2-STAT5 Score") +
      theme_bw() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        legend.position = "none",
        plot.title = element_text(size = 12)
      )

    vlnPlots[[clName]] <- pVln
  }

  combined <- vlnPlots[["RNA"]] | vlnPlots[["ATAC"]] | vlnPlots[["WNN"]]
  combined <- combined + patchwork::plot_annotation(
    title = paste(comp, "\u2014 HALLMARK IL2-STAT5 Signaling by Cluster"),
    theme = theme(plot.title = element_text(size = 16, face = "bold")))

  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_IL2_STAT5_violins"),
    height = 6, width = 14)
}
```

## IL2-STAT5 Heatmap and Trajectories

```{r il2Stat5SummaryPlots, fig.height = 8, fig.width = 14, cache = FALSE}
for (comp in compartments) {
  phaseSummary <- il2Stat5Summaries[[comp]]$phaseSummary
  donorSummary <- il2Stat5Summaries[[comp]]$donorPhaseCluster

  pHeat <- plotIl2Stat5Heatmap(
    phaseSummary,
    paste(comp, "— IL2-STAT5 Median Donor Score by WNN Cluster")
  )
  print(pHeat)
  savePlot(
    plot = pHeat,
    plotDir = plotDir,
    filename = paste0(comp, "_IL2_STAT5_heatmap"),
    height = 6,
    width = 10
  )

  pTraj <- plotIl2Stat5Trajectories(
    donorSummary,
    paste(comp, "— IL2-STAT5 Donor Trajectories by WNN Cluster")
  )
  print(pTraj)
  savePlot(
    plot = pTraj,
    plotDir = plotDir,
    filename = paste0(comp, "_IL2_STAT5_trajectories"),
    height = max(6, 2.5 * ceiling(length(unique(donorSummary$cluster)) / 3)),
    width = 14
  )
}
```

# Save WNN Bundle

```{r saveBundle, cache = FALSE}
wnnBundle_path <- file.path(dataOutputDir,
  paste0(filenameSuffix, "_WNN_bundle.rds"))

wnnBundle <- list(
  seuratNK = seuratList[["NK"]],
  seuratTreg = seuratList[["Treg"]],
  macs2PeaksNK = macs2PeaksList[["NK"]],
  macs2PeaksTreg = macs2PeaksList[["Treg"]],
  wnnResolution = selectedRes,
  clusterMarkers = allMarkers,
  rnaClusterMarkers = rnaClusterMarkers,
  atacClusterMarkers = atacClusterMarkers,
  evalResults = evalResults,
  clusterCountSummary = clusterCountSummary,
  primaryContrastDefs = primaryContrastDefs,
  integrationSettings = integrationSettings,
  clusterCoverage = clusterCoverage,
  clusterReadiness = clusterReadiness,
  clusterAbundance = clusterAbundance,
  clusterAbundanceSummary = clusterAbundanceSummary,
  clusterMappings = clusterMappings,
  clusterPhenotypes = clusterPhenotypes,
  il2Stat5Summaries = il2Stat5Summaries,

  # Carry forward
  tcrGeneList = tcrGeneList,
  palDonorId = palDonorId,
  palVisit = palVisit,
  palTreatmentPhase = palTreatmentPhase,
  sourcePeakCallingBundles = sourcePeakCallingBundles,

  # Metadata
  dataDate = dataDate,
  bundle_created = Sys.time(),
  source_script = "P697_WNN.rmd"
)

saveRDS(wnnBundle, file = wnnBundle_path)
cat("WNN bundle saved to:", wnnBundle_path, "\n")
cat("Bundle contains:", paste(names(wnnBundle), collapse = ", "), "\n")
```

# Session Info

```{r sessionInfo, results = "markup"}
sessionInfo()
```
