---
title: "P697: Differential Accessibility, Pseudobulk DGE, and GSEA"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
setwd("/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

library(knitr)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(viridis)
library(stringr)
library(patchwork)
library(Seurat)
library(Signac)
library(edgeR)
library(limma)
library(clusterProfiler)
library(msigdbr)
library(ggrepel)
library(RColorBrewer)

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      axis.text = element_text(colour = "black"),
      axis.ticks = element_line(colour = "black"),
      legend.key = element_blank(),
      strip.text.x = element_text(size = 14, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(fill = "white", colour = "black")))

opts_chunk$set(fig.width = 6, fig.height = 4.0, cache = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE, cache.lazy = FALSE, results = "hide")
opts_knit$set(root.dir = "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

options(stringsAsFactors = FALSE)
options(future.globals.maxSize = 1591289600)
```

# Setup

```{r setUpDirectories}
projNumber <- "P697"
baseDir <- "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2"
dataInputDir <- file.path(baseDir, "data/inputData")
plotDir <- file.path(baseDir, "figures")
dataOutputDir <- file.path(baseDir, "data/outputData")
tableDir <- file.path(baseDir, "tables")
if (!dir.exists(tableDir)) dir.create(tableDir, recursive = TRUE)
dataDate <- Sys.Date()
filenameSuffix <- paste0(projNumber, ".", dataDate)
```

```{r setupFunctions}
source(file.path(getwd(), "code/P697_functions.R"))
```

# Load WNN Bundle

```{r loadBundle}
bundleFiles <- list.files(dataOutputDir,
  pattern = "P697\\..*_WNN_bundle\\.rds$", full.names = TRUE)
bundleFile <- bundleFiles[which.max(file.mtime(bundleFiles))]
cat("Loading WNN bundle:", basename(bundleFile), "\n")

wnnBundle <- readRDS(bundleFile)

seuratList <- list(
  NK = wnnBundle$seuratNK,
  Treg = wnnBundle$seuratTreg
)
compartments <- c("NK", "Treg")

# Palettes
palDonorId <- wnnBundle$palDonorId
palVisit <- wnnBundle$palVisit
palTreatmentPhase <- wnnBundle$palTreatmentPhase

for (comp in compartments) {
  cat(sprintf("%s: %d cells, %d WNN clusters\n",
    comp, ncol(seuratList[[comp]]),
    length(unique(seuratList[[comp]]$wnn_clusters))))
}

rm(wnnBundle); gc(verbose = FALSE)
```

# Define Contrasts

```{r defineContrasts}
# Longitudinal contrasts (all relative to baseline and consecutive)
contrastDefs <- list(
  PostIL2_vs_BL    = c("Post-IL2", "Baseline"),
  PostRAPA_vs_BL   = c("Post-RAPA", "Baseline"),
  Washout_vs_BL    = c("Washout", "Baseline"),
  PostRAPA_vs_PostIL2  = c("Post-RAPA", "Post-IL2"),
  Washout_vs_PostRAPA  = c("Washout", "Post-RAPA")
)

cat("Contrasts defined:\n")
for (nm in names(contrastDefs)) {
  cat(sprintf("  %s: %s vs %s\n", nm, contrastDefs[[nm]][1], contrastDefs[[nm]][2]))
}
```

# Helper Functions

```{r helperFunctions}
# Pseudobulk aggregation: sum counts per donor × treatment phase
aggregatePseudobulk <- function(countMat, metadata, groupCols) {
  metadata$groupDGE <- apply(metadata[, groupCols, drop = FALSE], 1, paste, collapse = "_")

  # Sum counts per group
  groups <- unique(metadata$groupDGE)
  aggMat <- matrix(0, nrow = nrow(countMat), ncol = length(groups),
    dimnames = list(rownames(countMat), groups))

  for (g in groups) {
    cells <- which(metadata$groupDGE == g)
    if (length(cells) == 1) {
      aggMat[, g] <- countMat[, cells]
    } else {
      aggMat[, g] <- Matrix::rowSums(countMat[, cells])
    }
  }

  # Build design info
  designInfo <- metadata %>%
    distinct(groupDGE, .keep_all = TRUE) %>%
    dplyr::select(all_of(c(groupCols, "groupDGE")))
  designInfo$nCells <- as.integer(table(metadata$groupDGE)[designInfo$groupDGE])
  designInfo <- designInfo[match(colnames(aggMat), designInfo$groupDGE), ]

  list(counts = aggMat, designInfo = designInfo)
}

# Run limma-voom DGE/DA with paired donor design
runLimmaVoom <- function(aggCounts, designInfo, contrastDefs,
                         filterExpr = TRUE, minCellsPerGroup = 3) {

  # Require minimum cells per group
  keep <- designInfo$nCells >= minCellsPerGroup
  if (sum(!keep) > 0) {
    cat(sprintf("  Dropped %d groups with < %d cells\n", sum(!keep), minCellsPerGroup))
    aggCounts <- aggCounts[, keep]
    designInfo <- designInfo[keep, ]
  }

  # Paired design: ~ 0 + treatmentPhase + donorID
  designInfo$treatmentPhase <- factor(designInfo$treatmentPhase,
    levels = c("Baseline", "Post-IL2", "Post-RAPA", "Washout"))
  designInfo$donorID <- factor(designInfo$donorID)

  design <- model.matrix(~ 0 + treatmentPhase + donorID, data = designInfo)
  colnames(design) <- gsub("treatmentPhase", "", colnames(design))
  colnames(design) <- gsub("donorID", "D", colnames(design))
  colnames(design) <- make.names(colnames(design))

  # DGEList + filtering + normalization
  dge <- DGEList(counts = aggCounts)
  if (filterExpr) {
    keep <- filterByExpr(dge, design = design)
    cat(sprintf("  filterByExpr: keeping %d / %d features\n", sum(keep), length(keep)))
    dge <- dge[keep, , keep.lib.sizes = FALSE]
  }
  dge <- calcNormFactors(dge, method = "TMM")

  # voom
  v <- voom(dge, design, plot = FALSE)

  # Fit
  fit <- lmFit(v, design)

  # Build contrast matrix from contrastDefs
  # Only include contrasts where both levels are present in the design
  validContrasts <- list()
  for (nm in names(contrastDefs)) {
    lvl1 <- make.names(contrastDefs[[nm]][1])
    lvl2 <- make.names(contrastDefs[[nm]][2])
    if (lvl1 %in% colnames(design) && lvl2 %in% colnames(design)) {
      validContrasts[[nm]] <- paste0(lvl1, " - ", lvl2)
    }
  }

  if (length(validContrasts) == 0) {
    cat("  No valid contrasts — skipping\n")
    return(NULL)
  }

  contStr <- unlist(validContrasts)
  cont.matrix <- makeContrasts(contrasts = contStr, levels = design)
  colnames(cont.matrix) <- names(validContrasts)

  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2)

  # Extract results
  results <- list()
  for (nm in names(validContrasts)) {
    tt <- topTable(fit2, coef = nm, adjust.method = "BH",
      sort.by = "P", number = Inf)
    tt$feature <- rownames(tt)
    tt$contrast <- nm
    results[[nm]] <- tt
  }

  list(results = results, fit = fit2, voom = v, design = design,
    contMatrix = cont.matrix, designInfo = designInfo)
}

# Volcano plot
plotVolcano <- function(tt, contrastName, featureType = "Gene",
                        pCutoff = 0.05, fcCutoff = 0.5, nLabel = 15) {
  tt$sig <- ifelse(tt$adj.P.Val < pCutoff & abs(tt$logFC) > fcCutoff,
    ifelse(tt$logFC > 0, "Up", "Down"), "NS")
  tt$sig <- factor(tt$sig, levels = c("Up", "Down", "NS"))

  nUp <- sum(tt$sig == "Up")
  nDown <- sum(tt$sig == "Down")

  # Label top features
  topFeatures <- tt %>%
    dplyr::filter(sig != "NS") %>%
    arrange(adj.P.Val) %>%
    head(nLabel)

  ggplot(tt, aes(x = logFC, y = -log10(adj.P.Val), color = sig)) +
    geom_point(size = 0.8, alpha = 0.6) +
    geom_hline(yintercept = -log10(pCutoff), linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = c(-fcCutoff, fcCutoff), linetype = "dashed", color = "grey50") +
    scale_color_manual(values = c("Up" = "#E41A1C", "Down" = "#377EB8", "NS" = "grey70"),
      drop = FALSE) +
    geom_text_repel(data = topFeatures,
      aes(label = feature), size = 3, max.overlaps = 20, color = "black") +
    labs(x = "log2 Fold Change", y = "-log10(Adj. P-value)",
      title = contrastName,
      subtitle = sprintf("%s: %d up, %d down (FDR < %.2f, |logFC| > %.1f)",
        featureType, nUp, nDown, pCutoff, fcCutoff)) +
    theme(legend.position = "none")
}
```

# Global Pseudobulk RNA DGE {.tabset}

## Aggregation and Model Fitting

```{r globalRNADGE}
rnaResults <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Global RNA DGE ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  # Get raw counts
  counts <- GetAssayData(seuratObj, layer = "counts")

  # Filter genes: keep those expressed in >= 30% of cells
  cellCount <- ncol(counts)
  geneFilter <- Matrix::rowSums(counts > 0) >= 0.3 * cellCount
  cat(sprintf("  Gene filter (>=30%% cells): %d / %d genes pass\n",
    sum(geneFilter), length(geneFilter)))
  counts <- counts[geneFilter, ]

  # Build metadata for aggregation
  meta <- seuratObj@meta.data[, c("donorID", "treatmentPhase")]

  # Aggregate by donor × treatment phase
  agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))
  cat(sprintf("  Pseudobulk groups: %d\n", ncol(agg$counts)))

  # Run limma-voom
  rnaResults[[comp]] <- runLimmaVoom(agg$counts, agg$designInfo, contrastDefs)

  if (!is.null(rnaResults[[comp]])) {
    for (nm in names(rnaResults[[comp]]$results)) {
      tt <- rnaResults[[comp]]$results[[nm]]
      nSig <- sum(tt$adj.P.Val < 0.05)
      nFC <- sum(tt$adj.P.Val < 0.05 & abs(tt$logFC) > 0.5)
      cat(sprintf("  %s: %d DE genes (FDR<0.05), %d with |logFC|>0.5\n", nm, nSig, nFC))
    }
  }
}
```

## RNA Volcano Plots

```{r rnaVolcanos, fig.height = 10, fig.width = 14}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next

  plots <- list()
  for (nm in names(rnaResults[[comp]]$results)) {
    plots[[nm]] <- plotVolcano(rnaResults[[comp]]$results[[nm]],
      paste(comp, "RNA:", nm), featureType = "Gene")
  }

  nPlots <- length(plots)
  nCols <- min(3, nPlots)
  nRows <- ceiling(nPlots / nCols)

  combined <- wrap_plots(plots, ncol = nCols)
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_RNA_DGE_volcanos"),
    height = 6 * nRows, width = 6 * nCols)
}
```

## RNA Results Tables

```{r rnaResultsTables}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next

  for (nm in names(rnaResults[[comp]]$results)) {
    tt <- rnaResults[[comp]]$results[[nm]]
    outFile <- file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_RNA_DGE_", nm, ".csv"))
    write.csv(tt, file = outFile, row.names = FALSE)
    cat(sprintf("Saved: %s\n", basename(outFile)))
  }
}
```

# Global Pseudobulk ATAC DA {.tabset}

## Aggregation and Model Fitting

```{r globalATACDA}
atacResults <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Global ATAC DA ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  # Get ATAC peak counts
  counts <- GetAssayData(seuratObj, layer = "counts")

  # Filter peaks: keep those accessible in >= 5% of cells
  # (lower threshold for ATAC since peaks are sparser than genes)
  cellCount <- ncol(counts)
  peakFilter <- Matrix::rowSums(counts > 0) >= 0.05 * cellCount
  cat(sprintf("  Peak filter (>=5%% cells): %d / %d peaks pass\n",
    sum(peakFilter), length(peakFilter)))
  counts <- counts[peakFilter, ]

  # Build metadata for aggregation
  meta <- seuratObj@meta.data[, c("donorID", "treatmentPhase")]

  # Aggregate by donor × treatment phase
  agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))
  cat(sprintf("  Pseudobulk groups: %d\n", ncol(agg$counts)))

  # Run limma-voom
  atacResults[[comp]] <- runLimmaVoom(agg$counts, agg$designInfo, contrastDefs)

  if (!is.null(atacResults[[comp]])) {
    for (nm in names(atacResults[[comp]]$results)) {
      tt <- atacResults[[comp]]$results[[nm]]
      nSig <- sum(tt$adj.P.Val < 0.05)
      nFC <- sum(tt$adj.P.Val < 0.05 & abs(tt$logFC) > 0.5)
      cat(sprintf("  %s: %d DA peaks (FDR<0.05), %d with |logFC|>0.5\n", nm, nSig, nFC))
    }
  }
}
```

## ATAC Volcano Plots

```{r atacVolcanos, fig.height = 10, fig.width = 14}
for (comp in compartments) {
  if (is.null(atacResults[[comp]])) next

  plots <- list()
  for (nm in names(atacResults[[comp]]$results)) {
    plots[[nm]] <- plotVolcano(atacResults[[comp]]$results[[nm]],
      paste(comp, "ATAC:", nm), featureType = "Peak")
  }

  nPlots <- length(plots)
  nCols <- min(3, nPlots)
  nRows <- ceiling(nPlots / nCols)

  combined <- wrap_plots(plots, ncol = nCols)
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_ATAC_DA_volcanos"),
    height = 5 * nRows, width = 5 * nCols)
}
```

## ATAC Results Tables

```{r atacResultsTables}
for (comp in compartments) {
  if (is.null(atacResults[[comp]])) next

  for (nm in names(atacResults[[comp]]$results)) {
    tt <- atacResults[[comp]]$results[[nm]]
    outFile <- file.path(tableDir,
      paste0(filenameSuffix, "_", comp, "_ATAC_DA_", nm, ".csv"))
    write.csv(tt, file = outFile, row.names = FALSE)
    cat(sprintf("Saved: %s\n", basename(outFile)))
  }
}
```

# Per-Cluster RNA DGE {.tabset}

## Aggregation and Model Fitting

```{r perClusterRNADGE}
perClusterRNA <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Per-Cluster RNA DGE ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  clusters <- sort(unique(seuratObj$wnn_clusters))
  perClusterRNA[[comp]] <- list()

  for (cl in clusters) {
    cat(sprintf("\n  Cluster %s:\n", cl))

    # Subset to cluster
    cellIdx <- which(seuratObj$wnn_clusters == cl)
    if (length(cellIdx) < 20) {
      cat(sprintf("    Skipping: only %d cells\n", length(cellIdx)))
      next
    }

    counts <- GetAssayData(seuratObj, layer = "counts")[, cellIdx]

    # Filter genes: >= 30% of cells in this cluster
    geneFilter <- Matrix::rowSums(counts > 0) >= 0.3 * length(cellIdx)
    counts <- counts[geneFilter, ]
    cat(sprintf("    %d cells, %d genes pass filter\n", length(cellIdx), nrow(counts)))

    if (nrow(counts) < 50) {
      cat("    Skipping: too few genes pass filter\n")
      next
    }

    meta <- seuratObj@meta.data[cellIdx, c("donorID", "treatmentPhase")]

    # Check: need at least 2 treatment phases with >= 2 donors each
    phaseTab <- table(meta$treatmentPhase, meta$donorID)
    phasesWithData <- rowSums(phaseTab > 0) >= 2
    if (sum(phasesWithData) < 2) {
      cat("    Skipping: insufficient donor coverage across phases\n")
      next
    }

    agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))

    result <- tryCatch(
      runLimmaVoom(agg$counts, agg$designInfo, contrastDefs),
      error = function(e) { cat(sprintf("    Error: %s\n", e$message)); NULL }
    )

    if (!is.null(result)) {
      perClusterRNA[[comp]][[as.character(cl)]] <- result
      for (nm in names(result$results)) {
        nSig <- sum(result$results[[nm]]$adj.P.Val < 0.05)
        cat(sprintf("    %s: %d DE genes\n", nm, nSig))
      }
    }
  }
}
```

## Per-Cluster RNA Summary Heatmap

```{r perClusterRNASummary, fig.height = 6, fig.width = 10}
for (comp in compartments) {
  if (length(perClusterRNA[[comp]]) == 0) next

  # Count significant genes per cluster × contrast
  summaryDf <- do.call(rbind, lapply(names(perClusterRNA[[comp]]), function(cl) {
    do.call(rbind, lapply(names(perClusterRNA[[comp]][[cl]]$results), function(nm) {
      tt <- perClusterRNA[[comp]][[cl]]$results[[nm]]
      data.frame(
        cluster = cl,
        contrast = nm,
        nDE = sum(tt$adj.P.Val < 0.05),
        nUp = sum(tt$adj.P.Val < 0.05 & tt$logFC > 0),
        nDown = sum(tt$adj.P.Val < 0.05 & tt$logFC < 0)
      )
    }))
  }))

  pSummary <- ggplot(summaryDf, aes(x = contrast, y = cluster, fill = nDE)) +
    geom_tile(color = "white") +
    geom_text(aes(label = nDE), size = 3.5) +
    scale_fill_gradient(low = "white", high = "#E41A1C", name = "# DE Genes") +
    labs(x = "", y = "WNN Cluster",
      title = paste(comp, "— Per-Cluster DE Genes (RNA, FDR < 0.05)")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pSummary)

  savePlot(plot = pSummary, plotDir = plotDir,
    filename = paste0(comp, "_perCluster_RNA_DGE_summary"),
    height = 6, width = 10)
}
```

## Per-Cluster RNA Tables

```{r perClusterRNATables}
for (comp in compartments) {
  for (cl in names(perClusterRNA[[comp]])) {
    for (nm in names(perClusterRNA[[comp]][[cl]]$results)) {
      tt <- perClusterRNA[[comp]][[cl]]$results[[nm]]
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_cluster", cl, "_RNA_DGE_", nm, ".csv"))
      write.csv(tt, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: per-cluster RNA tables saved\n", comp))
}
```

# Per-Cluster ATAC DA {.tabset}

## Aggregation and Model Fitting

```{r perClusterATACDA}
perClusterATAC <- list()

for (comp in compartments) {
  cat(sprintf("\n=== %s: Per-Cluster ATAC DA ===\n", comp))
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  clusters <- sort(unique(seuratObj$wnn_clusters))
  perClusterATAC[[comp]] <- list()

  for (cl in clusters) {
    cat(sprintf("\n  Cluster %s:\n", cl))

    cellIdx <- which(seuratObj$wnn_clusters == cl)
    if (length(cellIdx) < 20) {
      cat(sprintf("    Skipping: only %d cells\n", length(cellIdx)))
      next
    }

    counts <- GetAssayData(seuratObj, layer = "counts")[, cellIdx]

    # Filter peaks: >= 5% of cells in this cluster
    peakFilter <- Matrix::rowSums(counts > 0) >= 0.05 * length(cellIdx)
    counts <- counts[peakFilter, ]
    cat(sprintf("    %d cells, %d peaks pass filter\n", length(cellIdx), nrow(counts)))

    if (nrow(counts) < 50) {
      cat("    Skipping: too few peaks pass filter\n")
      next
    }

    meta <- seuratObj@meta.data[cellIdx, c("donorID", "treatmentPhase")]

    phaseTab <- table(meta$treatmentPhase, meta$donorID)
    phasesWithData <- rowSums(phaseTab > 0) >= 2
    if (sum(phasesWithData) < 2) {
      cat("    Skipping: insufficient donor coverage across phases\n")
      next
    }

    agg <- aggregatePseudobulk(counts, meta, c("donorID", "treatmentPhase"))

    result <- tryCatch(
      runLimmaVoom(agg$counts, agg$designInfo, contrastDefs),
      error = function(e) { cat(sprintf("    Error: %s\n", e$message)); NULL }
    )

    if (!is.null(result)) {
      perClusterATAC[[comp]][[as.character(cl)]] <- result
      for (nm in names(result$results)) {
        nSig <- sum(result$results[[nm]]$adj.P.Val < 0.05)
        cat(sprintf("    %s: %d DA peaks\n", nm, nSig))
      }
    }
  }
}
```

## Per-Cluster ATAC Summary Heatmap

```{r perClusterATACSummary, fig.height = 6, fig.width = 10}
for (comp in compartments) {
  if (length(perClusterATAC[[comp]]) == 0) next

  summaryDf <- do.call(rbind, lapply(names(perClusterATAC[[comp]]), function(cl) {
    do.call(rbind, lapply(names(perClusterATAC[[comp]][[cl]]$results), function(nm) {
      tt <- perClusterATAC[[comp]][[cl]]$results[[nm]]
      data.frame(
        cluster = cl,
        contrast = nm,
        nDA = sum(tt$adj.P.Val < 0.05),
        nUp = sum(tt$adj.P.Val < 0.05 & tt$logFC > 0),
        nDown = sum(tt$adj.P.Val < 0.05 & tt$logFC < 0)
      )
    }))
  }))

  pSummary <- ggplot(summaryDf, aes(x = contrast, y = cluster, fill = nDA)) +
    geom_tile(color = "white") +
    geom_text(aes(label = nDA), size = 3.5) +
    scale_fill_gradient(low = "white", high = "#FF7F00", name = "# DA Peaks") +
    labs(x = "", y = "WNN Cluster",
      title = paste(comp, "— Per-Cluster DA Peaks (ATAC, FDR < 0.05)")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(pSummary)

  savePlot(plot = pSummary, plotDir = plotDir,
    filename = paste0(comp, "_perCluster_ATAC_DA_summary"),
    height = 6, width = 10)
}
```

## Per-Cluster ATAC Tables

```{r perClusterATACTables}
for (comp in compartments) {
  for (cl in names(perClusterATAC[[comp]])) {
    for (nm in names(perClusterATAC[[comp]][[cl]]$results)) {
      tt <- perClusterATAC[[comp]][[cl]]$results[[nm]]
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_cluster", cl, "_ATAC_DA_", nm, ".csv"))
      write.csv(tt, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: per-cluster ATAC tables saved\n", comp))
}
```

# GSEA on Global RNA DGE {.tabset}

## Load Gene Sets

```{r loadGeneSets}
hallmarkSets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

goSets <- msigdbr(species = "Homo sapiens", category = "C5",
  subcategory = "GO:BP") %>%
  dplyr::select(gs_name, gene_symbol)

cat(sprintf("Hallmark: %d gene sets\n", length(unique(hallmarkSets$gs_name))))
cat(sprintf("GO:BP: %d gene sets\n", length(unique(goSets$gs_name))))
```

## Run GSEA

```{r runGSEA}
gseaResults <- list()

for (comp in compartments) {
  if (is.null(rnaResults[[comp]])) next
  gseaResults[[comp]] <- list()

  for (nm in names(rnaResults[[comp]]$results)) {
    tt <- rnaResults[[comp]]$results[[nm]]

    # Rank by t-statistic (more stable than logFC for GSEA)
    geneRanks <- tt$t
    names(geneRanks) <- tt$feature
    geneRanks <- sort(geneRanks, decreasing = TRUE)

    # Remove duplicates and NAs
    geneRanks <- geneRanks[!is.na(geneRanks)]
    geneRanks <- geneRanks[!duplicated(names(geneRanks))]

    cat(sprintf("\n%s %s: GSEA with %d ranked genes\n", comp, nm, length(geneRanks)))

    # Hallmark GSEA
    hallmarkGSEA <- tryCatch(
      GSEA(geneRanks, TERM2GENE = hallmarkSets,
        minGSSize = 10, maxGSSize = 500,
        pvalueCutoff = 1.0, verbose = FALSE),
      error = function(e) { cat(sprintf("  Hallmark error: %s\n", e$message)); NULL }
    )

    # GO:BP GSEA
    goGSEA <- tryCatch(
      GSEA(geneRanks, TERM2GENE = goSets,
        minGSSize = 10, maxGSSize = 500,
        pvalueCutoff = 1.0, verbose = FALSE),
      error = function(e) { cat(sprintf("  GO:BP error: %s\n", e$message)); NULL }
    )

    gseaResults[[comp]][[nm]] <- list(
      hallmark = hallmarkGSEA,
      go_bp = goGSEA
    )

    if (!is.null(hallmarkGSEA)) {
      nSigH <- sum(hallmarkGSEA@result$p.adjust < 0.05)
      cat(sprintf("  Hallmark: %d significant (FDR<0.05)\n", nSigH))
    }
    if (!is.null(goGSEA)) {
      nSigG <- sum(goGSEA@result$p.adjust < 0.05)
      cat(sprintf("  GO:BP: %d significant (FDR<0.05)\n", nSigG))
    }
  }
}
```

## GSEA Bubble Plots

```{r gseaBubblePlots, fig.height = 10, fig.width = 12}
for (comp in compartments) {
  if (is.null(gseaResults[[comp]])) next

  for (nm in names(gseaResults[[comp]])) {
    gseaSets <- gseaResults[[comp]][[nm]]
    plotData <- list()

    # Hallmark results
    if (!is.null(gseaSets$hallmark) && nrow(gseaSets$hallmark@result) > 0) {
      hRes <- gseaSets$hallmark@result %>%
        dplyr::filter(p.adjust < 0.1) %>%
        mutate(GeneSet = "Hallmark",
          Description_clean = gsub("^HALLMARK_", "", Description),
          Description_clean = gsub("_", " ", Description_clean),
          Description_clean = str_to_title(Description_clean))

      if (nrow(hRes) > 0) plotData[["Hallmark"]] <- hRes
    }

    # GO:BP results (top 20 by p.adjust)
    if (!is.null(gseaSets$go_bp) && nrow(gseaSets$go_bp@result) > 0) {
      gRes <- gseaSets$go_bp@result %>%
        dplyr::filter(p.adjust < 0.1) %>%
        arrange(p.adjust) %>%
        head(20) %>%
        mutate(GeneSet = "GO:BP",
          Description_clean = gsub("^GOBP_", "", Description),
          Description_clean = gsub("_", " ", Description_clean),
          Description_clean = str_to_title(Description_clean))

      if (nrow(gRes) > 0) plotData[["GO:BP"]] <- gRes
    }

    if (length(plotData) == 0) {
      cat(sprintf("%s %s: No significant GSEA results to plot\n", comp, nm))
      next
    }

    combined <- bind_rows(plotData)

    pBubble <- ggplot(combined, aes(x = NES, y = reorder(Description_clean, NES))) +
      geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
      geom_segment(aes(y = reorder(Description_clean, NES),
        yend = reorder(Description_clean, NES),
        x = 0, xend = NES), color = "black", linewidth = 0.5) +
      geom_point(aes(size = setSize, fill = -log10(p.adjust)),
        shape = 21, color = "black", stroke = 0.5) +
      facet_grid(GeneSet ~ ., scales = "free_y", space = "free_y") +
      scale_fill_gradient(low = "lightblue", high = "red",
        name = "-log10(FDR)") +
      scale_size_area(max_size = 8,
        guide = guide_legend(override.aes = list(fill = NA, color = "black"))) +
      labs(x = "Normalized Enrichment Score (NES)", y = "",
        size = "Gene Set Size",
        title = paste(comp, "—", nm)) +
      theme(
        strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
        strip.text = element_text(face = "bold", size = 12),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.y = element_text(size = 10),
        panel.grid.major.x = element_line(color = "gray95"),
        panel.grid.major.y = element_blank())

    print(pBubble)

    savePlot(plot = pBubble, plotDir = plotDir,
      filename = paste0(comp, "_GSEA_", nm),
      height = max(6, 0.4 * nrow(combined) + 2), width = 12)
  }
}
```

## GSEA Results Tables

```{r gseaResultsTables}
for (comp in compartments) {
  if (is.null(gseaResults[[comp]])) next

  for (nm in names(gseaResults[[comp]])) {
    gseaSets <- gseaResults[[comp]][[nm]]

    for (setType in c("hallmark", "go_bp")) {
      if (is.null(gseaSets[[setType]])) next
      res <- gseaSets[[setType]]@result
      outFile <- file.path(tableDir,
        paste0(filenameSuffix, "_", comp, "_GSEA_", setType, "_", nm, ".csv"))
      write.csv(res, file = outFile, row.names = FALSE)
    }
  }
  cat(sprintf("%s: GSEA tables saved\n", comp))
}
```

# Cross-Modality Integration Summary {.tabset}

## DA-DGE Overlap

```{r daDgeOverlap, fig.height = 6, fig.width = 8}
# For each contrast, identify peaks near DE genes and vice versa
for (comp in compartments) {
  if (is.null(rnaResults[[comp]]) || is.null(atacResults[[comp]])) next

  overlapSummary <- data.frame()

  for (nm in intersect(names(rnaResults[[comp]]$results),
                       names(atacResults[[comp]]$results))) {
    rna <- rnaResults[[comp]]$results[[nm]]
    atac <- atacResults[[comp]]$results[[nm]]

    nDE <- sum(rna$adj.P.Val < 0.05)
    nDA <- sum(atac$adj.P.Val < 0.05)

    overlapSummary <- rbind(overlapSummary, data.frame(
      contrast = nm,
      DE_genes = nDE,
      DA_peaks = nDA
    ))
  }

  cat(sprintf("\n%s: DA-DGE Summary\n", comp))
  print(overlapSummary)

  if (nrow(overlapSummary) > 0) {
    overlapLong <- overlapSummary %>%
      pivot_longer(cols = c(DE_genes, DA_peaks),
        names_to = "modality", values_to = "count")

    pOverlap <- ggplot(overlapLong, aes(x = contrast, y = count, fill = modality)) +
      geom_col(position = "dodge") +
      scale_fill_manual(values = c("DE_genes" = "#377EB8", "DA_peaks" = "#FF7F00"),
        labels = c("DE Genes (RNA)", "DA Peaks (ATAC)")) +
      labs(x = "", y = "# Significant Features (FDR < 0.05)",
        fill = "", title = paste(comp, "— Significant Features per Contrast")) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    print(pOverlap)

    savePlot(plot = pOverlap, plotDir = plotDir,
      filename = paste0(comp, "_DA_DGE_overlap_barplot"),
      height = 6, width = 8)
  }
}
```

## LogFC Correlation (Top DE Peaks Nearest Genes)

```{r logFCCorrelation, fig.height = 6, fig.width = 8}
for (comp in compartments) {
  if (is.null(rnaResults[[comp]]) || is.null(atacResults[[comp]])) next
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "ATAC"

  for (nm in intersect(names(rnaResults[[comp]]$results),
                       names(atacResults[[comp]]$results))) {
    atac <- atacResults[[comp]]$results[[nm]]
    rna <- rnaResults[[comp]]$results[[nm]]

    # Get significant DA peaks
    sigPeaks <- atac %>% dplyr::filter(adj.P.Val < 0.1)
    if (nrow(sigPeaks) < 5) next

    # Find nearest gene for each significant peak
    peakGR <- StringToGRanges(sigPeaks$feature)

    nearestGene <- tryCatch({
      nearest <- ClosestFeature(seuratObj, regions = peakGR)
      nearest
    }, error = function(e) NULL)

    if (is.null(nearestGene) || nrow(nearestGene) == 0) next

    # Merge ATAC logFC with RNA logFC for nearest gene
    sigPeaks$nearestGene <- nearestGene$gene_name[
      match(sigPeaks$feature, nearestGene$query_region)]
    sigPeaks$distance <- nearestGene$distance[
      match(sigPeaks$feature, nearestGene$query_region)]

    # Filter: only peaks within 100kb of a gene
    sigPeaks <- sigPeaks %>%
      dplyr::filter(!is.na(nearestGene) & distance <= 100000)

    if (nrow(sigPeaks) < 5) next

    # Join with RNA results
    merged <- sigPeaks %>%
      inner_join(rna %>% dplyr::select(feature, logFC, adj.P.Val) %>%
        dplyr::rename(RNA_logFC = logFC, RNA_FDR = adj.P.Val),
        by = c("nearestGene" = "feature"))

    if (nrow(merged) < 5) next

    corTest <- cor.test(merged$logFC, merged$RNA_logFC, method = "spearman")

    pCorr <- ggplot(merged, aes(x = logFC, y = RNA_logFC)) +
      geom_point(aes(color = -log10(adj.P.Val)), size = 2, alpha = 0.7) +
      geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.5) +
      scale_color_viridis_c(name = "-log10(ATAC FDR)") +
      geom_text_repel(
        data = merged %>% arrange(adj.P.Val) %>% head(10),
        aes(label = nearestGene), size = 3, max.overlaps = 15) +
      labs(x = "ATAC Peak logFC", y = "Nearest Gene RNA logFC",
        title = paste(comp, nm),
        subtitle = sprintf("Spearman rho = %.2f, p = %.2e (n = %d peaks)",
          corTest$estimate, corTest$p.value, nrow(merged)))

    print(pCorr)

    savePlot(plot = pCorr, plotDir = plotDir,
      filename = paste0(comp, "_logFC_correlation_", nm),
      height = 6, width = 8)
  }
}
```

# Export for SCENIC+

```{r scenicExport}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  exportDir <- file.path(dataOutputDir, paste0("scenic_export_", comp))
  if (!dir.exists(exportDir)) dir.create(exportDir, recursive = TRUE)

  # RNA counts (sparse matrix)
  DefaultAssay(seuratObj) <- "RNA"
  rnaCounts <- GetAssayData(seuratObj, layer = "counts")
  Matrix::writeMM(rnaCounts, file.path(exportDir, "RNA_counts.mtx"))
  writeLines(rownames(rnaCounts), file.path(exportDir, "RNA_genes.txt"))
  writeLines(colnames(rnaCounts), file.path(exportDir, "RNA_barcodes.txt"))

  # ATAC counts (sparse matrix)
  DefaultAssay(seuratObj) <- "ATAC"
  atacCounts <- GetAssayData(seuratObj, layer = "counts")
  Matrix::writeMM(atacCounts, file.path(exportDir, "ATAC_counts.mtx"))
  writeLines(rownames(atacCounts), file.path(exportDir, "ATAC_peaks.txt"))
  writeLines(colnames(atacCounts), file.path(exportDir, "ATAC_barcodes.txt"))

  # Metadata
  metaCols <- c("donorID", "visitNum", "treatmentPhase", "wnn_clusters",
    "RNA.weight", "ATAC.weight")
  metaCols <- intersect(metaCols, colnames(seuratObj@meta.data))
  meta <- seuratObj@meta.data[, metaCols]
  write.csv(meta, file.path(exportDir, "metadata.csv"))

  # Embeddings
  for (redName in c("umap.wnn", "harmony", "harmony_lsi")) {
    if (redName %in% names(seuratObj@reductions)) {
      emb <- Embeddings(seuratObj, reduction = redName)
      write.csv(emb, file.path(exportDir, paste0("embedding_", redName, ".csv")))
    }
  }

  cat(sprintf("%s: SCENIC+ export saved to %s\n", comp, exportDir))
  cat(sprintf("  RNA: %d genes × %d cells\n", nrow(rnaCounts), ncol(rnaCounts)))
  cat(sprintf("  ATAC: %d peaks × %d cells\n", nrow(atacCounts), ncol(atacCounts)))
}
```

# Save DA Bundle

```{r saveDABundle}
daBundle_path <- file.path(dataOutputDir,
  paste0(filenameSuffix, "_DA_bundle.rds"))

daBundle <- list(
  # Global results
  rnaResults = rnaResults,
  atacResults = atacResults,

  # Per-cluster results
  perClusterRNA = perClusterRNA,
  perClusterATAC = perClusterATAC,

  # GSEA
  gseaResults = gseaResults,

  # Contrasts used
  contrastDefs = contrastDefs,

  # Carry forward palettes
  palDonorId = palDonorId,
  palVisit = palVisit,
  palTreatmentPhase = palTreatmentPhase,

  # Metadata
  dataDate = dataDate,
  bundle_created = Sys.time(),
  source_script = "P697_DA.rmd"
)

saveRDS(daBundle, file = daBundle_path)
cat("DA bundle saved to:", daBundle_path, "\n")
cat("Bundle contains:", paste(names(daBundle), collapse = ", "), "\n")
```

# Session Info

```{r sessionInfo, results = "markup"}
sessionInfo()
```
