---
title: "P697: WNN Multimodal Analysis (RNA + ATAC)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
setwd("/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

library(knitr)
library(dplyr)
library(tibble)
library(ggplot2)
library(viridis)
library(stringr)
library(patchwork)
library(Seurat)
library(Signac)
library(ggpointdensity)
library(RColorBrewer)
library(cowplot)
library(cluster)
library(ComplexHeatmap)
library(circlize)

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      axis.text = element_text(colour = "black"),
      axis.ticks = element_line(colour = "black"),
      legend.key = element_blank(),
      strip.text.x = element_text(size = 14, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(fill = "white", colour = "black")))

opts_chunk$set(fig.width = 6, fig.height = 4.0, cache = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE, cache.lazy = FALSE, results = "hide")
opts_knit$set(root.dir = "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2")

options(stringsAsFactors = FALSE)
options(future.globals.maxSize = 1591289600)
```

# Setup

```{r setUpDirectories}
projNumber <- "P697"
baseDir <- "/Users/tedwards/Documents/projects/P697_TEAseq_T1D_low_dose_IL2"
dataInputDir <- file.path(baseDir, "data/inputData")
plotDir <- file.path(baseDir, "figures")
dataOutputDir <- file.path(baseDir, "data/outputData")
dataDate <- Sys.Date()
filenameSuffix <- paste0(projNumber, ".", dataDate)
```

```{r setupFunctions}
source(file.path(getwd(), "code/P697_functions.R"))
```

# Load Peak Calling Bundles

```{r loadBundles}
compartments <- c("NK", "Treg")
seuratList <- list()
macs2PeaksList <- list()

for (comp in compartments) {
  pattern <- sprintf("P697\\..*_%s_peakCallingBundle\\.rds$", comp)
  bundleFiles <- list.files(dataOutputDir, pattern = pattern, full.names = TRUE)
  bundleFile <- bundleFiles[which.max(file.mtime(bundleFiles))]
  cat(sprintf("Loading %s bundle: %s\n", comp, basename(bundleFile)))

  bundle <- readRDS(bundleFile)

  objName <- paste0("seurat", comp)
  seuratList[[comp]] <- bundle[[objName]]
  macs2PeaksList[[comp]] <- bundle$macs2Peaks

  if (comp == compartments[1]) {
    tcrGeneList <- bundle$tcrGeneList
    palDonorId <- bundle$palDonorId
    palVisit <- bundle$palVisit
  }

  cat(sprintf("  %s: %d cells, assays: %s\n", comp,
    ncol(seuratList[[comp]]),
    paste(names(seuratList[[comp]]@assays), collapse = ", ")))
  cat(sprintf("  Reductions: %s\n",
    paste(names(seuratList[[comp]]@reductions), collapse = ", ")))

  rm(bundle); gc(verbose = FALSE)
}
```

# Create Treatment Phase Variable

```{r createTreatmentPhase}
# Map visitNum to treatment phase based on trial design:
#   LD IL-2 treatment → stopped → RAPA treatment → stopped → washout
#   V0:  Baseline (pre-treatment)
#   V4/5: End of IL-2 phase (Post-IL2)
#   V8/10: End of RAPA phase (Post-RAPA)
#   V12: ~6 months after all treatment cessation (Washout)
treatmentPhaseMap <- c(
  "0" = "Baseline",
  "4" = "Post-IL2", "5" = "Post-IL2",
  "8" = "Post-RAPA", "10" = "Post-RAPA",
  "12" = "Washout"
)

palTreatmentPhase <- c(
  "Baseline" = "#4DAF4A",
  "Post-IL2" = "#377EB8",
  "Post-RAPA" = "#FF7F00",
  "Washout" = "#984EA3"
)

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  seuratObj$treatmentPhase <- factor(
    unname(treatmentPhaseMap[as.character(seuratObj$visitNum)]),
    levels = c("Baseline", "Post-IL2", "Post-RAPA", "Washout")
  )

  cat(sprintf("\n%s treatmentPhase distribution:\n", comp))
  print(table(seuratObj$treatmentPhase, seuratObj$donorID))

  nNA <- sum(is.na(seuratObj$treatmentPhase))
  if (nNA > 0) cat(sprintf("  WARNING: %d cells with unmapped visitNum\n", nNA))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj)
```

# Recompute RNA UMAP on Subset

```{r recomputeRNAUMAP}
# The RNA UMAP from preprocessing was computed on all cells (including
# non-ATAC-passing). After subsetting in peak calling, coordinates are stale.
# Recompute for fair comparison with ATAC and WNN UMAPs.
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  set.seed(6022)
  seuratObj <- RunUMAP(
    seuratObj,
    reduction = "harmony",
    dims = 1:18,
    reduction.name = "umap.rna",
    reduction.key = "rnaUMAP_",
    verbose = FALSE
  )

  cat(sprintf("%s: RNA UMAP recomputed (%d cells)\n", comp, ncol(seuratObj)))
  seuratList[[comp]] <- seuratObj
}
rm(seuratObj); gc(verbose = FALSE)
```

# WNN Integration

```{r wnnIntegration}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  cat(sprintf("\n%s: Running FindMultiModalNeighbors...\n", comp))
  cat(sprintf("  RNA: harmony (dims 1:18)\n"))
  cat(sprintf("  ATAC: harmony_lsi (dims 1:29)\n"))

  seuratObj <- FindMultiModalNeighbors(
    seuratObj,
    reduction.list = list("harmony", "harmony_lsi"),
    dims.list = list(1:18, 1:29),
    modality.weight.name = c("RNA.weight", "ATAC.weight"),
    verbose = FALSE
  )

  cat(sprintf("  Median RNA weight: %.3f\n", median(seuratObj$RNA.weight)))
  cat(sprintf("  Median ATAC weight: %.3f\n", median(seuratObj$ATAC.weight)))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj); gc(verbose = FALSE)
```

# WNN UMAP

```{r wnnUMAP}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  set.seed(6022)
  seuratObj <- RunUMAP(
    seuratObj,
    nn.name = "weighted.nn",
    reduction.name = "umap.wnn",
    reduction.key = "wnnUMAP_",
    verbose = FALSE
  )

  cat(sprintf("%s: WNN UMAP computed\n", comp))
  seuratList[[comp]] <- seuratObj
}
rm(seuratObj); gc(verbose = FALSE)
```

# WNN Clustering Evaluation

```{r evaluateResolutions}
resolutions <- seq(0.1, 2.0, by = 0.1)
evalResults <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  # Build combined scaled embedding for silhouette computation
  rnaEmbed <- scale(Embeddings(seuratObj, "harmony")[, 1:18])
  atacEmbed <- scale(Embeddings(seuratObj, "harmony_lsi")[, 1:29])
  combinedEmbed <- cbind(rnaEmbed, atacEmbed)
  rm(rnaEmbed, atacEmbed)

  # Subsample for efficiency
  set.seed(42)
  nSample <- min(3000, nrow(combinedEmbed))
  sampleIdx <- sample(nrow(combinedEmbed), nSample)
  distSub <- dist(combinedEmbed[sampleIdx, ])
  rm(combinedEmbed)

  metrics <- data.frame(
    resolution = resolutions,
    nClusters = NA_integer_,
    meanSilhouette = NA_real_
  )

  cat(sprintf("\n%s: evaluating %d resolutions...\n", comp, length(resolutions)))
  for (j in seq_along(resolutions)) {
    res <- resolutions[j]
    seuratObj <- FindClusters(seuratObj, graph.name = "wsnn",
      resolution = res, verbose = FALSE)
    clusters <- as.integer(seuratObj$seurat_clusters)
    clustersSub <- clusters[sampleIdx]
    nUnique <- length(unique(clustersSub))

    metrics$nClusters[j] <- length(unique(clusters))

    if (nUnique > 1 && nUnique < nSample) {
      sil <- cluster::silhouette(clustersSub, distSub)
      metrics$meanSilhouette[j] <- mean(sil[, 3])
    }

    cat(sprintf("  res %.1f: %d clusters, sil %.3f\n",
      res, metrics$nClusters[j],
      ifelse(is.na(metrics$meanSilhouette[j]), NA, metrics$meanSilhouette[j])))
  }

  evalResults[[comp]] <- metrics
  seuratList[[comp]] <- seuratObj
  rm(distSub)
}
rm(seuratObj); gc(verbose = FALSE)
```

```{r clusterMetricsPlot, fig.height = 5, fig.width = 12}
allMetrics <- bind_rows(
  lapply(compartments, function(comp) {
    evalResults[[comp]] %>% mutate(compartment = comp)
  })
)

pSil <- ggplot(allMetrics, aes(x = resolution, y = meanSilhouette, color = compartment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("NK" = "#E41A1C", "Treg" = "#377EB8")) +
  labs(x = "Resolution", y = "Mean Silhouette Width",
    title = "WNN Clustering: Silhouette") +
  theme(legend.position = "bottom")

pNCl <- ggplot(allMetrics, aes(x = resolution, y = nClusters, color = compartment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("NK" = "#E41A1C", "Treg" = "#377EB8")) +
  labs(x = "Resolution", y = "Number of Clusters",
    title = "WNN Clustering: Cluster Count") +
  theme(legend.position = "bottom")

pSil | pNCl

savePlot(plot = pSil | pNCl, plotDir = plotDir,
  filename = "WNN_clusterMetrics", height = 5, width = 12)

for (comp in compartments) {
  bestIdx <- which.max(evalResults[[comp]]$meanSilhouette)
  cat(sprintf("%s best silhouette: res %.1f → %d clusters (sil = %.3f)\n",
    comp, evalResults[[comp]]$resolution[bestIdx],
    evalResults[[comp]]$nClusters[bestIdx],
    evalResults[[comp]]$meanSilhouette[bestIdx]))
}
```

# Select Final Resolution

```{r selectResolution}
# Default: resolution with highest mean silhouette.
# Override by uncommenting the manual lines below after inspecting the metrics.
selectedRes <- list()
for (comp in compartments) {
  bestIdx <- which.max(evalResults[[comp]]$meanSilhouette)
  selectedRes[[comp]] <- evalResults[[comp]]$resolution[bestIdx]
}

# Manual override (uncomment to use):
# selectedRes[["NK"]] <- 0.5
# selectedRes[["Treg"]] <- 0.5

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  seuratObj <- FindClusters(seuratObj, graph.name = "wsnn",
    resolution = selectedRes[[comp]], verbose = FALSE)

  seuratObj$wnn_clusters <- seuratObj$seurat_clusters

  cat(sprintf("\n%s: %d WNN clusters at resolution %.2f\n",
    comp, length(unique(seuratObj$wnn_clusters)), selectedRes[[comp]]))
  print(table(seuratObj$wnn_clusters))

  seuratList[[comp]] <- seuratObj
}
rm(seuratObj); gc(verbose = FALSE)
```

# WNN Visualization {.tabset}

## Overview UMAPs

```{r overviewUMAPs, fig.height = 5, fig.width = 18}
# Side-by-side RNA, ATAC, and WNN UMAPs colored by WNN clusters
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pRNA <- DimPlot(seuratObj, reduction = "umap.rna",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle(paste(comp, "— RNA UMAP")) + NoLegend()

  pATAC <- DimPlot(seuratObj, reduction = "umap.atac",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle(paste(comp, "— ATAC UMAP")) + NoLegend()

  pWNN <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle(paste(comp, "— WNN UMAP"))

  print(pRNA | pATAC | pWNN)

  savePlot(plot = pRNA | pATAC | pWNN, plotDir = plotDir,
    filename = paste0(comp, "_WNN_UMAP_overview"),
    height = 5, width = 18)
}
```

## WNN UMAP by Metadata

```{r wnnMetadataUMAPs, fig.height = 10, fig.width = 14}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pClust <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "wnn_clusters", label = TRUE, repel = TRUE) +
    ggtitle("WNN Clusters")

  pDonor <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "donorID", cols = palDonorId) +
    ggtitle("Donor")

  pPhase <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "treatmentPhase", cols = palTreatmentPhase) +
    ggtitle("Treatment Phase")

  pVisit <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "visitNum", cols = palVisit) +
    ggtitle("Visit")

  combined <- (pClust | pDonor) / (pPhase | pVisit)
  print(combined)

  savePlot(plot = combined, plotDir = plotDir,
    filename = paste0(comp, "_WNN_UMAP_metadata"),
    height = 10, width = 14)
}
```

## WNN UMAP Faceted by Phase

```{r wnnUMAPByPhase, fig.height = 5, fig.width = 20}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pFaceted <- DimPlot(seuratObj, reduction = "umap.wnn",
    group.by = "wnn_clusters", split.by = "treatmentPhase",
    label = TRUE, repel = TRUE, ncol = 4) +
    ggtitle(paste(comp, "— WNN Clusters by Treatment Phase"))

  print(pFaceted)

  savePlot(plot = pFaceted, plotDir = plotDir,
    filename = paste0(comp, "_WNN_UMAP_byPhase"),
    height = 5, width = 20)
}
```

## Modality Weights

```{r modalityWeights, fig.height = 8, fig.width = 10}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  pDens <- ggplot(seuratObj@meta.data, aes(x = RNA.weight)) +
    geom_density(fill = "#377EB8", alpha = 0.5) +
    geom_vline(xintercept = 0.5, linetype = "dashed") +
    labs(x = "RNA Weight (ATAC Weight = 1 - RNA Weight)", y = "Density",
      title = paste(comp, "— Modality Weight Distribution")) +
    xlim(0, 1)

  pViolin <- VlnPlot(seuratObj, features = "RNA.weight",
    group.by = "wnn_clusters", pt.size = 0) +
    geom_hline(yintercept = 0.5, linetype = "dashed") +
    ggtitle(paste(comp, "— RNA Weight per WNN Cluster")) +
    NoLegend()

  print(pDens / pViolin)

  savePlot(plot = pDens / pViolin, plotDir = plotDir,
    filename = paste0(comp, "_WNN_modalityWeights"),
    height = 8, width = 10)
}
```

# WNN vs RNA Cluster Comparison

```{r clusterComparison, fig.height = 6, fig.width = 8}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]

  # Find RNA cluster column from preprocessing
  rnaClusCol <- intersect(
    c("seurat_clusters_RNA_0p3", "RNA_snn_res.0.3", "seurat_clusters"),
    colnames(seuratObj@meta.data)
  )

  if (length(rnaClusCol) == 0) {
    cat(sprintf("%s: no RNA cluster column found, skipping comparison\n", comp))
    next
  }
  rnaClusCol <- rnaClusCol[1]
  cat(sprintf("%s: comparing WNN clusters to RNA clusters (%s)\n", comp, rnaClusCol))

  ct <- table(
    RNA = seuratObj@meta.data[[rnaClusCol]],
    WNN = seuratObj$wnn_clusters
  )
  ctProp <- prop.table(ct, margin = 1)
  ctDf <- as.data.frame(ctProp)
  colnames(ctDf) <- c("RNA_cluster", "WNN_cluster", "Proportion")

  pHeat <- ggplot(ctDf, aes(x = WNN_cluster, y = RNA_cluster, fill = Proportion)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.0f%%", 100 * Proportion)),
      size = 3, color = "black") +
    scale_fill_gradient(low = "white", high = "#377EB8") +
    labs(x = "WNN Cluster", y = "RNA Cluster",
      title = paste(comp, "— RNA → WNN Cluster Mapping"))

  print(pHeat)

  savePlot(plot = pHeat, plotDir = plotDir,
    filename = paste0(comp, "_WNN_vs_RNA_clusters"),
    height = 6, width = 8)
}
```

# WNN Cluster Characterization

```{r clusterMarkers}
allMarkers <- list()

for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  DefaultAssay(seuratObj) <- "RNA"
  Idents(seuratObj) <- "wnn_clusters"

  cat(sprintf("\n%s: Finding markers for %d WNN clusters...\n",
    comp, length(unique(seuratObj$wnn_clusters))))

  markers <- FindAllMarkers(
    seuratObj,
    only.pos = TRUE,
    min.pct = 0.25,
    logfc.threshold = 0.25,
    test.use = "wilcox",
    verbose = FALSE
  )

  nSig <- sum(markers$p_val_adj < 0.05)
  cat(sprintf("  %d significant markers (p_val_adj < 0.05)\n", nSig))

  allMarkers[[comp]] <- markers
}
```

```{r markerDotplots, fig.height = 6, fig.width = 12}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  markers <- allMarkers[[comp]]
  DefaultAssay(seuratObj) <- "RNA"
  Idents(seuratObj) <- "wnn_clusters"

  topMarkers <- markers %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) %>%
    ungroup()

  if (nrow(topMarkers) > 0) {
    topGenes <- unique(topMarkers$gene)

    pDot <- createGeneDotPlot(
      seuratObj = seuratObj,
      genes = topGenes,
      groupBy = "wnn_clusters",
      title = paste0(comp, " — Top WNN Cluster Markers (",
        length(unique(seuratObj$wnn_clusters)), " clusters)"),
      xlab = "Marker Genes",
      ylab = "WNN Cluster",
      plotDir = plotDir,
      filename = paste0(comp, "_WNN_clusterMarkers_dotplot"),
      orderGenes = TRUE,
      height = 6,
      width = max(10, 4 + length(topGenes) * 0.25)
    )

    if (!is.null(pDot)) print(pDot)
  }
}
```

## Marker Heatmap by Treatment Phase

```{r markerHeatmapByPhase, fig.height = 8, fig.width = 12}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  markers <- allMarkers[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  topMarkers <- markers %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) %>%
    ungroup()

  if (nrow(topMarkers) == 0) next
  topGenes <- unique(topMarkers$gene)

  # Get scaled expression for marker genes
  seuratObj <- ScaleData(seuratObj, features = topGenes, verbose = FALSE)
  matScaled <- as.matrix(GetAssayData(seuratObj, layer = "scale.data")[topGenes, , drop = FALSE])

  # Order cells by treatment phase, then by WNN cluster within phase
  cellOrder <- order(seuratObj$treatmentPhase, seuratObj$wnn_clusters)
  matScaled <- matScaled[, cellOrder]

  # Clamp extreme values for visualization
  matScaled[matScaled > 2.5] <- 2.5
  matScaled[matScaled < -2.5] <- -2.5

  # Build annotations
  phaseOrdered <- seuratObj$treatmentPhase[cellOrder]
  clusterOrdered <- factor(seuratObj$wnn_clusters[cellOrder])

  htAnnot <- ComplexHeatmap::HeatmapAnnotation(
    Phase = phaseOrdered,
    Cluster = clusterOrdered,
    col = list(
      Phase = palTreatmentPhase,
      Cluster = setNames(
        scales::hue_pal()(length(levels(clusterOrdered))),
        levels(clusterOrdered))
    ),
    show_annotation_name = TRUE,
    annotation_name_side = "left",
    annotation_legend_param = list(
      Phase = list(title = "Treatment Phase"),
      Cluster = list(title = "WNN Cluster")
    )
  )

  ht <- ComplexHeatmap::Heatmap(matScaled,
    name = "z-score",
    top_annotation = htAnnot,
    column_split = phaseOrdered,
    cluster_columns = FALSE,
    show_column_names = FALSE,
    row_names_gp = grid::gpar(fontsize = 8),
    column_title_gp = grid::gpar(fontsize = 10),
    column_title_rot = 0,
    col = colorRamp2(c(-2.5, 0, 2.5), c("blue", "white", "red")),
    column_title = paste0(comp, " — Marker Genes by Treatment Phase"),
    heatmap_legend_param = list(title = "z-score"),
    use_raster = TRUE,
    raster_quality = 3
  )

  ComplexHeatmap::draw(ht)

  pdf(file.path(plotDir, paste0(filenameSuffix, "_", comp,
    "_WNN_markerHeatmap_byPhase.pdf")), width = 12, height = 8)
  ComplexHeatmap::draw(ht)
  dev.off()

  png(file.path(plotDir, paste0(filenameSuffix, "_", comp,
    "_WNN_markerHeatmap_byPhase.png")),
    width = 12, height = 8, units = "in", res = 600)
  ComplexHeatmap::draw(ht)
  dev.off()
}
```

## Marker Heatmap by Visit

```{r markerHeatmapByVisit, fig.height = 8, fig.width = 14}
for (comp in compartments) {
  seuratObj <- seuratList[[comp]]
  markers <- allMarkers[[comp]]
  DefaultAssay(seuratObj) <- "RNA"

  topMarkers <- markers %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) %>%
    ungroup()

  if (nrow(topMarkers) == 0) next
  topGenes <- unique(topMarkers$gene)

  # Get scaled expression for marker genes
  seuratObj <- ScaleData(seuratObj, features = topGenes, verbose = FALSE)
  matScaled <- as.matrix(GetAssayData(seuratObj, layer = "scale.data")[topGenes, , drop = FALSE])

  # Order cells by visitNum, then by WNN cluster within visit
  visitFactor <- factor(seuratObj$visitNum,
    levels = sort(unique(as.numeric(as.character(seuratObj$visitNum)))))
  cellOrder <- order(visitFactor, seuratObj$wnn_clusters)
  matScaled <- matScaled[, cellOrder]

  matScaled[matScaled > 2.5] <- 2.5
  matScaled[matScaled < -2.5] <- -2.5

  visitOrdered <- visitFactor[cellOrder]
  clusterOrdered <- factor(seuratObj$wnn_clusters[cellOrder])
  phaseOrdered <- seuratObj$treatmentPhase[cellOrder]

  htAnnot <- ComplexHeatmap::HeatmapAnnotation(
    Visit = visitOrdered,
    Phase = phaseOrdered,
    Cluster = clusterOrdered,
    col = list(
      Visit = palVisit,
      Phase = palTreatmentPhase,
      Cluster = setNames(
        scales::hue_pal()(length(levels(clusterOrdered))),
        levels(clusterOrdered))
    ),
    show_annotation_name = TRUE,
    annotation_name_side = "left",
    annotation_legend_param = list(
      Visit = list(title = "Visit"),
      Phase = list(title = "Treatment Phase"),
      Cluster = list(title = "WNN Cluster")
    )
  )

  ht <- ComplexHeatmap::Heatmap(matScaled,
    name = "z-score",
    top_annotation = htAnnot,
    column_split = visitOrdered,
    cluster_columns = FALSE,
    show_column_names = FALSE,
    row_names_gp = grid::gpar(fontsize = 8),
    column_title_gp = grid::gpar(fontsize = 10),
    column_title_rot = 0,
    col = colorRamp2(c(-2.5, 0, 2.5), c("blue", "white", "red")),
    column_title = paste0(comp, " — Marker Genes by Visit"),
    heatmap_legend_param = list(title = "z-score"),
    use_raster = TRUE,
    raster_quality = 3
  )

  ComplexHeatmap::draw(ht)

  pdf(file.path(plotDir, paste0(filenameSuffix, "_", comp,
    "_WNN_markerHeatmap_byVisit.pdf")), width = 14, height = 8)
  ComplexHeatmap::draw(ht)
  dev.off()

  png(file.path(plotDir, paste0(filenameSuffix, "_", comp,
    "_WNN_markerHeatmap_byVisit.png")),
    width = 14, height = 8, units = "in", res = 600)
  ComplexHeatmap::draw(ht)
  dev.off()
}
```

# Save WNN Bundle

```{r saveBundle}
wnnBundle_path <- file.path(dataOutputDir,
  paste0(filenameSuffix, "_WNN_bundle.rds"))

wnnBundle <- list(
  seuratNK = seuratList[["NK"]],
  seuratTreg = seuratList[["Treg"]],
  macs2PeaksNK = macs2PeaksList[["NK"]],
  macs2PeaksTreg = macs2PeaksList[["Treg"]],
  wnnResolution = selectedRes,
  clusterMarkers = allMarkers,
  evalResults = evalResults,

  # Carry forward
  tcrGeneList = tcrGeneList,
  palDonorId = palDonorId,
  palVisit = palVisit,
  palTreatmentPhase = palTreatmentPhase,

  # Metadata
  dataDate = dataDate,
  bundle_created = Sys.time(),
  source_script = "P697_WNN.rmd"
)

saveRDS(wnnBundle, file = wnnBundle_path)
cat("WNN bundle saved to:", wnnBundle_path, "\n")
cat("Bundle contains:", paste(names(wnnBundle), collapse = ", "), "\n")
```

# Session Info

```{r sessionInfo, results = "markup"}
sessionInfo()
```
